---
title: "Describing Your Data"
subtitle: "Mean, spread, counting correctly, and group summaries"
description: "Mean, median, variance, standard deviation, and standard error; quantiles, range, and IQR; the length() trap and the sum(!is.na()) fix; a fast full-dataset overview with skimr; and getting tidy per-group summaries with group_by() + summarize() — all on the pine needle dataset."
categories: [Descriptive Statistics, R, tidyverse]
order: 3
week: 2
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 3
    fig-height: 3
    fig-dpi: 300
  pptx: default
execute:
  echo: true
---

## Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Imported data two ways (`read_csv()`, `read_excel()`)
- Wrangled with `filter()`, `select()`, `mutate()`, `arrange()`
- Caught duplicate rows with `distinct()`
- Saved a new file with `write_csv()`

::: callout-note
**✅ Key idea from Module 2**

You can reshape a dataset and hand off a clean file. Today we start **describing** it — the step that comes before any formal test.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  Mean and median
2.  Variance and standard deviation
3.  Quantiles, range, and IQR
4.  The `length()` trap — and the fix
5.  Standard error
6.  `skimr` — a fast full-dataset overview
7.  `group_by()` + `summarize()`
:::
::::
:::::::

## Part 1 · Our Question, Revisited {.divider}

## Our Question, Revisited

::::::: columns
:::: {.column width="60%"}
- **H₀:** No difference in needle length between the lee and windward sides
- **Hₐ:** Needle length differs between the two sides

Before we can formally *test* that with statistics, we need to *describe* what we actually measured.

::: callout-note
**Plan for this unit**

1.  **Today** — describe the data: center, spread, and group summaries
2.  **A later module** — formally test H₀ vs. Hₐ
:::
::::

:::: {.column width="40%"}
```{r}
library(tidyverse)

pine_df <- read_csv("data/pine_needles.csv")
```

::: callout-tip
**🖐 Recap**

`n_s` codes lee (`n`) vs. windward (`s`); `group` is the field team name. Four teams, four trees, six needles per side.
:::
::::
:::::::

## Part 2 · The Mean {.divider}

## The Mean

::::::: columns
:::: {.column width="60%"}
$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

- Add up every value, divide by *n*
- The **balance point** of the distribution
- **Sensitive to outliers** — one very large value pulls it

**Example:** Values 5, 6, 7, 8, **50** → mean = 15.2. Is that a fair summary? Not really — the outlier dominates.

::: callout-note
**✅ Why it matters**

The mean is the most commonly reported statistic — and it's the number every t-test, ANOVA, and regression is ultimately built around.
:::
::::

:::: {.column width="40%"}
```{r}
# Pull one side's values out as a vector -----
lee_lengths <- pine_df |>
  filter(n_s == "n") |>
  pull(length_mm)

lee_lengths

mean(lee_lengths)
```

::: callout-tip
**📖 New pattern**

`filter()` picks **rows**. `pull()` extracts one column as a plain **vector**, not a data frame.
:::
::::
:::::::

## Part 3 · The Median {.divider}

## The Median

::::::: columns
:::: {.column width="60%"}
**The middle value once the data are sorted**

| Sorted data            | Median                |
|------------------------|------------------------|
| 5, 6, **7**, 8, 50     | **7**                  |
| 5, 6, **7, 8**, 50, 51 | **(7 + 8) / 2 = 7.5**  |

- Not pulled around by outliers — it's **robust**
- If mean ≫ median → right-skewed data
- If mean ≈ median → roughly symmetric
::::

:::: {.column width="40%"}
```{r}
wind_lengths <- pine_df |>
  filter(n_s == "s") |>
  pull(length_mm)

median(lee_lengths)
median(wind_lengths)
```

::: callout-note
**✅ Key idea**

Report mean *and* median together the first time you look at a new variable — a big gap between them is a signal worth investigating.
:::
::::
:::::::

## Part 4 · Variance and Standard Deviation {.divider}

## Variance and Standard Deviation

::::::: columns
:::: {.column width="60%"}
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n - 1} \qquad s = \sqrt{s^2}$$

1.  Distance of each value from the mean
2.  **Square** it — makes everything positive
3.  Average those squared distances (over $n-1$)
4.  **Square root** brings variance back into the original units

**Larger spread = larger variance and SD.**

::: callout-note
Variance is in *squared* units (mm²) — hard to interpret directly. SD is why we usually report it instead.
:::
::::

:::: {.column width="40%"}
```{r}
var(lee_lengths)
sd(lee_lengths)

var(wind_lengths)
sd(wind_lengths)
```

::: callout-tip
**📖 Rule of thumb**

For roughly normal data: \~68% of values fall within mean ± 1 SD, \~95% within mean ± 2 SD.
:::
::::
:::::::

## Part 4b · Quantiles, Range, and IQR {.divider}

## Quantiles, Range, and IQR

::::::: columns
:::: {.column width="60%"}
```{r}
range(lee_lengths)              # min and max
max(lee_lengths) - min(lee_lengths)  # range as a single number

quantile(lee_lengths)           # 0/25/50/75/100th percentiles
IQR(lee_lengths)                # 75th percentile minus 25th
```

::: callout-note
**✅ Key idea**

The **IQR** is the range of the *middle 50%* of the data — it ignores the most extreme quarter on each end, which makes it robust to outliers the same way the median is.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

`quantile()`'s 50th percentile is exactly the median. `IQR()` is just `quantile(x, 0.75) - quantile(x, 0.25)` computed for you.

📖 R4DS §12 — EDA
:::
::::
:::::::

## Quantiles — Why This Matters for Module 4

::::::: columns
:::: {.column width="60%"}
```{r}
quantile(wind_lengths)
IQR(wind_lengths)
```
:::

:::: {.column width="40%"}
::: callout-tip
**📖 Preview**

The box in a boxplot **is** the IQR — its bottom and top edges are the 25th and 75th percentiles you just computed. Module 4's boxplots are these five numbers, drawn.
:::
::::
:::::::

## Part 5 · The `length()` Trap {.divider}

## The `length()` Trap

::: callout-important
**🔮 Predict first:** A vector has 5 slots, but 2 are `NA`. What does `length()` return? What *should* your sample size be for a mean or SE calculation?
:::

::::::: columns
:::: {.column width="60%"}
```{r}
x <- c(4, 3, NA, 7, NA)   # a vector with missing values

length(x)          # counts every slot — including NAs!
sum(is.na(x))       # how many are missing?
```
::::

:::: {.column width="40%"}
::: callout-note
**📖 New words**

- **`is.na(x)`** — `TRUE` where a value is missing
- **`!is.na(x)`** — the flip: `TRUE` where a value is real
- **`sum(!is.na(x))`** — counts the real values = the correct *n*
:::
::::
:::::::

## The `length()` Trap — Why It Matters

::::::: columns
:::: {.column width="60%"}
```{r}
# Using length() as n silently gives the WRONG answer
wrong_n  <- length(x)
wrong_se <- sd(x, na.rm = TRUE) / sqrt(wrong_n)

wrong_n
round(wrong_se, 3)
```
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

R will **not** warn you. `length()` happily returns 5 when only 3 values are real — a silent, common error.
:::
::::
:::::::

## Fix — `sum(!is.na())`

::::::: columns
:::: {.column width="60%"}
```{r}
is.na(x)          # TRUE where NA
!is.na(x)         # flipped: TRUE where real
sum(!is.na(x))    # count the TRUEs
```

::: callout-tip
**Best practice:** use `sum(!is.na(x))` for *n* every time you compute an SE — even when you're sure there are no missing values. Field data rarely stays that clean.
:::
::::

:::: {.column width="40%"}
**How it works, step by step:**

1.  `is.na(x)` → `TRUE` for each missing value
2.  `!is.na(x)` → flips it to `TRUE` for **real** values
3.  `sum(...)` → adds up the `TRUE`s (`TRUE` = 1)

📖 R4DS §18 — Missing Values
::::
:::::::

## Fix — Applied to Real Data

::::::: columns
:::: {.column width="60%"}
```{r}
# Applied to our real data — even with no NAs, make it a habit
n_lee  <- sum(!is.na(lee_lengths))
n_wind <- sum(!is.na(wind_lengths))

n_lee
n_wind
```
::::

:::: {.column width="40%"}
::: callout-tip
Even when you're confident a column has zero missing values, run it through `sum(!is.na())` anyway — it costs nothing and field data rarely stays clean for long.
:::
::::
:::::::

## Part 6 · Standard Error {.divider}

## Standard Error

::::::: columns
:::: {.column width="60%"}
$$SE = \frac{s}{\sqrt{n}}$$

- Measures how precisely you know the **mean** — not the spread of raw data
- Gets **smaller** as *n* grows: more data → a more precise estimate of the mean

|        | Describes                          |
|--------|-------------------------------------|
| **SD** | spread of *individual* data points |
| **SE** | precision of the *sample mean*     |

We report SE whenever we're making a claim about a group mean.
::::

:::: {.column width="40%"}
```{r}
se_lee  <- sd(lee_lengths)  / sqrt(n_lee)
se_wind <- sd(wind_lengths) / sqrt(n_wind)

round(se_lee, 2)
round(se_wind, 2)
```

::: callout-warning
**⚠️ Watch out!**

`sqrt(length(x))` and `sqrt(sum(!is.na(x)))` give the *same* answer only when there are zero missing values. Don't assume — check.
:::
::::
:::::::

## All Stats, One Group, Base R

::::::: columns
:::: {.column width="60%"}
```{r}
n_l    <- sum(!is.na(lee_lengths))
mean_l <- mean(lee_lengths, na.rm = TRUE)
med_l  <- median(lee_lengths, na.rm = TRUE)
sd_l   <- sd(lee_lengths, na.rm = TRUE)
se_l   <- sd_l / sqrt(n_l)

cat("--- Lee side ---\n",
    "n      =", n_l, "\n",
    "Mean   =", round(mean_l, 2), "\n",
    "Median =", round(med_l, 2), "\n",
    "SD     =", round(sd_l, 2), "\n",
    "SE     =", round(se_l, 2), "\n")
```
::::

:::: {.column width="40%"}
- Every function takes `na.rm = TRUE`
- `cat()` prints clean, labeled output

::: callout-note
**✅ Key idea**

Copy-pasting this block for every group is exactly the repetition `group_by()` + `summarize()` exists to remove.
:::
::::
:::::::

## Part 6b · `skimr` — a Fast Full-Dataset Overview {.divider}

## `skimr` — Every Column, One Command

::::::: columns
:::: {.column width="60%"}
```{r}
#| eval: false
install.packages("skimr")   # once per machine
```

```{r}
library(skimr)

skim(pine_df)
```
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`skim()` gives you n, missing count, mean, SD, and a tiny inline histogram for **every numeric column** — and counts for every character column — in one call. It's the fastest way to sanity-check a dataset the moment you load it.
:::
::::
:::::::

## `skimr` — Doesn't Replace, Just Speeds Up

::::::: columns
:::: {.column width="60%"}
```{r}
# Same idea, restricted to numeric columns
pine_df |>
  select(length_mm, tree_no) |>
  skim()
```
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`skim()` is a first look, not a final answer. It doesn't know that `length_mm` should be split by `n_s` — for real per-group comparisons, you still need `group_by()` + `summarize()`, next.
:::
::::
:::::::

## Part 7 · Tidy Group Stats — `group_by()` + `summarize()` {.divider}

## Tidy Group Stats — `group_by()` + `summarize()`

::: callout-note
**🔮 Predict first:** How many **rows** will the result have? (Hint: how many values does `n_s` take?)
:::

::::::: columns
:::: {.column width="60%"}
```{r}
side_stats_df <- pine_df |>
  group_by(n_s) |>
  summarize(
    n       = sum(!is.na(length_mm)),
    mean_mm = round(mean(length_mm, na.rm = TRUE), 2),
    med_mm  = round(median(length_mm, na.rm = TRUE), 2),
    sd_mm   = round(sd(length_mm, na.rm = TRUE), 2),
    se_mm   = round(sd_mm / sqrt(n), 2)
  )

side_stats_df
```
::::

:::: {.column width="40%"}
- `group_by(n_s)` splits the data by side
- `summarize()` collapses each group to **one row** of stats
- Both groups, one clean table — no copy-paste
- A column you just created (`sd_mm`) can feed the next one (`se_mm`)

📖 R4DS §3.5 — `summarize()`
::::
:::::::

## Grouping by More Than One Variable

::::::: columns
:::: {.column width="60%"}
```{r}
team_stats_df <- pine_df |>
  group_by(group, n_s) |>
  summarize(
    n       = sum(!is.na(length_mm)),
    mean_mm = round(mean(length_mm, na.rm = TRUE), 2),
    se_mm   = round(sd(length_mm, na.rm = TRUE) / sqrt(n), 2),
    .groups = "drop"
  )

team_stats_df
```
::::

:::: {.column width="40%"}
- `group_by(group, n_s)` splits by **field team, then side** — one row per combination
- `.groups = "drop"` turns off the grouping once you're done — good habit, avoids surprises later

::: callout-tip
**🖐 Notice**

Four teams × two sides = 8 rows. Does every team show the same lee-vs-windward pattern, or do some disagree?
:::
::::
:::::::

## Part 7b · Bringing Back `case_when()` from Module 2 {.divider}

## Summarizing the Categories You Built Last Class

::: callout-note
**🔮 Predict first:** We're grouping by the `size_class` bins (`"short"`/`"medium"`/`"long"`) from Module 2's `case_when()`. How many rows will the summary have?
:::

::::::: columns
:::: {.column width="60%"}
```{r}
size_class_stats_df <- pine_df |>
  mutate(
    size_class = case_when(
      length_mm < 16 ~ "short",
      length_mm < 20 ~ "medium",
      TRUE           ~ "long"
    )
  ) |>
  group_by(size_class) |>
  summarize(
    n       = sum(!is.na(length_mm)),
    mean_mm = round(mean(length_mm, na.rm = TRUE), 2),
    se_mm   = round(sd(length_mm, na.rm = TRUE) / sqrt(n), 2)
  )

size_class_stats_df
```
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`mutate()` and `group_by()` + `summarize()` chain together like any other verbs — you don't need to save the `case_when()` column separately first. This is the wrangling → describing pipeline in one block.
:::
::::
:::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Computed **mean** and **median**, and know why they can disagree
- Computed **variance** and **SD** by hand, with real formulas
- Computed **quantiles**, **range**, and **IQR** — and saw the IQR is literally the boxplot's box
- Found the **`length()` trap** and fixed it with `sum(!is.na())`
- Computed **SE** and know how it differs from SD
- Got a fast full-dataset overview with **`skimr::skim()`**
- Used `group_by()` + `summarize()` to get every group's stats in one table — including the `case_when()` categories from Module 2

::: callout-tip
**🖐 Before next class**

Finish the worksheet: compute n, mean, SD, and SE for `length_mm` grouped by `n_s`, and again grouped by `group`. Save the summary table with `write_csv()`.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Up next — Module 4**

- Boxplots and jittered points, done right
- Plotting mean ± SE directly with `stat_summary()`
- Faceting, themes, and `ggsave()` for a publication-ready figure
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` loaded? Spelling? Did you forget `na.rm = TRUE`?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
