---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Describing Your Data"
subtitle: "Mean, spread, counting correctly, and group summaries"
description: "Mean, median, variance, standard deviation, and standard error; quantiles, range, and IQR; the length() trap and the sum(!is.na()) fix; a fast full-dataset overview with skimr; and getting tidy per-group summaries with group_by() + summarize() — all on the pine needle dataset."
categories: [Descriptive Statistics, R, tidyverse]
order: 3
week: 2
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true
---

## Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Imported data two ways (`read_csv()`, `read_excel()`)
- Wrangled with `filter()`, `select()`, `mutate()`, `arrange()`
- Caught duplicate rows with `distinct()`
- Saved a new file with `write_csv()`

::: callout-note
**✅ Key idea from Wrangling Your Data**

You can reshape a dataset and hand off a clean file. Today we start **describing** it — the step that comes before any formal test.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  Mean and median
2.  Variance and standard deviation
3.  Quantiles, range, and IQR
4.  The `length()` trap — and the fix
5.  Standard error
6.  `skimr` — a fast full-dataset overview
7.  `group_by()` + `summarize()`
:::
::::
:::::::

## Part 1 · Our Question, Revisited {.divider}

## Our Question, Revisited

::::::: columns
:::: {.column width="60%"}
- **H₀:** No difference in needle length between the shady and sunny sides
- **Hₐ:** Needle length differs between the two sides

Before we can formally *test* that with statistics, we need to *describe* what we actually measured.

::: callout-note
**Plan for this unit**

1.  **Today** — describe the data: center, spread, and group summaries
2.  **A later module** — formally test H₀ vs. Hₐ
:::
::::

:::: {.column width="40%"}
```{r descriptive-stats-01}
library(tidyverse)

pine_df <- read_csv("data/pine_data.csv")
```

[📥 Download the data](data/pine_data.csv){.btn .btn-primary download="pine_data.csv"}

::: callout-tip
**🖐 Recap**

`side` is `"shady"` or `"sunny"`; `team` is the field team name.\
Four teams, four trees, eight needles per side — 64 rows, but only 4 trees.
:::
::::
:::::::

## Part 2 · The Mean {.divider}

## The Mean

::::::: columns
:::: {.column width="60%"}
$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

- Add up every value, divide by *n*
- The **balance point** of the distribution
- **Sensitive to outliers** — one very large value disproportionately affects the mean

**Example:** Values 5, 6, 7, 8, **50** → mean = 15.2.\
Is that a fair summary?\
Not really — the outlier dominates.

::: callout-note
**✅ Why it matters**

The mean is the most commonly reported statistic\
— and it's the number every t-test, ANOVA, and regression is ultimately built around.
:::
::::

:::: {.column width="40%"}
```{r descriptive-stats-02}
# Keep one side's rows -- still a data frame -----
shady_df <- pine_df |>
  filter(side == "shady")

shady_df |>
  summarize(mean_mm = mean(needle_length_mm))
```

::: callout-tip
**📖 New pattern**

`filter()` picks **rows** and gives back a data frame. `summarize()` collapses that data frame to **one row** of statistics — so we stay in data frames the whole way, exactly like every other analysis this term.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 3 — Describing Data, on the mean as a measure of location.\
Gotelli & Ellison, *A Primer of Ecological Statistics*, Ch. 3 — Summary Statistics: Measures of Location and Spread.
:::
::::
:::::::

## Part 3 · The Median {.divider}

## The Median

:::::: columns
::: {.column width="60%"}
**The middle value once the data are sorted**

| Sorted data            | Median                |
|------------------------|-----------------------|
| 5, 6, **7**, 8, 50     | **7**                 |
| 5, 6, **7, 8**, 50, 51 | **(7 + 8) / 2 = 7.5** |

- Not pulled around by outliers — it's **robust**
- If mean ≫ median → right-skewed data
- If mean ≈ median → roughly symmetric
:::

:::: {.column width="40%"}
```{r descriptive-stats-03}
sunny_df <- pine_df |>
  filter(side == "sunny")

shady_df |>
  summarize(med_mm = median(needle_length_mm))

sunny_df |>
  summarize(med_mm = median(needle_length_mm))
```

::: callout-note
**✅ Key idea**

Report mean *and* median together the first time you look at a new variable — a big gap between them is a signal worth investigating.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 3 — Describing Data, on the median and why a robust measure of location matters with skewed data.
:::
::::
::::::

## Part 4 · Variance (s^2^) and Standard Deviation (sd) {.divider}

## Variance and Standard Deviation

::::::: columns
:::: {.column width="60%"}
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n - 1} \qquad s = \sqrt{s^2}$$

1.  Distance of each value from the mean
2.  **Square** it — makes everything positive
3.  Average those squared distances (over $n-1$)
4.  **Square root** brings variance back into the original units as the standard deviation

**Larger spread = larger variance and SD.**

::: callout-note
- Variance is in *squared* units (mm²)
- hard to interpret directly.
- Standard deviation (SD) is why we usually report it instead
:::
::::

:::: {.column width="40%"}
```{r descriptive-stats-04}
shady_df |>
  summarize(var_mm = var(needle_length_mm),
            sd_mm  = sd(needle_length_mm))

sunny_df |>
  summarize(var_mm = var(needle_length_mm),
            sd_mm  = sd(needle_length_mm))
```

::: callout-tip
**📖 Rule of thumb**

- For roughly normal data:
  - \~68% of values fall within mean ± 1 SD
  - \~95% within mean ± 2 SD
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 3 — Describing Data, on variance, standard deviation, and the coefficient of variation.\
Gotelli & Ellison, Ch. 3 — Summary Statistics, for the same measures of spread.
:::
::::
:::::::

## Part 4b · Quantiles, Range, and IQR {.divider}

## Quantiles, Range, and IQR

::::::: columns
:::: {.column width="60%"}
```{r descriptive-stats-05}
shady_df |>
  summarize(
    min_mm   = min(needle_length_mm),      # the five numbers a
    q25_mm   = quantile(needle_length_mm, 0.25),   # boxplot draws
    med_mm   = median(needle_length_mm),
    q75_mm   = quantile(needle_length_mm, 0.75),
    max_mm   = max(needle_length_mm),
    range_mm = max_mm - min_mm,            # max minus min
    iqr_mm   = IQR(needle_length_mm)       # q75 minus q25
  )

box_plot_1 <- ggplot(shady_df, aes(y = needle_length_mm)) +
  geom_boxplot()
```

::: callout-note
**✅ Key idea**

The **IQR** is the range of the *middle 50%* of the data — it ignores the most extreme quarter on each end, which makes it robust to outliers the same way the median is.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

`med_mm` and `q50` are the same number. `iqr_mm` is just `q75_mm - q25_mm`, computed for you. And because `shady_df` is a data frame, it drops straight into `ggplot()` — no special handling.

[📖 R4DS §10 — Exploratory data analysis](https://r4ds.hadley.nz/EDA.html)

```{r descriptive-stats-06}
box_plot_1
```
:::
::::
:::::::

## Quantiles — Why This Matters for Graphing Effectively

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-07}
sunny_df |>
  summarize(
    q25_mm = quantile(needle_length_mm, 0.25),
    q75_mm = quantile(needle_length_mm, 0.75),
    iqr_mm = IQR(needle_length_mm)
  )

```
:::

:::: {.column width="40%"}
::: callout-tip
**📖 Preview**

The box in a boxplot **is** the IQR — its bottom and top edges are the 25th and 75th percentiles you just computed. The boxplots you'll build in Graphing Effectively are these five numbers, drawn.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 2 — Displaying Data, on the boxplot and the five-number summary behind it.
:::
::::
::::::

## Part 5 · The `length()` Trap {.divider}

## The `length()` Trap

::: callout-important
**🔮 Predict first:** A data frame has 5 rows, but 2 values are `NA`. What does `n()` return? What *should* your sample size be for a mean or SE calculation?
:::

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-08}
na_demo_df <- tibble(length_mm = c(4, 3, NA, 7, NA))

na_demo_df |>
  summarize(
    rows    = n(),                     # every row — including NAs!
    missing = sum(is.na(length_mm))    # how many are missing?
  )
```
:::

:::: {.column width="40%"}
::: callout-note
**📖 New words**

- **`n()`** — counts **rows**, whether or not the value is there
- **`is.na(x)`** — `TRUE` where a value is missing
- **`!is.na(x)`** — the flip: `TRUE` where a value is real
- **`sum(!is.na(x))`** — counts the real values = the correct *n*
:::
::::
::::::

## The `length()` Trap — Why It Matters

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-09}
# Using n() as your sample size silently gives the WRONG answer
na_demo_df |>
  summarize(
    wrong_n  = n(),
    right_n  = sum(!is.na(length_mm)),
    sd_mm    = sd(length_mm, na.rm = TRUE),
    wrong_se = round(sd_mm / sqrt(wrong_n), 3),
    right_se = round(sd_mm / sqrt(right_n), 3)
  )
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

R will **not** warn you. `n()` happily returns 5 when only 3 values are real — a silent, common error, and the two SEs above differ because of it.
:::
::::
::::::

## Fix — `sum(!is.na())`

:::::: columns
:::: {.column width="60%"}
```{r descriptive-stats-10}
na_demo_df |>
  mutate(
    is_missing = is.na(length_mm),    # TRUE where NA
    is_real    = !is.na(length_mm)    # flipped: TRUE where real
  )

na_demo_df |>
  summarize(n_real = sum(!is.na(length_mm)))   # count the TRUEs
```

::: callout-tip
**Best practice:** use `sum(!is.na(needle_length_mm))` for *n* every time you compute an SE — never `n()`. Even when you're sure there are no missing values. Field data rarely stays that clean.
:::
::::

::: {.column width="40%"}
**How it works, step by step:**

1.  `is.na(x)` → `TRUE` for each missing value
2.  `!is.na(x)` → flips it to `TRUE` for **real** values
3.  `sum(...)` → adds up the `TRUE`s (`TRUE` = 1)

`mutate()` shows the two logical columns; `summarize()` collapses them to the count.

[📖 R4DS §18 — Missing values](https://r4ds.hadley.nz/missing-values.html)
:::
::::::

## Fix — Applied to Real Data

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-11}
# Applied to our real data — even with no NAs, make it a habit
shady_df |>
  summarize(n = sum(!is.na(needle_length_mm)))

sunny_df |>
  summarize(n = sum(!is.na(needle_length_mm)))
```
:::

:::: {.column width="40%"}
::: callout-tip
Even when you're confident a column has zero missing values, run it through `sum(!is.na())` anyway — it costs nothing and field data rarely stays clean for long.
:::
::::
::::::

## Part 6 · Standard Error {.divider}

## Standard Error

:::::: columns
::: {.column width="60%"}
$$SE = \frac{s}{\sqrt{n}}$$

- Measures how precisely you know the **mean** — not the spread of raw data
- Gets **smaller** as *n* grows: more data → a more precise estimate of the mean

|        | Describes                          |
|--------|------------------------------------|
| **SD** | spread of *individual* data points |
| **SE** | precision of the *sample mean*     |

We report SE whenever we're making a claim about a group mean.
:::

:::: {.column width="40%"}
```{r descriptive-stats-12}
shady_df |>
  summarize(
    n     = sum(!is.na(needle_length_mm)),
    sd_mm = sd(needle_length_mm, na.rm = TRUE),
    se_mm = round(sd_mm / sqrt(n), 2)
  )

sunny_df |>
  summarize(
    n     = sum(!is.na(needle_length_mm)),
    sd_mm = sd(needle_length_mm, na.rm = TRUE),
    se_mm = round(sd_mm / sqrt(n), 2)
  )
```

::: callout-warning
**⚠️ Watch out!**

`sqrt(n())` and `sqrt(sum(!is.na(x)))` give the *same* answer only when there are zero missing values. Don't assume — check.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 4 — Estimating with Uncertainty, is the chapter for the standard error and the sampling distribution of the mean.\
Gotelli & Ellison, Ch. 3 — Summary Statistics, covers the standard error alongside the standard deviation.
:::
::::
::::::

## All Stats, One Group, One `summarize()`

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-13}
shady_df |>
  summarize(
    n       = sum(!is.na(needle_length_mm)),
    mean_mm = round(mean(needle_length_mm, na.rm = TRUE), 2),
    med_mm  = round(median(needle_length_mm, na.rm = TRUE), 2),
    sd_mm   = round(sd(needle_length_mm, na.rm = TRUE), 2),
    se_mm   = round(sd(needle_length_mm, na.rm = TRUE) / sqrt(n), 2)
  )
```
:::

:::: {.column width="40%"}
- Every function takes `na.rm = TRUE`
- One row out, already labeled — and it is a data frame, so it pipes onward into `write_csv()` or `ggplot()`

::: callout-note
**✅ Key idea**

Running this once for `shady_df` and again for `sunny_df` is exactly the repetition `group_by()` + `summarize()` exists to remove.
:::
::::
::::::

## Part 6b · `skimr` — a Fast Full-Dataset Overview {.divider}

## `skimr` — Every Column, One Command

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-14}
#| eval: false
install.packages("skimr")   # once per machine
```

```{r descriptive-stats-15}
library(skimr)

skim(pine_df)
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`skim()` gives you n, missing count, mean, SD, and a tiny inline histogram for **every numeric column** — and counts for every character column — in one call. It's the fastest way to sanity-check a dataset the moment you load it.

[📖 R4DS §10 — Exploratory data analysis](https://r4ds.hadley.nz/EDA.html) is this habit written out at length.
:::
::::
::::::

## `skimr` — Doesn't Replace, Just Speeds Up

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-16}
# Same idea, restricted to numeric columns
pine_df |>
  select(needle_length_mm, needle_width_mm) |>
  skim()
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

- `skim()` is a first look, not a final answer.
- you can pipe groups into it to show that `needle_length_mm` should be split by `side`
  - for real per-group comparisons, you still need `group_by()` + `summarize()`, next.
:::
::::
::::::

## Part 7 · Tidy Group Stats — `group_by()` + `summarize()` {.divider}

## Tidy Group Stats — `group_by()` + `summarize()`

::: callout-note
**🔮 Predict first:** How many **rows** will the result have? (Hint: how many values does `side` take?)
:::

::::: columns
::: {.column width="60%"}
```{r descriptive-stats-17}
side_stats_df <- pine_df |>
  group_by(side) |>
  summarize(
    n       = sum(!is.na(needle_length_mm)),
    mean_mm = round(mean(needle_length_mm, na.rm = TRUE), 2),
    med_mm  = round(median(needle_length_mm, na.rm = TRUE), 2),
    sd_mm   = round(sd(needle_length_mm, na.rm = TRUE), 2),
    se_mm   = round(sd(needle_length_mm, na.rm = TRUE) / sqrt(n), 2)
  )

side_stats_df
```
:::

::: {.column width="40%"}
- `group_by(side)` splits the data by side
- `summarize()` collapses each group to **one row** of stats
- Both groups, one clean table — no copy-paste

::: callout-warning
**⚠️ Round last, not first**

Tempting to write `se_mm = sd_mm / sqrt(n)` and reuse the column above. Don't — `sd_mm` has already been **rounded**, so the SE inherits that error. Compute from the raw values every time; round only what you print.
:::

[📖 R4DS §3.5 — `summarize()`](https://r4ds.hadley.nz/data-transform.html#sec-summarize)
:::
:::::

## Grouping by More Than One Variable

:::::: columns
::: {.column width="60%"}
```{r descriptive-stats-18}
team_stats_df <- pine_df |>
  group_by(team, side) |>
  summarize(
    n       = sum(!is.na(needle_length_mm)),
    mean_mm = round(mean(needle_length_mm, na.rm = TRUE), 2),
    se_mm   = round(sd(needle_length_mm, na.rm = TRUE) / sqrt(n), 2),
    .groups = "drop"
  )

team_stats_df
```
:::

:::: {.column width="40%"}
- `group_by(team, side)` splits by **field team, then side** — one row per combination
- `.groups = "drop"` turns off the grouping once you're done — good habit, avoids surprises later

::: callout-tip
**🖐 Notice**

Four teams × two sides = 8 rows. Does every team show the same shady-vs-sunny pattern, or do some disagree?
:::
::::
::::::

## Part 7c · The *n* You Report Is a Choice {.divider}

## Look again at that `n`

::::::: columns
:::: {.column width="60%"}
Our side summary reported **n = 32 per side**. But where did 32 come from?

```{r descriptive-stats-19}
pine_df |> count(side, team)
```

32 needles per side — but only **4 trees**. We measured 8 needles from each tree and then counted them as 8 independent facts about sun exposure.

::: callout-important
**This is the pseudoreplication from Getting Started, in a summary table.** The `n` in your table is a claim about how much independent evidence you have. Right now that claim is wrong.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Ask of every `n`: **"n of what?"**\
n = 32 needles is true.\
n = 32 *independent measurements of sun exposure* is not.
:::
::::
:::::::

## The two-stage summary — the right way

::::::: columns
:::: {.column width="60%"}
**Stage 1** — collapse each tree to one number:

```{r descriptive-stats-20}
team_means_df <- pine_df |>
  group_by(team, side) |>
  summarize(team_mean_mm = mean(needle_length_mm, na.rm = TRUE),
            .groups = "drop")
```

**Stage 2** — now summarize *those*:

```{r descriptive-stats-21}
side_correct_df <- team_means_df |>
  group_by(side) |>
  summarize(
    n_trees = sum(!is.na(team_mean_mm)),
    mean_mm = round(mean(team_mean_mm), 2),
    se_mm   = round(sd(team_mean_mm) / sqrt(n_trees), 2)
  )

side_correct_df
```
::::

:::: {.column width="40%"}
::: callout-note
**✅ The pattern**

`group_by()` + `summarize()` **twice**: once to get down to the unit you actually replicated, once to describe those units.

You will use this pattern for the rest of the course.
:::
::::
:::::::

## What changed, and what didn't

::::::: columns
:::: {.column width="60%"}
| | naive (n = 32 needles) | correct (n = 4 trees) |
|---|---|---|
| mean, shady | 14.12 | **14.12** |
| mean, sunny | 15.21 | **15.21** |
| SE, shady | 0.49 | **1.01** |
| SE, sunny | 0.46 | **0.79** |

::: callout-important
**The means did not move. The standard errors roughly doubled.**

Pseudoreplication doesn't bias your estimate — it fakes your **precision**. It makes you look twice as certain as your data entitles you to be.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Coming in Week 4**

An SE that is twice as big makes a difference that is much harder to call real. That is exactly what we will see when we run the t-test — and why the **paired** test, which respects the tree structure, is the honest one.
:::
::::
:::::::

## Part 7b · See It — Boxplot with Mean ± SE {.divider}

## Every Number Today Has a Picture

::: callout-note
**🔮 Predict first:** The box shows the middle 50% of 32 needles. The error bar shows the mean ± 1 SE. Which will be **narrower**, and why?
:::

::::::: columns
:::: {.column width="55%"}
```{r descriptive-stats-22}
#| fig-width: 5
#| fig-height: 4
#| output: false
side_plot <- ggplot(pine_df,
       aes(x = side, y = needle_length_mm)) +

  # 1 — spread of all 64 needles
  geom_boxplot(width = 0.5, fill = NA,
               outlier.shape = NA) +

  # 2 — each team's mean, faded, behind
  geom_point(data = team_means_df,
             aes(y = team_mean_mm, color = team),
             position = position_dodge(width = 0.5),
             size = 3, alpha = 0.5) +

  # 3 — mean and standard error, on top
  stat_summary(fun = mean,
               geom = "point", size = 3) +
  stat_summary(fun.data = mean_se,
               geom = "errorbar", width = 0.15) +

  labs(x = "Sun exposure",
       y = "Needle length (mm)",
       color = "Field team") +
  theme_bw()
```

::: callout-tip
**📖 New pattern**

`stat_summary()` computes a statistic **while plotting** — no summary table needed first. `fun = mean` draws the mean; `fun.data = mean_se` draws mean ± 1 SE.
:::
::::

:::: {.column width="45%"}
```{r descriptive-stats-23}
#| fig-width: 5
#| fig-height: 4
#| echo: false
side_plot
```

::: callout-note
**✅ Key idea**

The **box** describes the spread of individual needles. The **error bar** describes how precisely we know the mean. They answer different questions — never read one as the other.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 2 — Displaying Data, and Ch. 4 — Estimating with Uncertainty, on showing a mean with its uncertainty.
:::
::::
:::::::

## But Those Error Bars Are the Naive Ones

::::::: columns
:::: {.column width="55%"}
```{r descriptive-stats-24}
#| fig-width: 5
#| fig-height: 4
#| output: false
honest_plot <- ggplot(team_means_df,
       aes(x = side, y = team_mean_mm)) +

  geom_point(aes(color = team),
             position = position_dodge(width = 0.5),
             size = 3, alpha = 0.5) +

  stat_summary(fun = mean,
               geom = "point", size = 3) +
  stat_summary(fun.data = mean_se,
               geom = "errorbar", width = 0.15) +

  labs(x = "Sun exposure",
       y = "Team mean needle length (mm)",
       color = "Field team") +
  theme_bw()
```

::: callout-warning
**⚠️ Watch out!**

The plot on the previous slide fed `stat_summary()` all **32 needles** per side, so it drew the **n = 32** SE — the pseudoreplicated one. Feed it the **4 team means** instead and the error bars widen to the honest **n = 4**.
:::
::::

:::: {.column width="45%"}
```{r descriptive-stats-25}
#| fig-width: 5
#| fig-height: 4
#| echo: false
honest_plot
```

::: callout-note
**✅ Key idea**

Same two `stat_summary()` layers, different data in — and a very different claim about precision. **What you pipe into the plot is a statistical decision, not a plotting one.**
:::
::::
:::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Computed **mean** and **median**, and know why they can disagree
- Computed **variance** and **SD** for each group with `summarize()`
- Computed **quantiles**, **range**, and **IQR** — and saw the IQR is literally the boxplot's box
- Found the **`n()` trap** and fixed it with `sum(!is.na())`
- Computed **SE** and know how it differs from SD
- Got a fast full-dataset overview with **`skimr::skim()`**
- Used `group_by()` + `summarize()` to get every group's stats in one table
- Learned to ask **"n of what?"**, and used a **two-stage summary** to report the n we actually replicated
- Drew the whole thing: a boxplot with **`stat_summary()`** mean ± SE, and saw the error bars change when the n changes

::: callout-tip
**🖐 Before next class**

Finish the worksheet: compute n, mean, SD, and SE for `needle_length_mm` grouped by `side`, and again grouped by `team`. Save the summary table with `write_csv()`.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Up next — Graphing Effectively**

- Boxplots and jittered points, done right
- Plotting mean ± SE directly with `stat_summary()`
- Faceting, themes, and `ggsave()` for a publication-ready figure
:::
::::
:::::::

## Readings

::::::: columns
:::: {.column width="60%"}
**Before next class** — chapter PDFs are posted on **Canvas**:

- **Whitlock & Schluter**, *The Analysis of Biological Data* — **Ch. 3**, Describing Data (mean, median, variance, SD, quantiles) and **Ch. 4**, Estimating with Uncertainty (the standard error and what it does *not* measure)
- **Gotelli & Ellison**, *A Primer of Ecological Statistics* — **Ch. 3**, Summary Statistics: Measures of Location and Spread

**If you want more:**

- Whitlock & Schluter, Ch. 2 — Displaying Data, for the boxplot and the five-number summary
::::

:::: {.column width="40%"}
::: callout-note
**📖 R for Data Science (2e)**

Free online at [r4ds.hadley.nz](https://r4ds.hadley.nz/):

- [§3.5 — Groups and `summarize()`](https://r4ds.hadley.nz/data-transform.html#sec-summarize)
- [§10 — Exploratory data analysis](https://r4ds.hadley.nz/EDA.html)
- [§13 — Numbers](https://r4ds.hadley.nz/numbers.html) — counts and summary functions
- [§18 — Missing values](https://r4ds.hadley.nz/missing-values.html) — the `n()` trap
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` loaded? Spelling? Did you forget `na.rm = TRUE`?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

[📖 R4DS §8 — Workflow: getting help](https://r4ds.hadley.nz/workflow-help.html)

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
