---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Linear Mixed Models"
subtitle: "Random effects for nested and hierarchical data"
description: "The modern mixed-model (lmer) alternative to a traditional nested ANOVA: afex for unbalanced designs, random-effects notation, F vs. Chi-square tests, diagnostics, post-hoc comparisons, and a by-hand F-ratio calculation for comparison — on the sea urchin grazing / algae cover dataset."
categories: [ANOVA, Mixed Models]
order: 15
week: 10
author: "Bill Perry"

knitr:
  opts_chunk:
    paged.print: false
    collapse: true
format:
  html:
    downloads: [docx, pptx]  # This creates download links for all three
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true

editor:
  markdown:
    wrap: 72
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
library(car)
library(lme4)
library(lmerTest)
library(emmeans)
library(afex)
library(patchwork)
library(janitor)
library(tidyverse)

options(contrasts = c("contr.sum", "contr.poly"))
options(scipen = 999)  # no sci notation
```

## Where We Left Off

:::::: columns
:::: {.column width="60%"}
Covered last time — nested ANOVA, the traditional way:

- Factorial vs. nested designs
- The nested-design linear model and variance partitioning
- Fixed vs. random effects — B nested in A
- The naive (wrong) factorial model vs. `aov(..., Error(...))`

::: callout-note
**✅ Key idea from last lecture**

`patch` nested in `treat` needs the right error term — `MS_{patch(treat)}`,
not `MS_{resid}` — or the F-test for treatment is wrong. Today: the modern
alternative to specifying that error term by hand.
:::
::::

::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

Same sea urchin grazing / algae cover dataset as last time — same
question, same data, a different (and more flexible) way to fit the
model.
:::
:::
::::::

## Part 1 · Setup — The Sea Urchin Data, Again {.divider}

## Data Overview

::::: columns
::: {.column width="50%"}
Same dataset as the nested ANOVA lecture: `patch` (random patches 1–16
where treatments were applied), `treat` (urchin density treatment),
`quad` (replicate quadrats within each treatment-patch combination),
`algae` (percentage cover of filamentous algae — the response
variable).

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_df <- read_csv("data/andrew.csv") %>% clean_names()

u_df <- u_df %>%
  mutate(treat = case_when(
         treat == "removal" ~ "none",
         treat == "control" ~ "high",
         treat == "dens_33" ~ "low",
         treat == "dens_66" ~ "medium",
         TRUE ~ "other")) %>%
  mutate(treat = factor(treat,
                        levels = c("none", "low", "medium", "high"),
                        labels = c("none", "low", "medium", "high")),
         patch = as_factor(patch))

cat("Data\n\n")
head(u_df)
cat("\n\nTreatment levels:\n\n")
levels(u_df$treat)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_summary <- u_df %>%
  group_by(treat) %>%
  summarise(
    n = n(),
    mean = mean(algae),
    sd = sd(algae),
    se = sd / sqrt(n),
    min = min(algae),
    max = max(algae)
  )
cat("Summary statistics by treatment:\n\n")
u_summary
```
:::
:::::

## Part 2 · Modern Approaches — afex and Mixed Models {.divider}

## What to Do If It's an Unbalanced Design

::::: columns
::: {.column width="50%"}
The `afex` package is specifically designed for ANOVA with Type III SS,
and handles nested designs well — even when unbalanced.
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
options(contrasts = c("contr.sum", "contr.poly"))

u_afex_model <- aov_car(algae ~ treat + Error(patch),
                     data = u_df,
                     fun_aggregate = mean)
cat("AFEX nested ANOVA results:\n\n")
summary(u_afex_model)
```
:::
:::::

## The Modern Way — Mixed-Model ANOVA

::::: columns
::: {.column width="50%"}
Fit the model with treatment as a fixed effect, and patch nested within
treatment as a random effect:
`lmer(algae ~ treat + (1|treat:patch), ...)`.

**BOBYQA** (Bound Optimization BY Quadratic Approximation) is an
optimization algorithm used in mixed-effects modeling to find the
parameter values that maximize the likelihood function — useful for
fitting complex nested models.

**Notation we'll get into later:**

- Random intercept, fixed slope — `(1|patch)`
- Random intercept, random slope — `(1|treat:patch)`
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_lmer_model <- lmer(algae ~ treat + (1|treat:patch), data = u_df,
                    control = lmerControl(optimizer = "bobyqa",
                                         optCtrl = list(maxfun = 2e5)))
cat("Mixed model summary:\n\n")
summary(u_lmer_model)
```
:::
:::::

## Mixed-Model ANOVA — Method 1: F-Distribution

::::: columns
::: {.column width="50%"}
- Accounts for the uncertainty in variance-component estimation
- More conservative (higher p-values)
- Better for small samples
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_anova_result <- Anova(u_lmer_model, type = 3, ddf = "Satterthwaite",
                       test.statistic = "F")
cat("Type III ANOVA with F test:\n\n")
u_anova_result
```
:::
:::::

## Mixed-Model ANOVA — Method 2: Chi-Square

::::: columns
::: {.column width="50%"}
- Assumes variance components are known (not estimated)
- More liberal (lower p-values)
- Assumes large samples

**Why different results?** Chi-square = F × numerator df. The F-test
accounts for the denominator df (12, here — reflecting sample size);
Chi-square assumes infinite denominator df (i.e., a large sample).

**Rule of thumb:** under 100 observations or \< 20 random-effect levels,
use the F-test. Over 500 observations and \> 50 random-effect levels,
Chi-square is okay. In between, the F-test is safer.
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_anova_f <- Anova(u_lmer_model, type = 3, test.statistic = "Chisq")
cat("Type III ANOVA with Chi-square test:\n\n")
u_anova_f
```
:::
:::::

## The Nested Model, Restated

::::: columns
::: {.column width="50%"}
$$algae_{ijk} = \mu + \alpha_i + \beta_{j(i)} + \epsilon_{ijk}$$

- $\mu$ is the overall mean
- $\alpha_i$ is the fixed effect of treatment i
- $\beta_{j(i)}$ is the random effect of patch j nested within treatment
  i
- $\epsilon_{ijk}$ is the residual error for quadrat k in patch j within
  treatment i
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
cat("Final ANOVA results:\n\n")
u_anova_result
```
:::
:::::

## Part 3 · Post-Hoc and Interpretation {.divider}

## Interpreting the Variance Components

::: callout-important
The nested ANOVA reveals that there was no significant effect of urchin
density treatment on algae cover (χ² = 8.1513, df = 3, p = 0.04306). The
variance component for patches nested within treatments (294.3)
indicates substantial spatial heterogeneity in algae cover, highlighting
the importance of accounting for this spatial variation in the analysis.
:::

## Post-Hoc Comparisons

::::: columns
::: {.column width="60%"}
Although the main effect of treatment was marginally significant (p =
0.04306), we can examine mean differences between treatments to
understand patterns in the data.

```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_emm <- emmeans(u_lmer_model, ~ treat)
cat("Estimated marginal means:\n\n")
u_emm
```
:::

::: {.column width="40%"}
```{r}
u_pairs <- pairs(u_emm, adjust = "sidak")
cat("Pairwise comparisons (Sidak adjusted):\n\n")
u_pairs
```
:::
:::::

## Compact Letter Display

:::::: columns
::: {.column width="60%"}
```{r}
u_cld <- multcomp::cld(u_emm, alpha = 0.05, Letters = letters)
cat("Compact letter display:\n\n")
u_cld
```
:::

:::: {.column width="40%"}
::: callout-important
**Interpreting treatment comparisons**

Mean algae cover shows a pattern with increasing urchin density:
None/Removed (39.20%) \> Medium/33% Density (19.00%) \> High/66% Density
(21.55%) \> Low/Control (1.30%). This suggests an inverse relationship
between urchin density and algae cover. The high variability among
patches within treatments contributed to the marginal statistical
significance for the treatment effect.
:::
::::
::::::

## Part 4 · Assumptions and Visualization {.divider}

## ANOVA Assumptions Testing — Base R

:::::: columns
::: {.column width="60%"}
For valid inference from ANOVA, several assumptions must be met. We test
these below with a base R approach — note that it doesn't work all that
well for a mixed model.
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

`plot(mixed_model)` and friends were built for `lm()` objects — they
mostly work for `lmer()`, but purpose-built tools like `DHARMa` do a
better job (see the Nested ANOVA activity).
:::
::::
::::::

## Diagnostic Plots

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-width: 3.8
#| fig-height: 2.6
#| fig-alt: "Three diagnostic plots of the mixed-model residuals side by side: a normal Q-Q plot, a histogram of residuals, and a residuals-versus-fitted-values plot with a dashed reference line at zero."
u_fitted <- fitted(u_lmer_model)
u_residuals <- residuals(u_lmer_model)

u_qq_plot <- ggplot(data.frame(residuals = u_residuals), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 7)

u_hist_plot <- ggplot(data.frame(residuals = u_residuals), aes(x = residuals)) +
  geom_histogram(bins = 15, fill = "lightblue", color = "black") +
  labs(title = "Histogram", x = "Residuals", y = "Frequency") +
  theme_minimal(base_size = 7)

u_resid_plot <- ggplot(data.frame(fitted = u_fitted, residuals = u_residuals),
                    aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal(base_size = 7)

u_qq_plot + u_hist_plot + u_resid_plot
```

## Levene's Test for Homogeneity of Variance

:::::: columns
::: {.column width="50%"}
```{r}
u_levene <- leveneTest(algae ~ treat, data = u_df)
cat("Levene's test for homogeneity of variance:\n\n")
u_levene
```
:::

:::: {.column width="50%"}
::: callout-important
**Interpreting the assumption tests**

The Q-Q plot shows some deviation from normality, particularly in the
tails, and Levene's test indicates significant heterogeneity of
variances across treatments (p = 0.00008785). Quinn & Keough (2002)
noted "large differences in within-cell variances" in this dataset, and
transformations (including arcsin) did not improve variance homogeneity.
However, ANOVA is generally robust to heteroscedasticity with balanced
designs, which is why they analyzed untransformed data. The Residuals
vs. fitted plot also shows a pattern of increasing variance with
increasing fitted values, confirming the heteroscedasticity.
:::
::::
::::::

## Visualization — Boxplot

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-width: 3.6
#| fig-height: 3
#| fig-alt: "Boxplot of algae cover percentage by urchin density treatment, with jittered individual quadrat points overlaid, showing algae cover highest at 'none' treatment and lowest at 'low' treatment."
u_boxplot <- ggplot(u_df, aes(x = treat, y = algae, fill = treat)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  scale_fill_viridis_d(option = "D", end = 0.85) +
  labs(title = "Algae Cover by Urchin Density Treatment",
       x = "Treatment", y = "Algae Cover (%)") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
u_boxplot
```

## Visualization — Means Plot

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-width: 3.4
#| fig-height: 2.8
#| fig-alt: "Mean algae cover by treatment plotted as points with standard error error bars, connected to show the pattern across urchin density treatments."
u_means_plot <- ggplot(u_summary, aes(x = treat, y = mean, group = 1)) +
  geom_point(size = 3, shape = 21, fill = "white") +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(title = "Mean Algae Cover by Treatment",
       x = "Treatment", y = "Mean Algae Cover (%)") +
  theme_minimal(base_size = 8)
u_means_plot
```

## Part 5 · Discussion and Manual Calculation {.divider}

## Scientific Interpretation

::: callout-important
Our mixed-model analysis revealed substantial spatial heterogeneity in
algae cover, with significant variation among patches within each
treatment. The effect of urchin density treatments on filamentous algae
cover was marginally significant at the α = 0.05 level (p = 0.043).
Descriptive statistics show algae cover increasing as urchin density
decreases, with None/Removed showing the highest cover (39.20%),
followed by Medium/33% Density (19.00%), High/66% Density (21.55%), and
Low/Control showing minimal algae cover (1.30%). The variance component
for patches nested within treatments (294.31, \~39.5% of total variance)
shows that spatial heterogeneity is a major structuring force in these
algal communities, and needs to be accounted for when designing and
analyzing ecological field experiments.
:::

## We Can Do This Manually — but Ugh

::::: columns
::: {.column width="50%"}
Nobody would ever do this by hand in practice — but seeing it once makes
the "which MS goes on the bottom of the F-ratio" logic concrete, and
shows how the traditional nested-ANOVA F-ratio (last lecture) connects
to the mixed-model result above.

```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_man_model <- aov(algae ~ treat + treat:patch, data = u_df)
u_anova_summary <- summary(u_man_model)[[1]]
cat("Standard ANOVA table:\n\n")
u_anova_summary
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_MS_treat <- u_anova_summary[1, "Mean Sq"]
u_MS_patch <- u_anova_summary[2, "Mean Sq"]
u_MS_residual <- u_anova_summary[3, "Mean Sq"]

u_df_treat <- u_anova_summary[1, "Df"]
u_df_patch <- u_anova_summary[2, "Df"]
u_df_residual <- u_anova_summary[3, "Df"]

cat("MS Treatment:", u_MS_treat, "\n")
cat("MS Patch(Treatment):", u_MS_patch, "\n")
cat("MS Residual:", u_MS_residual, "\n\n")
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_F_treat_correct <- u_MS_treat / u_MS_patch        # Treatment tested against patches
u_F_patch <- u_MS_patch / u_MS_residual             # Patches tested against residual

cat("F Treatment:", u_F_treat_correct, "\n")
cat("F Patch(Treatment):", u_F_patch, "\n\n")
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_p_treat_correct <- pf(u_F_treat_correct, u_df_treat, u_df_patch, lower.tail = FALSE)
u_p_patch <- pf(u_F_patch, u_df_patch, u_df_residual, lower.tail = FALSE)

u_corrected_table <- data.frame(
  Source = c("Treatment", "Patches(Treatment)", "Residual"),
  Df = c(u_df_treat, u_df_patch, u_df_residual),
  MS = round(c(u_MS_treat, u_MS_patch, u_MS_residual), 1),
  F = c(round(u_F_treat_correct, 2), round(u_F_patch, 2), NA),
  p = c(ifelse(u_p_treat_correct < 0.001, "<0.001", round(u_p_treat_correct, 3)),
        ifelse(u_p_patch < 0.001, "<0.001", round(u_p_patch, 3)),
        NA)
)
cat("Corrected ANOVA table:\n\n")
u_corrected_table
```
:::
:::::
