---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Linear Mixed Models"
subtitle: "Random effects for nested and hierarchical data"
description: "The modern mixed-model (lmer) alternative to a traditional nested ANOVA: afex for unbalanced designs, random-effects notation, F vs. Chi-square tests, diagnostics, post-hoc comparisons, and a by-hand F-ratio calculation for comparison — on the sea urchin grazing / algae cover dataset."
categories: [ANOVA, Mixed Models]
order: 15
week: 10
author: "Bill Perry"

knitr:
  opts_chunk:
    paged.print: false
    collapse: true
format:
  html:
    downloads: [docx, pptx]  # This creates download links for all three
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true

editor:
  markdown:
    wrap: 72
---

```{r linear-mixed-models-01}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
library(car)
library(lme4)
library(lmerTest)
library(emmeans)
library(afex)
library(patchwork)
library(janitor)
library(tidyverse)

options(contrasts = c("contr.sum", "contr.poly"))
options(scipen = 999)  # no sci notation
```

## Where We Left Off

:::::: columns
:::: {.column width="60%"}
Covered last time — nested ANOVA, the traditional way:

- Factorial vs. nested designs
- The nested-design linear model and variance partitioning
- Fixed vs. random effects — B nested in A
- The naive (wrong) factorial model vs. `aov(..., Error(...))`

::: callout-note
**✅ Key idea from last lecture**

`patch` nested in `treat` needs the right error term — `MS_{patch(treat)}`,
not `MS_{resid}` — or the F-test for treatment is wrong. Today: the modern
alternative to specifying that error term by hand.
:::
::::

::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

Same sea urchin grazing / algae cover dataset as last time — same
question, same data, a different (and more flexible) way to fit the
model.
:::
:::
::::::

## Part 1 · Setup — The Sea Urchin Data, Again {.divider}

## Data Overview

::::: columns
::: {.column width="50%"}
Same dataset as the nested ANOVA lecture: `patch` (random patches 1–16
where treatments were applied), `treat` (urchin density treatment),
`quad` (replicate quadrats within each treatment-patch combination),
`algae` (percentage cover of filamentous algae — the response
variable).

```{r linear-mixed-models-02}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_df <- read_csv("data/andrew.csv") %>% clean_names()

u_df <- u_df %>%
  # Map to Quinn & Keough's Table 9.1 densities. NOTE the file's dens_33 /
  # dens_66 labels are the reverse of the book's percentages -- verified by
  # matching patch means: dens_33 is the book's 66% group, dens_66 its 33%.
  mutate(treat = case_when(
         treat == "removal" ~ "0%",     # book  0% density, mean algae 39.2
         treat == "dens_66" ~ "33%",    # book 33% density, mean algae 19.0
         treat == "dens_33" ~ "66%",    # book 66% density, mean algae 21.6
         treat == "control" ~ "100%",   # book 100% density, mean algae 1.3
         TRUE ~ "other")) %>%
  mutate(treat = factor(treat,
                        levels = c("0%", "33%", "66%", "100%"),
                        labels = c("0%", "33%", "66%", "100%")),
         patch = as_factor(patch))

cat("Data\n\n")
head(u_df)
cat("\n\nTreatment levels:\n\n")
levels(u_df$treat)
```
:::

::: {.column width="50%"}
```{r linear-mixed-models-03}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_summary <- u_df %>%
  group_by(treat) %>%
  summarise(
    n = n(),
    mean = mean(algae),
    sd = sd(algae),
    se = sd / sqrt(n),
    min = min(algae),
    max = max(algae)
  )
cat("Summary statistics by treatment:\n\n")
u_summary
```
:::
:::::

## Part 2 · Modern Approaches — afex and Mixed Models {.divider}

## What to Do If It's an Unbalanced Design

::::: columns
::: {.column width="50%"}
The `afex` package is specifically designed for ANOVA with Type III SS,
and handles nested designs well — even when unbalanced.
:::

::: {.column width="50%"}
```{r linear-mixed-models-04}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
options(contrasts = c("contr.sum", "contr.poly"))

u_afex_model <- aov_car(algae ~ treat + Error(patch),
                     data = u_df,
                     fun_aggregate = mean)
cat("AFEX nested ANOVA results:\n\n")
summary(u_afex_model)
```
:::
:::::

## What Makes an Effect "Random"?

::::::: columns
:::: {.column width="60%"}
Ask one question: **do I care about these specific levels, or are they a sample of levels I could have had?**

| | `treat` | `patch` |
|------------------|------------------|------------------|
| The levels are | 0/33/66/100% density | patches 1–16 |
| Chosen by | me, deliberately | whatever was there |
| Would a repeat study use the same ones? | **yes** | no — different patches |
| I want to estimate | **each level's mean** | **how much they vary** |
| So it is | **fixed** | **random** |

::: callout-important
**The key consequence**

A fixed factor with 16 levels costs you **15 parameters**. A random factor with 16 levels costs you **one** — a single variance, σ²_patch.

That is why mixed models cope with designs that `aov()` chokes on: they are estimating far less.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 A useful test**

If you can write a sentence a reader would care about that names a specific level — *"the 100% density treatment reduced algae"* — it is **fixed**.

If the sentence would be *"patch 7 happened to be lush"* — nobody cares about patch 7 in particular. **Random.**
:::

::: callout-warning
**⚠️ Rule of thumb: you need enough levels**

Estimating a variance from **fewer than ~5 levels** is unreliable. With 2 or 3 groups, fit them as fixed effects instead.

We have 16 patches — comfortably enough.
:::
::::
:::::::

## The Modern Way — Mixed-Model ANOVA

::::: columns
::: {.column width="50%"}
Fit the model with treatment as a fixed effect, and patch nested within
treatment as a random effect:
`lmer(algae ~ treat + (1|patch), ...)`.

**BOBYQA** (Bound Optimization BY Quadratic Approximation) is an
optimization algorithm used in mixed-effects modeling to find the
parameter values that maximize the likelihood function — useful for
fitting complex nested models.

**Reading the random-effects formula:**

`(` what varies `|` what it varies across `)`

- **`(1|patch)`** — the `1` means the **intercept**. Each patch gets its own baseline algae cover.
- **`(1|treat:patch)`** — the same thing, just naming the group as the *treatment-patch combination*. Still intercept-only.
- **`(1 + depth|patch)`** — *this* is a random **slope**: each patch gets its own intercept **and** its own slope for `depth`.

::: callout-warning
**⚠️ `(1|treat:patch)` is not a random slope**

Anything of the form `(1|...)` is **random intercepts only** — the `1` *is* the intercept. What changes between `(1|patch)` and `(1|treat:patch)` is the **grouping variable**, not what varies within it.

Here they are the identical model, because patch IDs are already unique (1–16) and each patch sits in one treatment. We'll use `(1|patch)`.
:::
:::

::: {.column width="50%"}
```{r linear-mixed-models-05}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_lmer_model <- lmer(algae ~ treat + (1|patch), data = u_df,
                    control = lmerControl(optimizer = "bobyqa",
                                         optCtrl = list(maxfun = 2e5)))
cat("Mixed model summary:\n\n")
summary(u_lmer_model)
```
:::
:::::

## What the Random Effect Buys You: ICC

::::::: columns
:::: {.column width="60%"}
The model splits the leftover variation in two:

```{r linear-mixed-models-04b}
#| echo: true
#| collapse: true
vc <- as.data.frame(VarCorr(u_lmer_model))
var_patch <- vc$vcov[vc$grp != "Residual"]
var_resid <- vc$vcov[vc$grp == "Residual"]

cat("among patches :", round(var_patch, 1), "\n")
cat("within patches:", round(var_resid, 1), "\n")
cat("ICC =", round(var_patch / (var_patch + var_resid), 3), "\n")
```

The **intraclass correlation** is the share of variance living *between* groups — equivalently, **how strongly two quadrats in the same patch resemble each other**.

::: callout-important
**ICC = 0.50 is the whole reason we are here**

Two quadrats from the same patch are **half-correlated**. They are not two independent observations — they are closer to one and a half.

An ICC near 0 would mean patches barely matter and ordinary ANOVA would have been fine. An ICC of 0.5 means ignoring patch would have been badly wrong — and last week's naive model showed exactly how wrong (F = 16.1 instead of 2.7).
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Same number, three ways**

You have now seen 294.3 arrive by three routes:

1. By hand from mean squares, `(MS_patch − MS_within)/n`
2. From `aov(Error(patch))`
3. From `VarCorr()` on the mixed model

They agree because **they are the same model**, written three ways.
:::
::::
:::::::

## Partial Pooling — What Actually Happens to Each Patch

::::::: columns
:::: {.column width="60%"}
A random effect does not just adjust the p-value. It changes each patch's **estimate**.

```{r linear-mixed-models-04c}
#| echo: false
#| fig-width: 3.6
#| fig-height: 2.8
#| message: false
#| fig-alt: "Slope plot comparing each patch's raw observed mean algae cover to its shrunken mixed-model estimate, with grey lines joining the two; estimates move toward their treatment mean, and patches with more extreme raw means move furthest."
raw_means <- u_df %>%
  group_by(treat, patch) %>%
  summarise(raw = mean(algae), .groups = "drop")

raw_means$fitted <- predict(
  u_lmer_model,
  newdata = raw_means[, c("treat", "patch")])

raw_means %>%
  pivot_longer(c(raw, fitted),
               names_to = "kind", values_to = "algae") %>%
  mutate(kind = factor(kind, c("raw", "fitted"),
                       c("raw patch mean", "model estimate"))) %>%
  ggplot(aes(kind, algae, group = patch, color = treat)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  labs(x = NULL, y = "Algae cover (%)", color = "Treatment") +
  theme_minimal(base_size = 8)
```
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word: shrinkage**

Every patch estimate is pulled **toward its treatment mean**. The model treats an extreme patch average as partly real signal and partly luck, and splits the difference.

Patches with fewer quadrats, or more internal scatter, get pulled further.
:::

::: callout-tip
**🖐 Why this is a feature**

A patch whose 5 quadrats happened to land high is probably *not* that extreme — the next 5 would come back closer to average. Shrinkage builds that expectation in, instead of taking every noisy group mean at face value.

This is called **partial pooling**: somewhere between treating every patch as unrelated and pretending they are all identical.
:::
::::
:::::::

## Mixed-Model ANOVA — Method 1: F-Distribution

::::: columns
::: {.column width="50%"}
- Accounts for the uncertainty in variance-component estimation
- More conservative (higher p-values)
- Better for small samples
:::

::: {.column width="50%"}
```{r linear-mixed-models-06}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
# car::Anova uses Kenward-Roger denominator df for test.statistic = "F"
# (that is what the printed header says -- there is no ddf argument here)
u_anova_result <- Anova(u_lmer_model, type = 3, test.statistic = "F")
cat("Type III ANOVA with F test:\n\n")
u_anova_result
```
:::
:::::

## Mixed-Model ANOVA — Method 2: Chi-Square

::::: columns
::: {.column width="50%"}
- Assumes variance components are known (not estimated)
- More liberal (lower p-values)
- Assumes large samples

**Why different results?** Chi-square = F × numerator df. The F-test
accounts for the denominator df (12, here — reflecting sample size);
Chi-square assumes infinite denominator df (i.e., a large sample).

**Rule of thumb:** under 100 observations or \< 20 random-effect levels,
use the F-test. Over 500 observations and \> 50 random-effect levels,
Chi-square is okay. In between, the F-test is safer.
:::

::: {.column width="50%"}
```{r linear-mixed-models-07}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_anova_chisq <- Anova(u_lmer_model, type = 3, test.statistic = "Chisq")
cat("Type III ANOVA with Chi-square test:\n\n")
u_anova_chisq
```
:::
:::::

## The Nested Model, Restated

::::: columns
::: {.column width="50%"}
$$algae_{ijk} = \mu + \alpha_i + \beta_{j(i)} + \epsilon_{ijk}$$

- $\mu$ is the overall mean
- $\alpha_i$ is the fixed effect of treatment i
- $\beta_{j(i)}$ is the random effect of patch j nested within treatment
  i
- $\epsilon_{ijk}$ is the residual error for quadrat k in patch j within
  treatment i
:::

::: {.column width="50%"}
```{r linear-mixed-models-08}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
cat("Final ANOVA results:\n\n")
u_anova_result
```
:::
:::::

## Part 3 · Post-Hoc and Interpretation {.divider}

## Interpreting the Variance Components

::: callout-important
**⚠️ Two tests, two answers — and you must pick one *before* you look**

| Test | Statistic | p | Verdict at α = 0.05 |
|-----------|-----------|-----------|-----------|
| F (Kenward-Roger df) | F₃,₁₂ = 2.72 | **0.091** | not significant |
| Wald χ² | χ² = 8.15 | **0.043** | significant |

Same model. Same data. **Opposite conclusions.**

The χ² test assumes it knows the variance components exactly and that the sample is large. We have **16 patches**. It is being over-confident, exactly as the rule of thumb two slides back predicted.

**Report the F-test.** It agrees with last week's `aov(Error(patch))` result to four decimal places — because it is the same test.
:::

::: callout-note
**How to report this result**

"Algal cover did not differ significantly among urchin density treatments (linear mixed model with patch as a random effect, **F₃,₁₂ = 2.72, p = 0.091**, Kenward-Roger denominator df). Among-patch variance was substantial (**σ²_patch = 294.3** vs **σ²_residual = 298.6**, ICC = 0.50), indicating strong spatial heterogeneity in algal cover."
:::

## Post-Hoc Comparisons

::::: columns
::: {.column width="60%"}
Although the main effect of treatment was marginally significant (p =
0.04306), we can examine mean differences between treatments to
understand patterns in the data.

```{r linear-mixed-models-09}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_emm <- emmeans(u_lmer_model, ~ treat)
cat("Estimated marginal means:\n\n")
u_emm
```
:::

::: {.column width="40%"}
```{r linear-mixed-models-10}
u_pairs <- pairs(u_emm, adjust = "sidak")
cat("Pairwise comparisons (Sidak adjusted):\n\n")
u_pairs
```
:::
:::::

## Compact Letter Display

:::::: columns
::: {.column width="60%"}
```{r linear-mixed-models-11}
u_cld <- multcomp::cld(u_emm, alpha = 0.05, Letters = letters)
cat("Compact letter display:\n\n")
u_cld
```
:::

:::: {.column width="40%"}
::: callout-important
**Interpreting treatment comparisons**

Mean algae cover shows a pattern with increasing urchin density:
None/Removed (39.20%) \> Medium/33% Density (19.00%) \> High/66% Density
(21.55%) \> Low/Control (1.30%). This suggests an inverse relationship
between urchin density and algae cover. The high variability among
patches within treatments contributed to the marginal statistical
significance for the treatment effect.
:::
::::
::::::

## Part 4 · Assumptions and Visualization {.divider}

## ANOVA Assumptions Testing — Base R

:::::: columns
::: {.column width="60%"}
For valid inference from ANOVA, several assumptions must be met. We test
these below with a base R approach — note that it doesn't work all that
well for a mixed model.
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

`plot(mixed_model)` and friends were built for `lm()` objects — they
mostly work for `lmer()`, but purpose-built tools like `DHARMa` do a
better job (see the Nested ANOVA activity).
:::
::::
::::::

## Diagnostic Plots

```{r linear-mixed-models-12}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-width: 3.8
#| fig-height: 2.6
#| fig-alt: "Three diagnostic plots of the mixed-model residuals side by side: a normal Q-Q plot, a histogram of residuals, and a residuals-versus-fitted-values plot with a dashed reference line at zero."
u_fitted <- fitted(u_lmer_model)
u_residuals <- residuals(u_lmer_model)

u_qq_plot <- ggplot(data.frame(residuals = u_residuals), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 7)

u_hist_plot <- ggplot(data.frame(residuals = u_residuals), aes(x = residuals)) +
  geom_histogram(bins = 15, fill = "lightblue", color = "black") +
  labs(title = "Histogram", x = "Residuals", y = "Frequency") +
  theme_minimal(base_size = 7)

u_resid_plot <- ggplot(data.frame(fitted = u_fitted, residuals = u_residuals),
                    aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted", x = "Fitted Values", y = "Residuals") +
  theme_minimal(base_size = 7)

u_qq_plot + u_hist_plot + u_resid_plot
```

## Levene's Test for Homogeneity of Variance

:::::: columns
::: {.column width="50%"}
```{r linear-mixed-models-13}
u_levene <- leveneTest(algae ~ treat, data = u_df)
cat("Levene's test for homogeneity of variance:\n\n")
u_levene
```
:::

:::: {.column width="50%"}
::: callout-important
**Interpreting the assumption tests**

The Q-Q plot shows some deviation from normality, particularly in the
tails, and Levene's test indicates significant heterogeneity of
variances across treatments (p = 0.00008785). Quinn & Keough (2002)
noted "large differences in within-cell variances" in this dataset, and
transformations (including arcsin) did not improve variance homogeneity.
However, ANOVA is generally robust to heteroscedasticity with balanced
designs, which is why they analyzed untransformed data. The Residuals
vs. fitted plot also shows a pattern of increasing variance with
increasing fitted values, confirming the heteroscedasticity.
:::
::::
::::::

## Visualization — Boxplot

```{r linear-mixed-models-14}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-width: 3.6
#| fig-height: 3
#| fig-alt: "Boxplot of algae cover percentage by urchin density treatment, with jittered individual quadrat points overlaid, showing algae cover highest at 'none' treatment and lowest at 'low' treatment."
u_boxplot <- ggplot(u_df, aes(x = treat, y = algae, fill = treat)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  scale_fill_viridis_d(option = "D", end = 0.85) +
  labs(title = "Algae Cover by Urchin Density Treatment",
       x = "Treatment", y = "Algae Cover (%)") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
u_boxplot
```

## Visualization — Means Plot

```{r linear-mixed-models-15}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-width: 3.4
#| fig-height: 2.8
#| fig-alt: "Mean algae cover by treatment plotted as points with standard error error bars, connected to show the pattern across urchin density treatments."
u_means_plot <- ggplot(u_summary, aes(x = treat, y = mean, group = 1)) +
  geom_point(size = 3, shape = 21, fill = "white") +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(title = "Mean Algae Cover by Treatment",
       x = "Treatment", y = "Mean Algae Cover (%)") +
  theme_minimal(base_size = 8)
u_means_plot
```

## Part 5 · Discussion and Manual Calculation {.divider}

## Scientific Interpretation

::: callout-important
Our mixed-model analysis revealed substantial spatial heterogeneity in
algae cover, with significant variation among patches within each
treatment. The effect of urchin density treatments on filamentous algae
cover was marginally significant at the α = 0.05 level (p = 0.043).
Descriptive statistics show algae cover increasing as urchin density
decreases, with None/Removed showing the highest cover (39.20%),
followed by Medium/33% Density (19.00%), High/66% Density (21.55%), and
Low/Control showing minimal algae cover (1.30%). The variance component
for patches nested within treatments (294.31, \~39.5% of total variance)
shows that spatial heterogeneity is a major structuring force in these
algal communities, and needs to be accounted for when designing and
analyzing ecological field experiments.
:::

## We Can Do This Manually — but Ugh

::::: columns
::: {.column width="50%"}
Nobody would ever do this by hand in practice — but seeing it once makes
the "which MS goes on the bottom of the F-ratio" logic concrete, and
shows how the traditional nested-ANOVA F-ratio (last lecture) connects
to the mixed-model result above.

```{r linear-mixed-models-16}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_man_model <- aov(algae ~ treat + treat:patch, data = u_df)
u_anova_summary <- summary(u_man_model)[[1]]
cat("Standard ANOVA table:\n\n")
u_anova_summary
```

```{r linear-mixed-models-17}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_MS_treat <- u_anova_summary[1, "Mean Sq"]
u_MS_patch <- u_anova_summary[2, "Mean Sq"]
u_MS_residual <- u_anova_summary[3, "Mean Sq"]

u_df_treat <- u_anova_summary[1, "Df"]
u_df_patch <- u_anova_summary[2, "Df"]
u_df_residual <- u_anova_summary[3, "Df"]

cat("MS Treatment:", u_MS_treat, "\n")
cat("MS Patch(Treatment):", u_MS_patch, "\n")
cat("MS Residual:", u_MS_residual, "\n\n")
```
:::

::: {.column width="50%"}
```{r linear-mixed-models-18}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_F_treat_correct <- u_MS_treat / u_MS_patch        # Treatment tested against patches
u_F_patch <- u_MS_patch / u_MS_residual             # Patches tested against residual

cat("F Treatment:", u_F_treat_correct, "\n")
cat("F Patch(Treatment):", u_F_patch, "\n\n")
```

```{r linear-mixed-models-19}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
u_p_treat_correct <- pf(u_F_treat_correct, u_df_treat, u_df_patch, lower.tail = FALSE)
u_p_patch <- pf(u_F_patch, u_df_patch, u_df_residual, lower.tail = FALSE)

u_corrected_table <- data.frame(
  Source = c("Treatment", "Patches(Treatment)", "Residual"),
  Df = c(u_df_treat, u_df_patch, u_df_residual),
  MS = round(c(u_MS_treat, u_MS_patch, u_MS_residual), 1),
  F = c(round(u_F_treat_correct, 2), round(u_F_patch, 2), NA),
  p = c(ifelse(u_p_treat_correct < 0.001, "<0.001", round(u_p_treat_correct, 3)),
        ifelse(u_p_patch < 0.001, "<0.001", round(u_p_patch, 3)),
        NA)
)
cat("Corrected ANOVA table:\n\n")
u_corrected_table
```
:::
:::::

## The `lmer` Formula Cheat-Sheet

::::::: columns
:::: {.column width="60%"}
Read every random term as **`( what varies | across what )`**.

| Formula | Meaning |
|------------------|------------------------------------|
| `y ~ x + (1|g)` | random **intercept** per group — each group has its own baseline |
| `y ~ x + (1 + x|g)` | random intercept **and slope** — each group has its own baseline *and* its own response to x |
| `y ~ x + (0 + x|g)` | random slope only, shared intercept |
| `y ~ x + (1|a) + (1|b)` | two **crossed** random factors |
| `y ~ x + (1|a/b)` | b **nested** in a — shorthand for `(1|a) + (1|a:b)` |

::: callout-important
**⚠️ The confusion to avoid**

`(1|patch)` and `(1|treat:patch)` are **both random-intercept** models. The `1` is the intercept; what differs is only the **grouping**.

You only have a random slope when something other than `1` appears on the left of the `|`.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Do I need `a/b` here?**

Only if the inner IDs **restart** inside each outer level (patch 1,2,3,4 in *every* treatment). Then R cannot tell nesting from crossing and you must write `(1|treat/patch)`.

Our patches are numbered 1–16 across the whole study, so `(1|patch)` already says it.
:::

::: callout-warning
**⚠️ "boundary (singular) fit"**

Means a variance was estimated as ~0 — the grouping factor explains nothing, or you asked for more random structure than the data can support. Simplify the random effects; do not ignore it.
:::
::::
:::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Decided **fixed vs random** by asking whether you care about *these* levels or about *how much they vary*
- Learned that a random factor costs **one parameter** (a variance) where a fixed one costs *k* − 1
- Read `(1|group)` correctly, and saw that it is **not** a random slope
- Got the **ICC = 0.50** — two quadrats in a patch are half-correlated, which is exactly why this design needed a mixed model
- Saw **partial pooling** move each patch estimate toward its treatment mean
- Chose between the **F-test (p = 0.091)** and the **Wald χ² (p = 0.043)** — and learned why the F-test is the right one at n = 16 groups

::: callout-important
**⚠️ Report the F-test, and say so**

With few random-effect levels the χ² is anti-conservative — it assumes it knows the variance components exactly. Always state which approximation you used and what the denominator df were.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ The one idea to keep**

A mixed model does not "remove" the patch effect — it **estimates** it, and then tests the treatment against the right amount of noise.

**σ²_patch = 294.3 arrived three ways:** by hand from mean squares, from `aov(Error(patch))`, and from `VarCorr()`. Same model, three notations. If you understood the nested ANOVA, you already understood mixed models — this is just a better way to write and fit them.
:::

::: callout-tip
**🖐 Before next class**

Take the pine needle data, **collapsed to one value per team per side** (the two-stage summary from Week 3):

``` r
p_df <- pine_df |>
  group_by(team, side) |>
  summarise(needle_length_mm = mean(needle_length_mm),
            .groups = "drop")

anova(lmer(needle_length_mm ~ side + (1|team), data = p_df))
```

Compare `F` to the **square of** the paired-t statistic from Week 4 (t = −2.72). You should get **F = 7.41, df 1 and 3, p = 0.072** — the paired t-test *is* this mixed model.

**🚀 If you finish early:** run the same formula on the raw 64 needles instead. You get p = 0.057 and **59** denominator df, not 3. What changed? (Hint: check `VarCorr()` — one variance is estimated as exactly zero, and R warns "singular fit".)
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: is the grouping variable a **factor**? Is `lmerTest` loaded (plain `lme4` gives no p-values on purpose)? Are you calling `car::Anova()` on the model rather than `anova()`?
3.  **"boundary (singular) fit"** → your random structure is too complex for the data, or that variance really is zero
4.  **"Model failed to converge"** → try `control = lmerControl(optimizer = "bobyqa")`, as we did here, or simplify the random effects
5.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
