---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Multiple Regression"
subtitle: "Several predictors"
description: "Multiple regression on Loyn's Victorian forest-bird dataset: partial regression slopes, the ANOVA partition, assumptions and diagnostics, collinearity, interactions, categorical predictors as factors with a chosen reference level, model comparison with AIC, and ranking predictor importance."
categories: [Regression]
order: 11
week: 7
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]  # This creates download links for all three
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true
---

```{r multiple-regression-01}
#| include: false
#| message: false
#| warning: false
library(plotly)
library(knitr)
library(patchwork)
library(car)
library(tidyverse)

# ---- Loyn's Victorian forest-bird patches, read once and reused all lecture ----
bird_df <- read_csv("data/loyn_birds.csv", show_col_types = FALSE) %>%
  mutate(
    log_area        = log10(area_ha),
    log_dist        = log10(dist_m),
    log_dist_larger = log10(dist_larger_m)
  )

# ---- numbers quoted in the text below, computed once so the prose cannot drift ----
alone <- function(v) {
  m <- lm(reformulate(v, "abundance"), data = bird_df)
  c(slope = unname(coef(m)[2]), p = summary(m)$coefficients[2, 4])
}
a_area  <- alone("log_area")
a_graze <- alone("grazing")
a_year  <- alone("year_isolated")
a_alt   <- alone("altitude_m")

full_model <- lm(abundance ~ log_area + log_dist + log_dist_larger +
                   year_isolated + grazing + altitude_m, data = bird_df)
full_sum <- summary(full_model)
b_area_full  <- full_sum$coefficients["log_area", 1]
b_graze_full <- full_sum$coefficients["grazing", 1]
p_graze_full <- full_sum$coefficients["grazing", 4]

reduced_model <- lm(abundance ~ log_area + grazing, data = bird_df)
red_sum <- summary(reduced_model)
b_graze_red <- red_sum$coefficients["grazing", 1]

anova_red <- Anova(reduced_model, type = "II")
ss_res  <- sum(residuals(reduced_model)^2)
ss_tot  <- sum((bird_df$abundance - mean(bird_df$abundance))^2)
ss_reg  <- ss_tot - ss_res

std_model <- lm(abundance ~ log_area + grazing,
                data = mutate(bird_df, across(c(abundance, log_area, grazing),
                                              ~ as.numeric(scale(.)))))

raw_area_model <- lm(abundance ~ area_ha, data = bird_df)
log_area_model <- lm(abundance ~ log_area, data = bird_df)

# small p-values must print as plain text inline, not knitr's LaTeX sci-notation
fmt_p <- function(p, digits = 3) {
  if (p < 0.001) formatC(p, format = "e", digits = 2) else formatC(p, format = "g", digits = digits)
}
```

## Where We Left Off

::::: columns
:::: {.column width="60%"}
Covered last time:

- One continuous predictor, one continuous response
- The **sums-of-squares partition**: $SS_{total} = SS_{regression} + SS_{residual}$
- Mean squares, the F-ratio, and **t² = F**
- The four assumptions, all checked on the **residuals**
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea from last lecture**

A simple regression has one predictor and one partition of variance. Today we add more predictors.

The logic barely changes. The **bookkeeping** does — and so does what each slope means.
:::
::::
:::::

## Part 1 · Why Multiple Regression? {.divider}

## Today's Objectives

::::: columns
:::: {.column width="60%"}
The multiple linear regression model:

- Partial regression slopes — and why they are not the same as simple slopes
- Analysis of variance and the null hypotheses
- Assumptions, diagnostics and collinearity
- Interactions
- **Factors, dummy variables, and choosing your reference level**
- Model comparison and predictor importance

[📥 Bird data](data/loyn_birds.csv){.btn .btn-primary download="loyn_birds.csv"} [📥 Script skeleton (.R)](data/11_multiple_regression.R){.btn .btn-primary download="11_multiple_regression.R"}
::::

:::: {.column width="40%"}
::: callout-note
**📖 Reading**

Quinn & Keough, Ch. 6 — Multiple and Complex Regression.

Whitlock & Schluter, Ch. 17; Gotelli & Ellison, Ch. 9.
:::
::::
:::::

## The Study · Birds in Fragmented Forest

::::: columns
:::: {.column width="52%"}
Loyn (1987) surveyed **`r nrow(bird_df)` remnant forest patches** in Victoria, Australia, and counted the birds in each one. For every patch he also recorded six things that might explain the count.

Small patches, isolated patches, recently cut-off patches and heavily grazed patches are all candidates — and they are not independent of each other.
::::

:::: {.column width="48%"}
::: {.compact-table}
| Column | What it is |
|:---|:---|
| `abundance` | birds per patch (the response) |
| `area_ha` | patch area |
| `dist_m` | distance to nearest patch |
| `dist_larger_m` | distance to nearest *larger* patch |
| `year_isolated` | year the patch was cut off |
| `grazing` | grazing intensity, 1–5 |
| `altitude_m` | elevation |
:::
::::
:::::

## Reading the Data In

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-02}
bird_df <- read_csv("data/loyn_birds.csv") %>%
  mutate(log_area        = log10(area_ha),
         log_dist        = log10(dist_m),
         log_dist_larger = log10(dist_larger_m))

glimpse(bird_df)
```
::::

:::: {.column width="45%"}
The three area and distance columns are **logged on the way in**.

That is not a stylistic choice — Part 4 shows what the raw scale does to the residuals, and it is the same failure you met in last week's homework.

::: callout-note
**`grazing_class`**

The same grazing intensity as a **word** rather than a number. Part 7 is about what changes when you use it.
:::
::::
:::::

## When Do We Need More Than One Predictor?

::::: columns
:::: {.column width="55%"}
Bird abundance in a patch plausibly depends on **all seven** of those things at once, and they are not independent of each other — small patches tend to be grazed harder, and isolated patches tend to be older.

Fitting one predictor at a time cannot answer "does area matter *once grazing is accounted for*."

- Predictors all continuous → **multiple regression**
- A mix of continuous and categorical → still multiple regression, with dummy variables
::::

:::: {.column width="45%"}
|  | Continuous X | Categorical X |
|:---|:---|:---|
| **Continuous Y** | Regression | ANOVA |
| **Categorical Y** | Logistic regression | Contingency tables |
::::
:::::

## One Predictor at a Time

::::: columns
:::: {.column width="55%"}
Start the way you would have last week — fit each predictor on its own.

```{r multiple-regression-03}
#| echo: false
#| message: false
one_at_a_time <- tibble(
  predictor = c("log_area", "log_dist", "log_dist_larger",
                "year_isolated", "grazing", "altitude_m")
) %>%
  rowwise() %>%
  mutate(
    slope = coef(lm(reformulate(predictor, "abundance"), bird_df))[2],
    p     = summary(lm(reformulate(predictor, "abundance"), bird_df))$coefficients[2, 4]
  ) %>%
  ungroup() %>%
  mutate(across(c(slope, p), ~ signif(.x, 3)))

one_at_a_time
```
::::

:::: {.column width="45%"}
**Four of the six are significant on their own**, two of them overwhelmingly.

Grazing looks like the strongest story in the dataset: every extra step of grazing intensity costs about **`r abs(round(a_graze["slope"], 1))` birds**, at p = `r fmt_p(a_graze["p"])`.

::: callout-warning
**⚠️ Hold that thought**

Every one of these fits ignores the other five predictors entirely.
:::
::::
:::::

## 🗳️ DECIDE · Which Predictors Will Survive?

::::: columns
:::: {.column width="60%"}
We are about to put all six predictors into one model together.

**What do you expect to happen?**

**(a)** All four significant predictors stay significant — they were each real

**(b)** Roughly the same four stay, but with smaller slopes

**(c)** Only one or two survive, and at least one strong predictor becomes non-significant

**(d)** Nothing survives — with six predictors and 56 patches there is not enough data

::: callout-warning
**✍️ Write your letter down, then vote.**
:::
::::

:::: {.column width="40%"}
::: callout-note
**The question to ask**

*How much of what grazing "explains" on its own is really being explained by patch area?*

Small patches are grazed harder. Look back at the table — is grazing telling you something area was not already telling you?
:::
::::
:::::

## From a Line to a (Hyper)plane

::::: columns
:::: {.column width="35%"}
With one predictor we fit a **line** through a scatterplot.

With two, we fit a **plane** through a cloud — here, bird abundance over log patch area and grazing intensity.

With three or more it is a **hyperplane**, which you cannot draw but can still fit.
::::

:::: {.column width="65%"}
```{r multiple-regression-04}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| fig-alt: "Fitted regression plane for bird abundance over log patch area and grazing intensity: an interactive 3D scatterplot with a fitted mesh surface in HTML output, or a 2D heat map with contour lines and observed patches in docx and pptx output. Abundance rises with patch area and falls with grazing."
plane_model <- lm(abundance ~ log_area + grazing, data = bird_df)

grid <- expand.grid(
  log_area = seq(min(bird_df$log_area), max(bird_df$log_area), length.out = 30),
  grazing  = seq(min(bird_df$grazing),  max(bird_df$grazing),  length.out = 30)
)
grid$abundance <- predict(plane_model, newdata = grid)

if (knitr::is_html_output()) {
  plot_ly() %>%
    add_trace(data = bird_df, x = ~log_area, y = ~grazing, z = ~abundance,
              type = "scatter3d", mode = "markers",
              marker = list(size = 4, color = "red"), name = "Patches") %>%
    add_trace(data = grid, x = ~log_area, y = ~grazing, z = ~abundance,
              type = "mesh3d", opacity = 0.5, name = "Fitted plane") %>%
    layout(scene = list(xaxis = list(title = "log10 area"),
                        yaxis = list(title = "Grazing"),
                        zaxis = list(title = "Abundance")))
} else {
  ggplot() +
    geom_tile(data = grid, aes(log_area, grazing, fill = abundance)) +
    geom_contour(data = grid, aes(log_area, grazing, z = abundance),
                 colour = "white", alpha = 0.5) +
    geom_point(data = bird_df, aes(log_area, grazing), colour = "red", size = 1.5) +
    scale_fill_viridis_c(name = "Fitted\nabundance") +
    labs(x = "log10 patch area (ha)", y = "Grazing intensity") +
    theme_minimal(base_size = 8)
}
```
::::
:::::

## Part 2 · The Multiple Regression Model {.divider}

## The Multiple Regression Model

::::: columns
:::: {.column width="55%"}
For i = 1 to n observations and p predictors:

$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_i$$

- $y_i$ — the response for observation i
- $\beta_0$ — the intercept, the mean of Y when **every** predictor is 0
- $\beta_1 \dots \beta_p$ — the **partial** regression slopes
- $\epsilon_i$ — the error, estimated by the residual $y_i - \hat{y}_i$
::::

:::: {.column width="45%"}
::: callout-note
**✅ Key idea — what "partial" means**

A partial slope is the change in Y per one-unit change in that predictor **while every other predictor is held constant**.

It is a *conditional* statement. Change what else is in the model and you have changed the condition — so the number is allowed to change too.
:::
::::
:::::

## The Full Model · The Answer to the Vote

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-05}
full_model <- lm(abundance ~ log_area + log_dist +
                   log_dist_larger + year_isolated +
                   grazing + altitude_m,
                 data = bird_df)

summary(full_model)
```
::::

:::: {.column width="45%"}
::: callout-important
**✅ The answer is (c).**

Four predictors were significant alone. In the model together, **only `log_area` survives** (p < 0.001).

Grazing has gone from a slope of **`r round(a_graze["slope"], 2)`** (p = `r fmt_p(a_graze["p"])`) to **`r round(b_graze_full, 2)`** (p = `r fmt_p(p_graze_full)`) — most of what grazing appeared to explain was patch area wearing a different hat.

Year of isolation and altitude have gone the same way.
:::
::::
:::::

## Why Not Just Regress Y on Each X Separately?

::::: columns
:::: {.column width="55%"}
Each simple regression asks *"does this predictor track abundance?"* The multiple regression asks *"does it track abundance **among patches that are otherwise alike**?"*

Those are different questions, and they can have different answers.

::: {.compact-table}
| Predictor | Alone | In the full model |
|:---|---:|---:|
| log area | `r round(a_area["slope"], 2)` | **`r round(b_area_full, 2)`** |
| grazing | `r round(a_graze["slope"], 2)` | **`r round(b_graze_full, 2)`** |
| year isolated | `r round(a_year["slope"], 3)` | **`r round(coef(full_model)["year_isolated"], 3)`** |
| altitude | `r round(a_alt["slope"], 3)` | **`r round(coef(full_model)["altitude_m"], 3)`** |
:::
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ Neither number is "wrong"**

If you want to predict bird abundance from grazing alone, `r round(a_graze["slope"], 2)` is the right slope to use.

If you want to claim grazing has an effect **beyond** the fact that grazed patches are small, `r round(b_graze_full, 2)` is the number you have to defend — and here you cannot.
:::
::::
:::::

## Estimation and Prediction

::::: columns
:::: {.column width="55%"}
**Estimation.** Ordinary least squares again: choose the coefficients that minimise $\sum(y_i - \hat{y}_i)^2$. With several predictors this is a set of simultaneous equations solved with matrix algebra — `lm()` does it instantly, and what matters is knowing *what* it minimises.

**Prediction.** Substitute new values for every predictor.

- More observations → narrower intervals
- More predictors → wider intervals
- Stay inside the range of **every** predictor
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ Extrapolation is worse here**

In simple regression you only have to stay inside one range. Here you have to stay inside the **data cloud** — and a combination can be far outside it even when each predictor separately looks ordinary.

There are no 1000-ha patches under intense grazing in these data. The model will still happily predict one.
:::
::::
:::::

## Part 3 · ANOVA for Multiple Regression {.divider}

## Partitioning Variance — Sums of Squares

::::: columns
:::: {.column width="55%"}
Exactly the partition from last week, with p predictors instead of one:

$$SS_{total} = SS_{regression} + SS_{residual}$$

- $SS_{regression}$ — variation the model explains
- $SS_{residual}$ — variation it does not
- Regression df is now **p**, not 1
::::

:::: {.column width="45%"}
::: {.compact-table}
| Source | SS | df | MS |
|:---|:---|:---|:---|
| Regression | $\sum (\hat{y}_i - \bar{y})^2$ | $p$ | SS/df |
| Residual | $\sum (y_i - \hat{y}_i)^2$ | $n-p-1$ | SS/df |
| Total | $\sum (y_i - \bar{y})^2$ | $n-1$ | |
:::

Mean squares are variances, so they do not grow with n — which is why the F-ratio is built from them and not from the sums of squares directly.
::::
:::::

## Hypotheses — the "Basic" Test

::::: columns
:::: {.column width="55%"}
The **overall F-test** at the bottom of `summary()` tests one null:

$$H_0: \beta_1 = \beta_2 = \dots = \beta_p = 0$$

- If it is true, $MS_{regression}$ and $MS_{residual}$ estimate the same variance and F ≈ 1
- If any β is non-zero, $MS_{regression}$ is inflated and F > 1

It tells you the model as a whole explains something. It does **not** tell you which predictor did the work.
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ A significant overall F with nothing significant inside it**

This happens, and it is the classic signature of collinearity: the predictors jointly explain a lot, but they are so tangled that no individual slope can be pinned down.
:::
::::
:::::

## Hypotheses — Testing One Predictor

::::: columns
:::: {.column width="55%"}
To ask whether one specific β is zero, fit the model **with** and **without** it and compare:

$$SS_{extra} = SS_{regression}^{full} - SS_{regression}^{reduced}$$

$$F_{1,\,n-p-1} = \frac{MS_{extra}}{MS_{residual}^{full}}$$

```{r multiple-regression-06}
reduced_model <- lm(abundance ~ log_area + grazing,
                    data = bird_df)

anova(reduced_model, full_model)
```
::::

:::: {.column width="45%"}
Dropping **four** predictors at once — distance, distance to a larger patch, year of isolation and altitude — costs nothing detectable (p = `r fmt_p(anova(reduced_model, full_model)[2, "Pr(>F)"])`).

::: callout-tip
**🖐 You rarely run this by hand**

The t-test beside each coefficient in `summary()` answers the same question for a single predictor. Use `anova()` when you want to drop a **group** of predictors at once.
:::
::::
:::::

## Explained Variance

::::: columns
:::: {.column width="55%"}
$$R^2 = \frac{SS_{regression}}{SS_{total}} = 1 - \frac{SS_{residual}}{SS_{total}}$$

For the reduced model: $\dfrac{`r round(ss_reg, 0)`}{`r round(ss_tot, 0)`}$ = **`r round(red_sum$r.squared, 3)`**

The six-predictor model manages `r round(full_sum$r.squared, 3)` — higher, as it must be.
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ Watch out**

- R² **always** rises when you add a predictor, even a useless one
- So R² cannot be used to compare models with different numbers of predictors
- R² from models with differently transformed Y are not comparable at all
:::

That is why adjusted R² and AIC exist, and we come back to them in Part 8.
::::
:::::

## Part 4 · Assumptions and Diagnostics {.divider}

## First: Does a Predictor Need a Transform?

::::: columns
:::: {.column width="55%"}
Patch area runs from **`r min(bird_df$area_ha)` ha to `r format(max(bird_df$area_ha), big.mark = ",")` ha** — four orders of magnitude. Fitted raw, it behaves exactly like last week's homework:

::: {.compact-table}
| | raw `area_ha` | `log10(area_ha)` |
|:---|---:|---:|
| slope | `r round(coef(raw_area_model)[2], 4)` | `r round(coef(log_area_model)[2], 2)` |
| R² | `r round(summary(raw_area_model)$r.squared, 3)` | `r round(summary(log_area_model)$r.squared, 3)` |
| Shapiro on residuals | `r fmt_p(shapiro.test(residuals(raw_area_model))$p.value)` | `r fmt_p(shapiro.test(residuals(log_area_model))$p.value)` |
:::

Every model in this lecture uses `log_area`, and the diagnostics are why.
::::

:::: {.column width="45%"}
```{r multiple-regression-07}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3.4
#| fig-width: 3.4
#| fig-alt: "Two stacked scatterplots of bird abundance against patch area: on the raw scale the points pile against the left axis, and on a log10 scale they spread into a straight-line relationship."
p_raw <- ggplot(bird_df, aes(area_ha, abundance)) +
  geom_point(size = 1.2) + geom_smooth(method = "lm", se = FALSE, colour = "red") +
  labs(title = "Raw area", x = NULL, y = NULL) + theme_minimal(base_size = 8)

p_log <- ggplot(bird_df, aes(log_area, abundance)) +
  geom_point(size = 1.2) + geom_smooth(method = "lm", se = FALSE, colour = "blue") +
  labs(title = "log10 area", x = NULL, y = NULL) + theme_minimal(base_size = 8)

p_raw / p_log
```
::::
:::::

## Assumptions and Diagnostics — Outliers

::::: columns
:::: {.column width="55%"}
- Fixed X values — unrealistic in field data, but the working assumption
- **No overly influential points** — check leverage and influence with **Cook's distance**
- A point can be extreme in Y, extreme in X, or extreme in the *combination*; only the third kind is uniquely a multiple-regression problem

Largest Cook's distance here is **`r round(max(cooks.distance(reduced_model)), 2)`** — nothing near the usual threshold of 1.
::::

:::: {.column width="45%"}
```{r multiple-regression-08}
#| fig-height: 3
#| fig-width: 3.2
#| fig-alt: "Cook's distance plot for the reduced bird-abundance model, showing each patch's influence as a vertical bar."
plot(reduced_model, which = 4)
```
::::
:::::

## Assumptions and Diagnostics — Residuals

::::: columns
:::: {.column width="55%"}
Normality, equal variance and independence — all read off the **residuals**, exactly as last week.

```{r multiple-regression-09}
shapiro.test(residuals(reduced_model))
```

Distribution and variance problems are usually fixed by transforming Y, or a predictor, before anything more exotic.
::::

:::: {.column width="45%"}
```{r multiple-regression-10}
#| echo: false
#| fig-width: 3.2
#| fig-height: 3.2
#| fig-alt: "Two stacked diagnostic plots for the reduced bird-abundance model: residuals versus fitted values, and a normal QQ plot of the residuals."
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(reduced_model, which = 1)
plot(reduced_model, which = 2)
par(mfrow = c(1, 1))
```
::::
:::::

## Assumptions — Sample Size and Added-Variable Plots

::::: columns
:::: {.column width="55%"}
- **More observations than predictors** — a common rule of thumb is **10 observations per predictor**. Six predictors would want 60; we have `r nrow(bird_df)`, so the full model is already at the edge
- **Predictors not collinear** — Part 5
- **Each X linear in Y after accounting for the others** — this is what an **added-variable plot** shows, and it is the multiple-regression version of "plot your data first"

``` r
avPlots(reduced_model)
```
::::

:::: {.column width="45%"}
```{r multiple-regression-11}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 3.4
#| fig-alt: "Added-variable plots for the reduced bird-abundance model, one for log area and one for grazing, each showing the relationship with abundance after removing the other predictor's effect."
avPlots(reduced_model, ask = FALSE, main = "")
```

Each panel is Y and one X **after** the other predictor's effect has been removed from both.
::::
:::::

## Part 5 · Collinearity {.divider}

## Collinearity — the Problem

::::: columns
:::: {.column width="55%"}
Predictors in field data are almost never independent. Here, **small patches are grazed harder** — r = `r round(cor(bird_df$log_area, bird_df$grazing), 2)`.

When predictors overlap:

- Parameter estimates become **unstable** — one changed value can move them a lot
- Standard errors **inflate**, so slopes lose significance
- The model can fit well while no single slope is trustworthy
::::

:::: {.column width="45%"}
```{r multiple-regression-12}
#| echo: false
bird_df %>%
  select(log_area, log_dist, log_dist_larger,
         year_isolated, grazing, altitude_m) %>%
  cor() %>%
  round(2)
```
::::
:::::

## Collinearity — Detecting It

::::: columns
:::: {.column width="55%"}
**Variance Inflation Factors:**

$$VIF_{X_j} = \frac{1}{1 - R^2_j}$$

where $R^2_j$ comes from regressing **$X_j$ on all the *other* predictors** — not from a pairwise correlation. (With two predictors they coincide; with more they do not.)

- Rule of thumb: **VIF > 10 is bad**; some authors use > 5
- Simplest fix: drop predictors that are measuring the same thing
::::

:::: {.column width="45%"}
```{r multiple-regression-13}
vif(full_model)
```

::: callout-note
**✅ Key idea**

A high VIF does not mean the model is wrong. It means you cannot trust the **individual** partial slopes, even when the model as a whole fits well.
:::
::::
:::::

## 🗳️ DECIDE · The Slope That Changed Its Mind

::::: columns
:::: {.column width="60%"}
Grazing went from **`r round(a_graze["slope"], 2)`** on its own to **`r round(b_graze_full, 2)`** in the full model, and lost its significance. The VIFs are all under 3.

**What is the right thing to say about grazing?**

**(a)** Grazing has no effect on bird abundance

**(b)** The effect of grazing is real but small — the full model estimates it better

**(c)** Grazing and patch area explain overlapping variation, and these data cannot cleanly separate them

**(d)** The full model is broken by collinearity and should be discarded

::: callout-warning
**✍️ One letter. Write it down, then argue it with your neighbour.**
:::
::::

:::: {.column width="40%"}
::: callout-note
**The question to ask**

*Is this a statement about the world, or a statement about what this dataset can resolve?*
:::
::::
:::::

## What the Data Can and Cannot Separate

::::: columns
:::: {.column width="55%"}
::: callout-important
**✅ The answer is (c).**

Grazing and area are correlated at r = `r round(cor(bird_df$log_area, bird_df$grazing), 2)`. The heavily grazed patches in this dataset are *also* the small ones, so no analysis can tell you which of the two is doing the work.

**(a)** overstates it — a non-significant partial slope is not evidence of no effect.
**(b)** is the common mistake: the full model is not "better", it answers a *different, conditional* question.
**(d)** is wrong too — VIFs under 3 are mild, and the model predicts perfectly well. It just cannot apportion credit.
:::
::::

:::: {.column width="45%"}
::: callout-tip
**🖐 What would settle it**

Find, or create, patches that break the correlation: **large heavily grazed** patches and **small lightly grazed** ones.

That is a design problem, not an analysis problem — which is where this course started.
:::
::::
:::::

## Part 6 · Interactions {.divider}

## Interactions — Additive vs. Multiplicative

::::: columns
:::: {.column width="55%"}
So far every model has been **additive** — the effect of area is the same at every grazing level.

An **interaction** says the effect of one predictor *depends on* the level of another:

$$y_i = \beta_0 + \beta_1X_{1} + \beta_2X_{2} + \epsilon_i$$
$$y_i = \beta_0 + \beta_1X_{1} + \beta_2X_{2} + \beta_3X_{1}X_{2} + \epsilon_i$$

An interaction term is just **another predictor**: the product of two others, with its own coefficient.
::::

:::: {.column width="45%"}
::: callout-note
**✅ Key idea**

Additive: parallel lines.

Interactive: lines with different slopes — the fitted surface is no longer flat.
:::
::::
:::::

## Interactions — Fitting One

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-14}
interaction_model <- lm(abundance ~ log_area * grazing,
                        data = bird_df)

summary(interaction_model)$coefficients
```
::::

:::: {.column width="45%"}
The interaction is **significant** (p = `r fmt_p(summary(lm(abundance ~ log_area * grazing, data = bird_df))$coefficients[4, 4])`), and `a * b` expands to `a + b + a:b` automatically.

**What it means:** patch area matters **more** where grazing is heavy. In lightly grazed country a small patch still holds birds; under heavy grazing only the big patches do.

::: callout-warning
**⚠️ You cannot read the main effects the same way any more**

With an interaction in the model, the `log_area` coefficient is the slope **when grazing = 0** — a value that does not exist in these data.
:::
::::
:::::

## Interactions — the Cost

::::: columns
:::: {.column width="55%"}
- **Parameter proliferation.** Every predictor you add roughly doubles the number of possible terms — six predictors with all interactions is 64 terms, and you have `r nrow(bird_df)` patches
- **Interpretation gets harder**, as the slide before shows
- **Include an interaction when it makes biological sense**, not because it might be significant

"Does fragmentation hurt more where grazing is heavy" is a real hypothesis. That is why this one belongs.
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ Watch out**

Test enough interactions and one will reach p < 0.05 by chance. With six predictors there are 15 two-way interactions alone — at α = 0.05 you expect roughly one false positive just from testing them all.
:::
::::
:::::

## Part 7 · Factors and Dummy Variables {.divider}

## Categorical Predictors Become Dummy Variables

::::: columns
:::: {.column width="55%"}
So far grazing has been a **number**, 1 to 5, which forces every step to have the same effect. That is a strong assumption — is the jump from light to moderate really the same size as moderate to heavy?

Treat it as a **category** instead and the model estimates each level separately. A categorical predictor with k levels becomes **k − 1 dummy variables**, each 0 or 1.

One level gets no dummy at all. That is the **reference level**, and every other coefficient is a comparison *against it*.
::::

:::: {.column width="45%"}
```{r multiple-regression-15}
count(bird_df, grazing, grazing_class)
```
::::
:::::

## R Picks the Reference Level for You — Alphabetically

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-16}
bird_df <- bird_df %>%
  mutate(grazing_class = factor(grazing_class))

levels(bird_df$grazing_class)
```

A character column becomes a factor in **alphabetical** order, so the reference level here is **heavy** — the fourth of five grazing intensities, chosen by spelling.
::::

:::: {.column width="45%"}
```{r multiple-regression-17}
#| echo: false
lm(abundance ~ log_area + grazing_class, data = bird_df) %>%
  summary() %>%
  coef() %>%
  round(2)
```

Every grazing coefficient is now "compared with **heavy** grazing," which is a comparison nobody asked for.
::::
:::::

## 🗳️ DECIDE · Which Level Should Be the Reference?

::::: columns
:::: {.column width="60%"}
The five grazing classes are `minimal`, `light`, `moderate`, `heavy`, `intense`.

**Which should be the reference level, and why?**

**(a)** `heavy` — leave R's default alone; the choice does not affect the model

**(b)** `intense` — the most extreme level makes the biggest contrasts

**(c)** `minimal` — the least-disturbed level, so every coefficient reads as the cost of adding grazing

**(d)** It does not matter, because R² and the F-test are identical either way

::: callout-warning
**✍️ Careful — two of these are true statements. Only one is the right decision.**
:::
::::

:::: {.column width="40%"}
::: callout-note
**The question to ask**

*Which comparison do I want to read straight off the output?*
:::
::::
:::::

## Choosing the Reference Yourself · `fct_relevel()`

::::: columns
:::: {.column width="55%"}
::: callout-important
**✅ The answer is (c).**

(d) is *true* — the fit, R², F and the ANOVA table are all identical whichever level is the reference. But it is not a reason to accept a bad default: the coefficients are what people read, and they should answer the question you asked.

`minimal` grazing is the closest thing this observational study has to a **control**, so putting it first makes every other coefficient "birds lost relative to the least-grazed patches."
:::

```{r multiple-regression-18}
bird_df <- bird_df %>%
  mutate(grazing_class = fct_relevel(
    grazing_class,
    "minimal", "light", "moderate", "heavy", "intense"))

levels(bird_df$grazing_class)
```
::::

:::: {.column width="45%"}
::: callout-tip
**🖐 The habit to build**

Any time a factor enters a model, set its reference level **on purpose**, in the same line where you create the factor.

`fct_relevel()` moves the levels you name to the front and leaves the rest in place, so naming just the control is enough:

``` r
fct_relevel(treatment, "control")
```
:::
::::
:::::

## Same Model, Same F, a Readable Story

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-19}
grazing_model <- lm(abundance ~ log_area + grazing_class,
                    data = bird_df)

summary(grazing_model)$coefficients %>% round(2)
```
::::

:::: {.column width="45%"}
Now every coefficient is a comparison with **minimal** grazing, and the pattern is obvious:

- light, moderate and heavy are all **indistinguishable** from minimal
- `intense` costs about **`r abs(round(coef(grazing_model)["grazing_classintense"], 1))` birds**, p < 0.001

**Grazing is a threshold, not a gradient** — and fitting it as the number 1–5 hid that completely, averaging the effect into `r round(b_graze_red, 2)` birds per step.
::::
:::::

## Why This Matters Even More in ANOVA

::::: columns
:::: {.column width="55%"}
With only continuous predictors there is no reference level at all — every slope is "per unit of X" and nothing is being compared with anything.

The moment a **factor** enters the model, every coefficient becomes a **contrast against the reference**, and that is the whole structure of the next five weeks:

- **One-way ANOVA** — treatment means against a control
- **Two-way ANOVA** — main effects read at the other factor's reference level
- **ANCOVA** — separate intercepts, all relative to one group
::::

:::: {.column width="45%"}
::: callout-important
**✅ The rule to carry forward**

`factor()` gives you alphabetical order. Alphabetical order is almost never the order your experiment was designed in.

**Set the control as the first level, every time**, and your model output reads as "what each treatment did to the control" without any further work.
:::
::::
:::::

## Part 8 · Comparing Models {.divider}

## Comparing Models — the Problem

::::: columns
:::: {.column width="55%"}
- Which predictors should be in the model?
- **Occam's razor:** the best model balances fit against complexity
- **Overfitting:** a high R² bought with useless predictors — which is why R² alone cannot choose between models

Compare **nested** models: one whose predictors are a subset of the other's.
::::

:::: {.column width="45%"}
::: callout-note
**📖 Reference**

Quinn & Keough, Ch. 6, works through exactly this dataset. Whitlock & Schluter, Ch. 17, covers model comparison for regression.
:::
::::
:::::

## Comparing Models — Penalizing Complexity

::::: columns
:::: {.column width="55%"}
Two ways to charge a model for its predictors:

$$\text{Adjusted } R^2 = 1 - \frac{SS_{Residual}/(n - p - 1)}{SS_{Total}/(n - 1)}$$

$$\text{AIC} = n\ln(SS_{Residual}) + 2(p + 1) - n\ln(n)$$

- **Higher** adjusted R² is better
- **Lower** AIC is better
- AIC is comparative only — the number on its own means nothing, and models must be fitted to **the same rows**
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ The NA trap**

If one predictor has missing values, `lm()` silently drops those rows. Two models fitted to different numbers of rows **cannot** be compared by AIC, and R will not warn you.
:::
::::
:::::

## Comparing Models — How to Search

::::: columns
:::: {.column width="55%"}
- Fit all possible models and compare — fine with a few predictors, hopeless with many
- **Automated stepwise** selection adds or removes terms by partial F or AIC
- We use **manual backward selection**: start full, drop what earns nothing, stop when dropping hurts
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ Stepwise selection is controversial**

It inflates R², biases p-values, and knows nothing about biology. A small set of models you can *justify* beats an automated search over all of them.
:::
::::
:::::

## Comparing Models — Putting It Together

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-20}
model_comparison <- tibble(
  model = c("all six predictors",
            "log_area + grazing",
            "log_area only",
            "grazing only"),
  adj_r2 = c(summary(full_model)$adj.r.squared,
             summary(reduced_model)$adj.r.squared,
             summary(lm(abundance ~ log_area, bird_df))$adj.r.squared,
             summary(lm(abundance ~ grazing, bird_df))$adj.r.squared),
  AIC = c(AIC(full_model),
          AIC(reduced_model),
          AIC(lm(abundance ~ log_area, bird_df)),
          AIC(lm(abundance ~ grazing, bird_df)))
) %>%
  mutate(delta_AIC = AIC - min(AIC)) %>%
  arrange(AIC)

model_comparison
```
::::

:::: {.column width="45%"}
::: callout-important
**✅ The two-predictor model wins**

It has the **lowest AIC** and essentially the same adjusted R² as the full model — with four fewer predictors.

A **ΔAIC above about 2** is usually treated as a real difference, so the full model's extra terms are not paying for themselves.
:::

The `anova()` test in Part 3 said the same thing. Two routes, one answer.
::::
:::::

## Part 9 · Relative Importance of Predictors {.divider}

## How Important Is Each Predictor?

::::: columns
:::: {.column width="55%"}
Three standard approaches, and they answer slightly different questions:

1. **F- or t-tests** on the partial slopes
2. **Coefficient of partial determination** — the variance a predictor adds
3. **Standardized partial slopes** — every predictor in SD units

None of them is "the" right answer.
::::

:::: {.column width="45%"}
::: callout-warning
**⚠️ Why the raw slopes cannot be compared**

`log_area` has a slope of `r round(coef(reduced_model)["log_area"], 2)` and `grazing` `r round(coef(reduced_model)["grazing"], 2)`.

One is "birds per tenfold increase in area," the other is "birds per step of grazing intensity." Comparing those two numbers is comparing metres with kilograms.
:::
::::
:::::

## Approaches 1 and 2 — Partial F and Partial R²

::::: columns
:::: {.column width="55%"}
```{r multiple-regression-21}
Anova(reduced_model, type = "II")
```

**Partial η²** = $\dfrac{SS_{predictor}}{SS_{predictor} + SS_{residual}}$ — variance explained *after* controlling for the other predictor.
::::

:::: {.column width="45%"}
::: {.compact-table}
| | log_area | grazing |
|:---|---:|---:|
| partial η² | `r round(anova_red[1,"Sum Sq"]/(anova_red[1,"Sum Sq"]+ss_res), 3)` | `r round(anova_red[2,"Sum Sq"]/(anova_red[2,"Sum Sq"]+ss_res), 3)` |
| ΔR² | `r round(anova_red[1,"Sum Sq"]/ss_tot, 3)` | `r round(anova_red[2,"Sum Sq"]/ss_tot, 3)` |
:::

**ΔR²** is what the model's R² would lose if you dropped that predictor — its *unique* contribution.

Both say area is the stronger of the two.
::::
:::::

## Approach 3 — Standardized Partial Slopes

::::: columns
:::: {.column width="55%"}
Rescale every variable to mean 0 and SD 1, and the slopes become directly comparable: "SDs of Y per SD of X."

```{r multiple-regression-22}
bird_z <- bird_df %>%
  mutate(across(c(abundance, log_area, grazing),
                ~ as.numeric(scale(.))))

lm(abundance ~ log_area + grazing, data = bird_z) %>%
  coef() %>% round(3)
```
::::

:::: {.column width="45%"}
**log_area `r round(coef(std_model)[2], 2)`** versus **grazing `r round(coef(std_model)[3], 2)`** — area is about `r round(abs(coef(std_model)[2]/coef(std_model)[3]), 1)`× the influence of grazing per standard deviation.

::: callout-note
**✅ All three agree here**

Partial F, partial R² and standardized slopes all rank area above grazing. When they *disagree*, that is worth reporting — it usually means the predictors are tangled.
:::
::::
:::::

## Wrap-up

Today you:

- Extended the regression line to a **plane**, and fitted `r nrow(bird_df)` forest patches against six predictors at once
- Watched four predictors that were significant alone collapse to **one** in the model together
- Read a **partial slope** as a conditional statement, and saw that neither it nor the simple slope is "wrong"
- Tested groups of predictors with `anova(reduced, full)` and single ones with the t-tests in `summary()`
- Diagnosed **collinearity** with `vif()`, and learned to say "these data cannot separate them" instead of "it has no effect"
- Fitted a real **interaction** — fragmentation costs more where grazing is heavy
- Turned a categorical predictor into **dummy variables**, and set the reference level on purpose with `fct_relevel()`
- Compared models with **adjusted R²** and **AIC**, and ranked predictors three different ways

## The Three Things to Carry Out of Here

::::: columns
:::: {.column width="50%"}
::: callout-important
**⚠️ R² always rises**

Adding predictors can only increase plain R². Adjusted R² and AIC exist precisely because "explains more variance" is not the same as "is a better model."
:::

::: callout-note
**✅ A partial slope is conditional**

"What does area do among patches grazed the same amount?" Change what else is in the model and you have changed the question.
:::
::::

:::: {.column width="50%"}
::: callout-important
**✅ Set your reference level**

`factor()` sorts alphabetically. Your control is almost never first alphabetically — use `fct_relevel()` and make the output say what you meant.

This is the habit that makes every ANOVA from Week 8 onward readable.
:::

::: callout-note
**Up next — Analysis of Variance**

What happens when **every** predictor is categorical, and why ANOVA and regression are the same machinery in different clothes.
:::
::::
:::::

## Readings

- **Quinn, G. P., & Keough, M. J.** (2002). *Experimental Design and Data Analysis for Biologists*, Ch. 6 — Multiple and Complex Regression. This lecture follows their Box 6-2 analysis of the Loyn data.
- **Whitlock & Schluter**, Ch. 17 — Regression.
- **Gotelli & Ellison**, Ch. 9 — Regression.
- Loyn, R. H. (1987). Effects of patch area and habitat on bird abundances, species numbers and tree health in fragmented Victorian forests. In *Nature Conservation: The Role of Remnants of Native Vegetation*. Surrey Beatty & Sons.

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: is it `y ~ x1 + x2` (additive) or `y ~ x1 * x2` (interaction)? Is `car` loaded for `vif()` and `Anova()`? Are you comparing models fitted to the **same rows** — an `NA` in one predictor silently drops that row
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
