---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Wrangling Your Data"
subtitle: "Importing more than one way, reshaping columns, and saving what you make"
description: "Reading CSV and Excel files, the filter/select/mutate/arrange verbs, the full boolean-operator toolkit (==, !=, >, <, is.na(), %in%, & vs. |), recoding and binning with case_when(), log and unit-conversion transforms, catching duplicate rows, and writing your cleaned data back out — all on the pine needle dataset."
categories: [Wrangling, R, tidyverse]
order: 2
week: 1
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true
---

## Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Turned a field observation into a testable **H₀ / Hₐ**
- Built a project folder (`data/`, `scripts/`, `figures/`, `output/`)
- Loaded `pine_data.csv` and made your first `ggplot`
- Saw why 64 rows is not 64 independent observations

::: callout-note
**✅ Key idea from Getting Started**

You can get data *into* R and make a plot. Today we learn to **reshape** what you have and **save** what you make — the two skills every later module depends on.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  Importing data more than one way
2.  The five wrangling verbs
3.  `filter()` in depth — every boolean operator, `%in%`, AND vs. OR
4.  `mutate()` — unit conversions and log transforms
5.  `case_when()` — recoding and binning
6.  Catching duplicate rows
7.  Saving your work with `write_csv()`
:::
::::
:::::::

## Part 1 · Importing Data More Than One Way {.divider}

## The same data, two file formats

::::::: columns
:::: {.column width="60%"}
Field data doesn't always arrive as a `.csv`. Excel workbooks are everywhere in ecology — you need both tools.

[📥 Download the data](data/pine_data.csv){.btn .btn-primary download="pine_data.csv"} [📥 The messy version](data/pine_needles_error.csv){.btn .btn-secondary download="pine_needles_error.csv"}

```{r wrangling-data-01}
# install.packages("readxl")
# install.packages("janitor")
library(tidyverse)
library(readxl)   # reading Excel files
library(janitor)  # cleans messy column names, Excel artifacts, rounding

pine_csv_df   <- read_csv("data/pine_data.csv")
pine_excel_df <- read_excel("data/pine_data.xlsx")
```

::: callout-note
**✅ Key idea**

`read_csv()` comes from `readr` (loaded with `tidyverse`). \
Excel files need the separate `readxl` package \
— install once, `library()` every session.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word**

**Package vs. bundle** — `tidyverse` is a bundle of packages (`readr`, `dplyr`, `ggplot2`, …). `readxl` isn't included; it always needs its own `library()` line.
:::
::::
:::::::

## Look before you trust it — every time

::::::: columns
:::: {.column width="60%"}
```{r wrangling-data-02}
glimpse(pine_csv_df)
```

```{r wrangling-data-03}
dim(pine_csv_df)
```

::: callout-tip
**🖐 Notice**

Six columns: `source_file`, `team`, `replicate`, `side`, `needle_length_mm`, `needle_width_mm`.\
Same structure whether it came from the `.csv` or the `.xlsx`.
:::
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Excel loves to turn a plain number column into a date, or a `replicate` like `1` into `1.0`. \
Always `glimpse()` an Excel import before doing anything else.
:::
::::
:::::::

## Part 2 · The Five Wrangling Verbs {.divider}

## The Five Wrangling Verbs

::::::: columns
:::: {.column width="60%"}
Almost everything you do to a data frame is one of these five functions, chained with the pipe `|>`:

| Function      | What it does                       |
|---------------|------------------------------------|
| `filter()`    | keep **rows** matching a condition |
| `select()`    | keep **columns** by name           |
| `mutate()`    | **add or change** a column         |
| `arrange()`   | **sort** rows                      |
| `summarize()` | **collapse** rows to a summary     |

All take a data frame as input and return a data frame — that's what makes them chainable.

::: callout-note
**✅ Key idea**

`summarize()` is Describing Your Data's whole topic. \
Today: the other four, plus saving what you build.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**Read a pipeline out loud**

```         
pine_csv_df |>
  filter(...) |>   # then
  select(...) |>   # then
  mutate(...) |>   # then
  arrange(...)     # then
```

Each `|>` reads as **"then."**
:::
::::
:::::::

## `filter()` — Picking Rows

:::::: columns
:::: {.column width="60%"}
```{r wrangling-data-04}
# Keep only the shady side
pine_csv_df |> filter(side == "shady") |> head()

# Keep only needles longer than 17 mm
pine_csv_df |> filter(needle_length_mm > 17) |> head()
```

::: callout-warning
**⚠️ Watch out!**

- `=` assigns a value
- `==` tests equality — `filter(side = "shady")` is an error
  - you want `filter(side == "shady")`.
:::
::::

::: {.column width="40%"}
**Common comparison operators:**

| Symbol     | Meaning                       |
|------------|-------------------------------|
| `==`       | exactly equal                 |
| `!=`       | not equal                     |
| `>` `<`    | greater / less than           |
| `>=` `<=`  | greater / less / or equal     |
| `is.na()`  | is the value missing?         |
| `!is.na()` | is the value **not** missing? |
:::
::::::

## `filter()` — Combining Conditions

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-05}
# Combine two conditions with a comma (AND)
pine_csv_df |> filter(side == "sunny", needle_length_mm > 15) |> head()
```
:::

:::: {.column width="40%"}
::: callout-tip
**AND vs. OR**

- A comma between conditions means **AND**
  - —every condition must be true for a row to be kept.
- Use `|` between conditions for **OR** instead
:::
::::
::::::

## `filter()` — OR, With a Real Example

::: callout-note
**🔮 Predict first:** `side == "shady" | needle_length_mm > 18` keeps a row if the side is shady **or** the needle is longer than 18 mm. \
Will this ever keep a *sunny* needle? \
Under what condition?
:::

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-06}
# Shady-side needles, OR any exceptionally long needle
pine_csv_df |>
  filter(side == "shady" | needle_length_mm > 18) |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

- `|` is the pipe **character** (logical OR), not the pipe **operator** `|>` (or `%>%`)
- They look alike but do completely different jobs — one is logical OR, the other chains steps together.
:::
::::
::::::

## `filter()` — Matching One of Several Values with `%in%`

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-07}
# Keep rows where team is one of two named teams
pine_csv_df |>
  filter(team %in% c("botany", "waterbugs")) |>
  head()
```

```{r wrangling-data-08}
# The long way, for comparison — %in% replaces a chain of |
pine_csv_df |>
  filter(team == "botany" | team == "waterbugs") |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

- `x %in% c(a, b, c)` asks "is `x` equal to *any* of these?"
  - it's the clean way to write an OR across many possible values without a long chain of `|`.
:::
::::
::::::

## `filter()` — Missing Values with `is.na()`

::::::: columns
:::: {.column width="60%"}
```{r wrangling-data-09}
# A tiny illustrative dataset with missing values
survey_df <- tibble(
  site  = c("A", "B", "C", "D"),
  count = c(12, NA, 8, NA)
)

survey_df |> filter(is.na(count))    # rows that ARE missing
survey_df |> filter(!is.na(count))   # rows that are NOT missing
```

::: callout-note
**✅ Key idea**

- `is.na(x)` is itself a boolean test
  -  `TRUE`/`FALSE` — so it plugs into `filter()`
  - exactly like `==` or `>` does.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**📖 Coming up**

`pine_csv_df` has no missing values, so `is.na()` has nothing to catch here. \
Describing Your Data digs into *why* missing values matter so much once you're computing a sample size or a mean.
:::
::::
:::::::

## `select()` — Picking Columns

:::::: columns
:::: {.column width="60%"}
```{r wrangling-data-10}
# Keep only side and length
pine_csv_df |> select(side, needle_length_mm) |> head()

# Drop the provenance column we do not need (minus sign)
pine_csv_df |> select(-source_file) |> head()
```

::: callout-note
**✅ Key idea**

`select()` doesn't touch rows — it only trims the columns you're carrying forward. Useful the moment a dataset has 30+ columns and you need 4 of them.
:::
::::

::: {.column width="40%"}
- Name the columns you want, in any order
- A leading `-` drops a column instead of keeping it
:::
::::::

## `select()` — Picking Columns by Pattern

::::: columns
::: {.column width="60%"}
```{r wrangling-data-11}
# Keep columns that contain a word
pine_csv_df |> select(contains("n")) |> head()
```
:::

::: {.column width="40%"}
**Useful helpers:**

- `starts_with("needle")` — super useful when related variables share a prefix
- `ends_with("_mm")`
- `contains("needle")`

📖 R4DS §3.3 — `select()`
:::
:::::

## `mutate()` — Creating and Changing Columns

::: callout-note
**🔮 Predict first:** `needle_length_mm / 10` converts millimeters to centimeters. \
Will `mutate()` **replace** `needle_length_mm`, or **add** a new column? \
How many columns will the result have?
:::

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-12}
# Add a converted-units column
pine_csv_df |>
  mutate(needle_length_cm = needle_length_mm / 10) |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`mutate()` keeps every existing column and adds new ones to the right. \
To *overwrite* a column, reuse its name: \
`mutate(needle_length_mm = round(needle_length_mm, 1))`.
:::
::::
::::::

## `mutate()` — More Examples

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-13}
# Add a log-transformed column
pine_csv_df |>
  mutate(log_length = log(needle_length_mm)) |>
  head()

# mutate() can build on a column it just created
pine_csv_df |>
  mutate(
    needle_length_cm  = needle_length_mm / 10,
    size_class = if_else(needle_length_mm > 15, "long", "short")
  ) |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-tip
**📖 Why log-transform?**

- Many biological measurements are **right-skewed**
  - a log transform pulls extreme values in and can make a distribution more symmetric before you summarize or test it.
- We'll check whether it matters for our data in a later module
- **Know which log you are using.** In R, `log()` is the *natural* log (ln, base e); `log2()` and `log10()` are the other two. Always say which one in your methods.
:::
::::
::::::

## Part 2b · `case_when()` — Recoding and Binning {.divider}

## `case_when()` — Turning a Number into a Category

::: callout-note
**🔮 Predict first:** We're about to bin `needle_length_mm` into `"short"` (< 13), `"medium"` (< 16), and `"long"`. The first needle in the data is 18.44 mm. Which bin does it land in — and would that change if the cutoffs used `<=` instead of `<`?
:::

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-14}
pine_csv_df |>
  mutate(
    size_class = case_when(
      needle_length_mm < 13 ~ "short",
      needle_length_mm < 16 ~ "medium",
      TRUE           ~ "long"
    )
  ) |>
  count(size_class)
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`case_when()` checks conditions **top to bottom** and uses the *first* one that's `TRUE`. \
Once a row matches `needle_length_mm < 13`, R never even looks at the later conditions for that row.
:::
::::
::::::

## `case_when()` — Always End With `TRUE ~`

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-15}
# What happens if we forget the catch-all?
pine_csv_df |>
  mutate(
    size_class = case_when(
      needle_length_mm < 13 ~ "short",
      needle_length_mm < 16 ~ "medium"
      # no TRUE ~ ... here
    )
  ) |>
  count(size_class)
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Drop the `TRUE ~ "long"` line and all 21 "long" needles silently become `NA` instead of erroring. `case_when()` never warns you — always finish with a `TRUE ~ ...` catch-all, even when you're sure you've covered every case.
:::
::::
::::::

## `case_when()` — Recoding Categories, Not Just Numbers

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-16}
# Give each field team a short code
pine_csv_df |>
  mutate(
    team_code = case_when(
      team == "botany"     ~ "BOT",
      team == "grace_jace" ~ "GRJ",
      team == "one"        ~ "ONE",
      team == "waterbugs"  ~ "WBG",
      TRUE                 ~ "UNK"
    )
  ) |>
  distinct(team, team_code)
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

`case_when()` isn't just for numeric cutoffs — every condition on the left of a `~` can be any boolean test: `==`, `%in%`, `>`, even combinations with `&` and `|`.

The `"UNK"` catch-all should never actually appear here — if it does, that's your signal a team name was misspelled somewhere upstream.
:::
::::
::::::

## `arrange()` — Sorting Rows

::::: columns
::: {.column width="60%"}
```{r wrangling-data-17}
# Shortest needles first
pine_csv_df |> arrange(needle_length_mm)

# Longest first
pine_csv_df |> arrange(desc(needle_length_mm))
```
:::

::: {.column width="40%"}
- Default is **ascending** (smallest first)
- `desc()` reverses it
:::
:::::

## `arrange()` — Sorting by Multiple Columns

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-18}
# Sort by side, then by length within each side
pine_csv_df |> arrange(side, desc(needle_length_mm))
```
:::

:::: {.column width="40%"}
- Most useful at the **end** of a pipeline, to inspect results in a sensible order

::: callout-tip
`arrange()` never changes *which* rows you have — only the order they print in.
:::
::::
::::::

## The Full Pipeline

:::::: columns
:::: {.column width="60%"}
```{r wrangling-data-19}
pine_wrangled_df <- pine_csv_df |>
  filter(needle_length_mm > 0) |>                 # drop impossible values
  select(team, side, needle_length_mm, needle_width_mm) |>  # keep what we need
  mutate(
    needle_length_cm  = needle_length_mm / 10,
    log_length = log(needle_length_mm),
    size_class = case_when(
      needle_length_mm < 13 ~ "short",
      needle_length_mm < 16 ~ "medium",
      TRUE                  ~ "long"
    )
  ) |>
  arrange(side, desc(needle_length_mm))            # sort for inspection

head(pine_wrangled_df)
```

::: callout-note
**✅ Key idea**

The pipe `|>` is what makes this readable top to bottom — a recipe, not a nest of parentheses.
:::
::::

::: {.column width="40%"}
**Read it out loud:**

> - Take `pine_csv_df`,
>   - **then** filter to real values,
>     - **then** select four columns,
>       - **then** add three new columns,
>         - **then** sort by side and length.
:::
::::::

## Part 3 · A Duplicate-Row Problem {.divider}

## A Duplicate-Row Problem

::::::: columns
:::: {.column width="60%"}
Remember **pseudoreplication** from Getting Started — one tree standing in for many? Duplicate rows are the data-entry version of the same mistake: the same measurement counted twice inflates your sample size without adding real information.

```{r wrangling-data-20}
needle_check_df <- read_csv("data/pine_needles_error.csv")

nrow(needle_check_df)
n_distinct(needle_check_df)
```

::: callout-important
**Something's wrong:** `nrow()` and `n_distinct()` don't match. Some rows are exact repeats.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 New words**

- **`n_distinct()`** — counts how many *unique* rows exist
- **`distinct()`** — returns the data with duplicate rows removed — use it deliberately, not reflexively
- **`unique()`** — lists the distinct values in a single column; the fastest way to spot misspellings
:::
::::
:::::::

## Finding the duplicates

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-21}
# Which rows are exact duplicates of an earlier row?
needle_check_df |> filter(duplicated(needle_check_df))
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Field habit**

Before you trust any dataset — yours or a collaborator's — check `nrow()` against `n_distinct()`. It takes five seconds and catches a real, common mistake.
:::
::::
::::::

## Removing duplicates with `distinct()`

:::::: columns
::: {.column width="60%"}
```{r wrangling-data-22}
# Keep only one copy of each unique row
needle_clean_df <- needle_check_df |> distinct()

nrow(needle_check_df)
nrow(needle_clean_df)
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`distinct()` only catches *exact* duplicate rows. A typo'd duplicate (`waterbugs` vs `Waterbugs`) slips right through — always eyeball your data too.
:::
::::
::::::

## Part 3b · The Rest of the Mess {.divider}

## `unique()` — what else is wrong in here?

::::::: columns
:::: {.column width="60%"}
Duplicates were only the *first* problem in this file. Look at the columns themselves:

```{r wrangling-data-23}
names(needle_check_df)
```

```{r wrangling-data-24}
unique(needle_clean_df$`Side of Tree`)
```

::: callout-important
**Two values became six.** `shady`, `Shady`, `SHADY` — R treats every one as a separate category. Any `group_by()` or `count()` you run right now is wrong.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 New words**

- **`clean_names()`** (janitor) — turns `Needle Length (mm)` into `needle_length_mm`
- **`str_to_lower()`** — force text to lowercase
- **`str_squish()`** — strip leading/trailing spaces
:::

::: callout-tip
**🖐 It was almost worse**

This file also has stray spaces (`"sunny "`, `" shady"`). You don't see them because **`read_csv()` trims whitespace for you by default.** Prove it to yourself:

``` r
read_csv("data/pine_needles_error.csv", trim_ws = FALSE)
```

Now there are **ten** values, not six. Not every import tool is that kind — which is why we still run `str_squish()` below.
:::
::::
:::::::

## Fixing names and text in one pipeline

::::::: columns
:::: {.column width="60%"}
```{r wrangling-data-25}
needle_fixed_df <- needle_check_df |>
  clean_names() |>                 # Needle Length (mm) -> needle_length_mm
  distinct() |>                    # drop the repeated rows
  mutate(
    side = str_squish(str_to_lower(side_of_tree)),
    team = str_squish(str_to_lower(team_name))
  )

unique(needle_fixed_df$side)
unique(needle_fixed_df$team)
```

::: callout-warning
**⚠️ Watch out!**

`str_to_lower()` collapsed `Waterbugs` into `waterbugs` — but **`water bugs` is still its own value**. Lowercasing fixes case, not spelling. That last one needs `case_when()`.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Field habit**

Run `unique()` on *every* text column the first time you open a dataset. It takes seconds and it is how you find the problems that silently split your groups.
:::
::::
:::::::

## Numbers that aren't numbers

::::::: columns
:::: {.column width="60%"}
```{r wrangling-data-26}
class(needle_fixed_df$needle_length_mm)
class(needle_fixed_df$needle_width_mm)
```

**Both** imported as text, not numbers — somebody typed `17.73 mm` in a cell, somebody used a comma decimal (`13,98`), and the missing values were written four different ways.

```{r wrangling-data-27}
needle_final_df <- needle_fixed_df |>
  mutate(
    team = case_when(team == "water bugs" ~ "waterbugs",
                     TRUE                 ~ team),
    needle_length_mm = needle_length_mm |>
      str_remove(" mm") |>                    # drop the typed unit
      str_replace(",", ".") |>                # comma decimal -> point
      na_if("") |> na_if("NA") |>             # blank and literal "NA"
      as.numeric(),
    needle_width_mm = needle_width_mm |>
      na_if("") |> na_if("na") |> na_if(".") |>   # three more ways to write "missing"
      as.numeric(),
    needle_length_mm = if_else(needle_length_mm == -999,
                               NA_real_, needle_length_mm)   # sentinel value
  )

class(needle_final_df$needle_length_mm)
sum(is.na(needle_final_df$needle_length_mm))
sum(is.na(needle_final_df$needle_width_mm))
```
::::

:::: {.column width="40%"}
::: callout-important
**⚠️ The trap:** `parse_number("13,98")` returns **1398** — it reads the comma as a thousands separator and never warns you.

A 1398 mm pine needle is obvious *if you look*. **Always summarize or plot a column right after you clean it.**
:::

::: callout-note
**📖 New words**

- **`na_if(x, "")`** — turn a specific value into `NA`
- **`-999`** — a "sentinel" code some field sheets use for missing. R has no idea it means missing until you tell it.
:::
::::
:::::::

## Part 4 · Saving Your Work {.divider}

## The one-way street

::::::: columns
:::: {.column width="60%"}
Everything we did today runs in **one direction**:

```
data/     ← raw, read-only, never written to
   │  read_csv()
   ▼
scripts/  ← the cleaning code you just wrote
   │  write_csv()
   ▼
output/   ← everything R produces
```

The raw file is **evidence**. The script is the **method**. The output is a **result you can always rebuild** by re-running the method on the evidence.

::: callout-important
**Never write into `data/`.** If a script can overwrite the raw file, then one bad run destroys the only copy of what you actually measured — and no one can check your work.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

If you can't reproduce a result from **raw data + script**, the problem is in the script — never in the raw file. Keeping `data/` read-only is what makes that guarantee possible.
:::
::::
:::::::

## Tidy up before you save

::::::: columns
:::: {.column width="60%"}
`needle_final_df` still carries the wreckage of cleaning — the original messy columns *and* our fixed ones:

```{r wrangling-data-28}
names(needle_final_df)
```

Nobody downstream needs `team_name` **and** `team`. Keep the columns you mean to keep, in the order you want them:

```{r wrangling-data-29}
needle_out_df <- needle_final_df |>
  select(team, side, replicate = replicate_number,
         needle_length_mm, needle_width_mm) |>
  arrange(team, side, replicate)

glimpse(needle_out_df)
```
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Field habit**

`select()` right before `write_csv()` is the last checkpoint. A saved file with `side` *and* `side_of_tree` in it will confuse someone in three months — and that someone is usually you.
:::

::: callout-note
`select(new_name = old_name)` renames while it selects — handy for dropping janitor's `replicate_number` back to `replicate`.
:::
::::
:::::::

## `write_csv()` — the mirror image of `read_csv()`

::::::: columns
:::: {.column width="60%"}
```{r wrangling-data-30}
#| eval: false
write_csv(needle_out_df, "output/pine_needles_cleaned.csv")
```

Read from `data/`, write to `output/`. Same shape of call, opposite direction:

| Reading | Writing |
|---------|---------|
| `read_csv("data/…")` | `write_csv(df, "output/…")` |
| file → data frame | data frame → file |

::: callout-warning
**⚠️ Watch out!**

`write_csv()` **overwrites without asking** — no warning, no undo. That is exactly why the target is `output/` and never `data/`: re-running your script should be free, and it is only free if the thing it clobbers is something the script can rebuild.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 Note the argument order**

`write_csv(df, path)` — **data frame first**, then where to put it. `read_csv(path)` takes only the path. Getting these backwards is one of the most common errors of the term.
:::
::::
:::::::

## Naming your saved files

::::::: columns
:::: {.column width="60%"}
- Name the file for **what happened to it**: `pine_needles_cleaned.csv`, not `pine2.csv` or `final.csv`
- If a script produces several stages, number them: `01_cleaned.csv`, `02_summarised.csv`
- Anyone re-running your script from scratch should get a **byte-identical** output file

::: callout-warning
**⚠️ Watch out!**

`final.csv`, `final_v2.csv`, `final_FINAL.csv` — version control by filename. This is exactly the mess a project folder with clear names and a re-runnable script is meant to prevent.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Try it yourself**

Delete your `output/` folder entirely, then re-run your script top to bottom.

Everything should come back. **If it doesn't, something important only exists in your console history — and that is a bug in your script.**
:::
::::
:::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Imported the same data from a `.csv` **and** an `.xlsx`
- Used `filter()`, `select()`, `mutate()`, and `arrange()` — and chained them into a pipeline
- Filtered with the full boolean toolkit: `==`, `!=`, `>`/`<`, `%in%`, `is.na()`, AND (`,`) vs. OR (`|`)
- Recoded and binned columns with `case_when()` — always ending in a `TRUE ~` catch-all
- Converted units and log-transformed a column with `mutate()`
- Caught and removed duplicate rows with `distinct()`
- Cleaned messy headers with `clean_names()`, messy text with `str_to_lower()`, and numbers-stored-as-text with `as.numeric()`
- Trimmed to the columns worth keeping and wrote the result to `output/` — leaving `data/` untouched

::: callout-tip
**🖐 Before next class**

Finish the worksheet: take `pine_needles_error.csv` from `data/`, clean it end to end, `select()` the columns worth keeping, and write the result to `output/`. Then delete `output/`, re-run your script, and check that it all comes back.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Up next — Describing Your Data**

- Mean, median, variance, SD, SE
- The `length()` trap and the `sum(!is.na())` fix
- `group_by()` + `summarize()` for tidy group comparisons
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` and `library(readxl)` loaded? Spelling? A missing `)` or `|>` at the *start* of a line?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
