---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Wrangling Your Data"
subtitle: "Importing more than one way, reshaping columns, and saving what you make"
description: "Reading CSV and Excel files, the filter/select/mutate/arrange verbs, the full boolean-operator toolkit (==, !=, >, <, is.na(), %in%, & vs. |), recoding and binning with case_when(), log and unit-conversion transforms, catching duplicate rows, and writing your cleaned data back out — all on the pine needle dataset."
categories: [Wrangling, R, tidyverse]
order: 2
week: 1
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 3
    fig-height: 3
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true
---

## Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Turned a field observation into a testable **H₀ / Hₐ**
- Built a project folder (`data/`, `scripts/`, `figures/` , `outout/)`
- Loaded `pine_needles.csv` and made your first `ggplot`

::: callout-note
**✅ Key idea from Getting Started**

You can get data *into* R and make a plot. Today we learn to **reshape** what you have and **save** what you make — the two skills every later module depends on.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  Importing data more than one way
2.  The five wrangling verbs
3.  `filter()` in depth — every boolean operator, `%in%`, AND vs. OR
4.  `mutate()` — unit conversions and log transforms
5.  `case_when()` — recoding and binning
6.  Catching duplicate rows
7.  Saving your work with `write_csv()`
:::
::::
:::::::

## Part 1 · Importing Data More Than One Way {.divider}

## The same data, two file formats

::::::: columns
:::: {.column width="60%"}
Field data doesn't always arrive as a `.csv`. Excel workbooks are everywhere in ecology — you need both tools.

```{r}
# install.packages("readxl")
# install.packages("janitor")
library(tidyverse)
library(readxl)   # reading Excel files
library(janitor)  # cleans variable names, excel poop, rounding issues

pine_csv_df   <- read_csv("data/pine_needles.csv")
pine_excel_df <- read_excel("data/pine_needles.xlsx")
```

::: callout-note
**✅ Key idea**

`read_csv()` comes from `readr` (loaded with `tidyverse`). \
Excel files need the separate `readxl` package \
— install once, `library()` every session.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 New word**

**Package vs. bundle** — `tidyverse` is a bundle of packages (`readr`, `dplyr`, `ggplot2`, …). `readxl` isn't included; it always needs its own `library()` line.
:::
::::
:::::::

## Look before you trust it — every time

::::::: columns
:::: {.column width="60%"}
```{r}
glimpse(pine_csv_df)
```

```{r}
dim(pine_csv_df)
```

::: callout-tip
**🖐 Notice**

Six columns: `date`, `group`, `n_s`, `sun`, `tree_no`, `length_mm` or what you call them.\
Same structure whether it came from the `.csv` or the `.xlsx`.
:::
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Excel loves to turn a plain number column into a date, or a `tree_no` like `1` into `1.0`. \
Always `glimpse()` an Excel import before doing anything else.
:::
::::
:::::::

## Part 2 · The Five Wrangling Verbs {.divider}

## The Five Wrangling Verbs

::::::: columns
:::: {.column width="60%"}
Almost everything you do to a data frame is one of these five functions, chained with the pipe `|>`:

| Function      | What it does                       |
|---------------|------------------------------------|
| `filter()`    | keep **rows** matching a condition |
| `select()`    | keep **columns** by name           |
| `mutate()`    | **add or change** a column         |
| `arrange()`   | **sort** rows                      |
| `summarize()` | **collapse** rows to a summary     |

All take a data frame as input and return a data frame — that's what makes them chainable.

::: callout-note
**✅ Key idea**

`summarize()` is Describing Your Data's whole topic. \
Today: the other four, plus saving what you build.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**Read a pipeline out loud**

```         
pine_csv_df |>
  filter(...) |>   # then
  select(...) |>   # then
  mutate(...) |>   # then
  arrange(...)     # then
```

Each `|>` reads as **"then."**
:::
::::
:::::::

## `filter()` — Picking Rows

:::::: columns
:::: {.column width="60%"}
```{r}
# Keep only the shady (sheltered) side
pine_csv_df |> filter(n_s == "n") |> head()

# Keep only needles longer than 20 mm
pine_csv_df |> filter(length_mm > 20) |> head()
```

::: callout-warning
**⚠️ Watch out!**

- `=` assigns a value
-  `==` tests equality. `filter(n_s = "n")` is an error
  -  you want `filter(n_s == "n")`.
:::
::::

::: {.column width="40%"}
**Common comparison operators:**

| Symbol     | Meaning                       |
|------------|-------------------------------|
| `==`       | exactly equal                 |
| `!=`       | not equal                     |
| `>` `<`    | greater / less than           |
| `>=` `<=`  | greater / less / or equal     |
| `is.na()`  | is the value missing?         |
| `!is.na()` | is the value **not** missing? |
:::
::::::

## `filter()` — Combining Conditions

:::::: columns
::: {.column width="60%"}
```{r}
# Combine two conditions with a comma (AND)
pine_csv_df |> filter(n_s == "s", length_mm > 15) |> head()
```
:::

:::: {.column width="40%"}
::: callout-tip
**AND vs. OR**

- A comma between conditions means **AND**
  - —every condition must be true for a row to be kept.
- Use `|` between conditions for **OR** instead
:::
::::
::::::

## `filter()` — OR, With a Real Example

::: callout-note
**🔮 Predict first:** `n_s == "n" | length_mm > 22` keeps a row if the side is shady **or** the needle is longer than 22mm. \
Will this ever keep a *sunny* needle? \
Under what condition?
:::

:::::: columns
::: {.column width="60%"}
```{r}
# Shady-side needles, OR any exceptionally long needle
pine_csv_df |>
  filter(n_s == "n" | length_mm > 22) |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

- `|` is the pipe **character**, not the pipe **operator** `|> or %>%`
-  Look related but do completely different jobs
  - one is logical OR, the other chains steps together.
:::
::::
::::::

## `filter()` — Matching One of Several Values with `%in%`

:::::: columns
::: {.column width="60%"}
```{r}
# Keep rows where group is one of two named teams
pine_csv_df |>
  filter(group %in% c("cephalopods", "salmon")) |>
  head()
```

```{r}
# The long way, for comparison — %in% replaces a chain of |
pine_csv_df |>
  filter(group == "cephalopods" | group == "salmon") |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

- `x %in% c(a, b, c)` asks "is `x` equal to *any* of these?"
  - it's the clean way to write an OR across many possible values without a long chain of `|`.
:::
::::
::::::

## `filter()` — Missing Values with `is.na()`

::::::: columns
:::: {.column width="60%"}
```{r}
# A tiny illustrative dataset with missing values
survey_df <- tibble(
  site  = c("A", "B", "C", "D"),
  count = c(12, NA, 8, NA)
)

survey_df |> filter(is.na(count))    # rows that ARE missing
survey_df |> filter(!is.na(count))   # rows that are NOT missing
```

::: callout-note
**✅ Key idea**

- `is.na(x)` is itself a boolean test
  -  `TRUE`/`FALSE` — so it plugs into `filter()`
  - exactly like `==` or `>` does.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**📖 Coming up**

`pine_csv_df` has no missing values, so `is.na()` has nothing to catch here. \
Describing Your Data digs into *why* missing values matter so much once you're computing a sample size or a mean.
:::
::::
:::::::

## `select()` — Picking Columns

:::::: columns
:::: {.column width="60%"}
```{r}
# Keep only side and length
pine_csv_df |> select(n_s, length_mm) |> head()

# Drop the date column (minus sign)
pine_csv_df |> select(-date) |> head()
```

::: callout-note
**✅ Key idea**

`select()` doesn't touch rows — it only trims the columns you're carrying forward. Useful the moment a dataset has 30+ columns and you need 4 of them.
:::
::::

::: {.column width="40%"}
- Name the columns you want, in any order
- A leading `-` drops a column instead of keeping it
:::
::::::

## `select()` — Picking Columns by Pattern

::::: columns
::: {.column width="60%"}
```{r}
# Keep columns that contain a word
pine_csv_df |> select(contains("n")) |> head()
```
:::

::: {.column width="40%"}
**Useful helpers:**

- `starts_with("x") - super useful for similar units or types of data in variable`
- `ends_with("_mm")`
- `contains("sun")`

📖 R4DS §3.3 — `select()`
:::
:::::

## `mutate()` — Creating and Changing Columns

::: callout-note
**🔮 Predict first:** `length_mm / 10` converts millimeters to centimeters. \
Will `mutate()` **replace** `length_mm`, or **add** a new column? \
How many columns will the result have?
:::

:::::: columns
::: {.column width="60%"}
```{r}
# Add a converted-units column
pine_csv_df |>
  mutate(length_cm = length_mm / 10) |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`mutate()` keeps every existing column and adds new ones to the right. \
To *overwrite* a column, reuse its name: \
`mutate(length_mm = round(length_mm, 1))`.
:::
::::
::::::

## `mutate()` — More Examples

:::::: columns
::: {.column width="60%"}
```{r}
# Add a log-transformed column
pine_csv_df |>
  mutate(log_length = log(length_mm)) |>
  head()

# mutate() can build on a column it just created
pine_csv_df |>
  mutate(
    length_cm  = length_mm / 10,
    size_class = if_else(length_mm > 18, "long", "short")
  ) |>
  head()
```
:::

:::: {.column width="40%"}
::: callout-tip
**📖 Why log-transform?**

- Many biological measurements are **right-skewed**
  - a log transform pulls extreme values in and can make a distribution more symmetric before you summarize or test it.
- We'll check whether it matters for our data in a later module
- Know what log you are doing log2 log10 or ln?????
:::
::::
::::::

## Part 2b · `case_when()` — Recoding and Binning {.divider}

## `case_when()` — Turning a Number into a Category

::: callout-note
**🔮 Predict first:** We're about to bin `length_mm` into `"short"`, `"medium"`, and `"long"`. Needle 1 in the data is 20mm. Which bin does it land in — and does that depend on whether the cutoff is `<` or `<=`?
:::

:::::: columns
::: {.column width="60%"}
```{r}
pine_csv_df |>
  mutate(
    size_class = case_when(
      length_mm < 16 ~ "short",
      length_mm < 20 ~ "medium",
      TRUE           ~ "long"
    )
  ) |>
  count(size_class)
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

`case_when()` checks conditions **top to bottom** and uses the *first* one that's `TRUE`. \
Once a row matches `length_mm < 16`, R never even looks at the later conditions for that row.
:::
::::
::::::

## `case_when()` — Always End With `TRUE ~`

:::::: columns
::: {.column width="60%"}
```{r}
# What happens if we forget the catch-all?
pine_csv_df |>
  mutate(
    size_class = case_when(
      length_mm < 16 ~ "short",
      length_mm < 20 ~ "medium"
      # no TRUE ~ ... here
    )
  ) |>
  count(size_class)
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

Drop the `TRUE ~ "long"` line and every "long" needle silently becomes `NA` instead of erroring. `case_when()` never warns you — always finish with a `TRUE ~ ...` catch-all, even when you're sure you've covered every case.
:::
::::
::::::

## `case_when()` — Recoding Categories, Not Just Numbers

:::::: columns
::: {.column width="60%"}
```{r}
# Give each field team a short code
pine_csv_df |>
  mutate(
    team_code = case_when(
      group == "cephalopods" ~ "CEPH",
      group == "salmon"      ~ "SALM",
      group == "crayfish"    ~ "CRAY",
      group == "snail"       ~ "SNAI",
      TRUE                   ~ "UNK"
    )
  ) |>
  distinct(group, team_code)
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

`case_when()` isn't just for numeric cutoffs — every condition on the left of a `~` can be any boolean test: `==`, `%in%`, `>`, even combinations with `&` and `|`.

The `"UNK"` catch-all should never actually appear here — if it does, that's your signal a team name was misspelled somewhere upstream.
:::
::::
::::::

## `arrange()` — Sorting Rows

::::: columns
::: {.column width="60%"}
```{r}
# Shortest needles first
pine_csv_df |> arrange(length_mm)

# Longest first
pine_csv_df |> arrange(desc(length_mm))
```
:::

::: {.column width="40%"}
- Default is **ascending** (smallest first)
- `desc()` reverses it
:::
:::::

## `arrange()` — Sorting by Multiple Columns

:::::: columns
::: {.column width="60%"}
```{r}
# Sort by side, then by length within each side
pine_csv_df |> arrange(n_s, desc(length_mm))
```
:::

:::: {.column width="40%"}
- Most useful at the **end** of a pipeline, to inspect results in a sensible order

::: callout-tip
`arrange()` never changes *which* rows you have — only the order they print in.
:::
::::
::::::

## The Full Pipeline

:::::: columns
:::: {.column width="60%"}
```{r}
pine_wrangled_df <- pine_csv_df |>
  filter(length_mm > 0) |>                 # drop impossible values
  select(group, n_s, sun, length_mm) |>    # keep what we need
  mutate(
    length_cm  = length_mm / 10,
    log_length = log(length_mm),
    size_class = case_when(
      length_mm < 16 ~ "short",
      length_mm < 20 ~ "medium",
      TRUE           ~ "long"
    )
  ) |>
  arrange(n_s, desc(length_mm))            # sort for inspection

head(pine_wrangled_df)
```

::: callout-note
**✅ Key idea**

The pipe `|>` is what makes this readable top to bottom — a recipe, not a nest of parentheses.
:::
::::

::: {.column width="40%"}
**Read it out loud:**

> - Take `pine_csv_df`,
>   - **then** filter to real values,
>     - **then** select four columns,
>       - **then** add three new columns,
>         - **then** sort by side and length.
:::
::::::

## Part 3 · A Duplicate-Row Problem {.divider}

## A Duplicate-Row Problem

::::::: columns
:::: {.column width="60%"}
Remember **pseudoreplication** from Getting Started — one tree standing in for many? Duplicate rows are the data-entry version of the same mistake: the same measurement counted twice inflates your sample size without adding real information.

```{r}
needle_check_df <- read_csv("data/pine_needles_error.csv")

nrow(needle_check_df)
n_distinct(needle_check_df)
```

::: callout-important
**Something's wrong:** `nrow()` and `n_distinct()` don't match. Some rows are exact repeats.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 New words**

- **`n_distinct()`** — counts how many *unique* rows exist
- **`distinct()`** — returns the data with duplicate rows removed - be careful!!!\\
- **unique()** - looks for the unique rows or names in rows for mispellinbg
:::
::::
:::::::

## Finding the duplicates

:::::: columns
::: {.column width="60%"}
```{r}
# Which rows are exact duplicates of an earlier row?
needle_check_df |> filter(duplicated(needle_check_df))
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Field habit**

Before you trust any dataset — yours or a collaborator's — check `nrow()` against `n_distinct()`. It takes five seconds and catches a real, common mistake.
:::
::::
::::::

## Removing duplicates with `distinct()`

:::::: columns
::: {.column width="60%"}
```{r}
# Keep only one copy of each unique row
needle_clean_df <- needle_check_df |> distinct()

nrow(needle_check_df)
nrow(needle_clean_df)
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`distinct()` only catches *exact* duplicate rows. A typo'd duplicate (`John` vs `john`) slips right through — always eyeball your data too.
:::
::::
::::::

## Part 4 · Saving Your Work {.divider}

## `write_csv()` — the mirror image of `read_csv()`

::::::: columns
:::: {.column width="60%"}
```{r}
#| eval: false
write_csv(pine_wrangled_df, "data/pine_needles_wrangled.csv")
```

::: callout-important
**Raw data is read-only.** `pine_needles.csv` never gets overwritten — `write_csv()` always creates a **new, separate file**. That's why we named it `pine_needles_wrangled.csv`, not `pine_needles.csv`.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

If you can't reproduce a result from raw data + a script, something's wrong with the script — never with the raw file. Keeping raw data untouched is what makes that guarantee possible.
:::
::::
:::::::

## Naming your saved files

::::::: columns
:::: {.column width="60%"}
- Never overwrite `data/pine_needles.csv` — that file is raw
- Save wrangled output with a name that says what happened: `pine_needles_wrangled.csv`, not `pine_needles2.csv` or `final.csv`
- Anyone re-running your script from scratch should get the exact same output file

::: callout-warning
**⚠️ Watch out!**

`final.csv`, `final_v2.csv`, `final_FINAL.csv` — this is exactly the mess a project folder with relative paths and clear names is meant to prevent.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Try it yourself**

Predict what `write_csv(pine_wrangled_df, "data/pine_needles_wrangled.csv")` does if that file doesn't exist yet. What if it already does?
:::
::::
:::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Imported the same data from a `.csv` **and** an `.xlsx`
- Used `filter()`, `select()`, `mutate()`, and `arrange()` — and chained them into a pipeline
- Filtered with the full boolean toolkit: `==`, `!=`, `>`/`<`, `%in%`, `is.na()`, AND (`,`) vs. OR (`|`)
- Recoded and binned columns with `case_when()` — always ending in a `TRUE ~` catch-all
- Converted units and log-transformed a column with `mutate()`
- Caught and removed duplicate rows with `distinct()`
- Saved a new file with `write_csv()` without touching the raw data

::: callout-tip
**🖐 Before next class**

Finish the worksheet: wrangle `pine_needles.csv` into a new data frame with at least one unit conversion, one log transform, and one `case_when()` category, then save it with `write_csv()`.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Up next — Describing Your Data**

- Mean, median, variance, SD, SE
- The `length()` trap and the `sum(!is.na())` fix
- `group_by()` + `summarize()` for tidy group comparisons
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` and `library(readxl)` loaded? Spelling? A missing `)` or `|>` at the *start* of a line?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
