---
title: "PARKED — power & Type I/II material moved out of Probability & Inference"
---

<!--
This file is NOT rendered (leading underscore = Quarto ignores it).

Moved here 2026-09-16 from content/probability_inference/probability_inference_lecture.qmd
so the Probability & Inference lecture could end at confidence intervals and
hypothesis testing. The Type I/II error slides duplicated material already in
power_and_error_lecture.qmd; the grayling power example below is the only piece
that is genuinely new and worth folding in later.

TO DO: merge the grayling power example into the "Statistical Power" section of
power_and_error_lecture.qmd, or keep it as a second worked example alongside the
pine-needle one.
-->

## Type I and Type II Errors  {#parked-errors}

**Type I Error (False Positive)**

- Rejecting H₀ when it is actually true
- Probability = α (significance level)
- "Finding an effect that isn't real"

**Type II Error (False Negative)**

- Failing to reject H₀ when it is actually false
- Probability = β
- "Missing an effect that is real"

**Statistical Power = 1 − β**

- Probability of correctly rejecting a false H₀
- Increases with: larger sample size, larger effect size, lower variability, higher α

**Reading the two-curve figure:**

- Black curve (null): distribution of the test statistic when H₀ is true
- Green curve (alternative): distribution when there really is an effect
- Red shaded area (Type I): area under the **null** curve right of the critical value
- Blue shaded area (Type II): area under the **alternative** curve left of the critical value
- Power = area under the **green** curve right of the critical value
- Moving the critical value trades one error against the other — you cannot shrink both

```{r parked-error-plot}
#| eval: false
x <- seq(-4, 8, length.out = 1000)
null_y <- dnorm(x, mean = 0, sd = 1)
alt_y  <- dnorm(x, mean = 3, sd = 1)
crit   <- qnorm(0.95)

type1 <- data.frame(x = seq(crit, 8, length.out = 200)) %>%
  mutate(y = dnorm(x, 0, 1))
type2 <- data.frame(x = seq(-4, crit, length.out = 200)) %>%
  mutate(y = dnorm(x, 3, 1))

ggplot() +
  geom_line(data = data.frame(x = x, y = null_y), aes(x, y)) +
  geom_line(data = data.frame(x = x, y = alt_y), aes(x, y), colour = "green4") +
  geom_area(data = type1, aes(x, y), fill = "red",  alpha = 0.4) +
  geom_area(data = type2, aes(x, y), fill = "blue", alpha = 0.4) +
  geom_vline(xintercept = crit, linetype = "dashed") +
  labs(x = "Test Statistic", y = "Probability Density") +
  theme_classic(base_size = 9)
```

## Power example on the grayling data  {#parked-power}

Power to detect a 30 mm difference in mean length between lakes I3 and I8.

```{r parked-power-grayling}
#| eval: false
grayling_df <- read_csv("data/gray_I3_I8.csv")
lake_I3 <- grayling_df %>% filter(lake == "I3")
lake_I8 <- grayling_df %>% filter(lake == "I8")

# count only the fish we actually measured (the length() trap from week 2)
n1 <- sum(!is.na(lake_I3$length_mm))
n2 <- sum(!is.na(lake_I8$length_mm))

sd_pooled <- sqrt((var(lake_I3$length_mm, na.rm = TRUE) * (n1 - 1) +
                   var(lake_I8$length_mm, na.rm = TRUE) * (n2 - 1)) /
                  (n1 + n2 - 2))

# power.t.test assumes EQUAL group sizes. Ours are 66 and 102, so use the
# harmonic mean - the effective n per group when the design is unbalanced.
n_effective <- 2 / (1/n1 + 1/n2)

power_result <- power.t.test(n = n_effective,
                             delta = 30,       # difference we care about, in mm
                             sd = sd_pooled,   # same scale as delta
                             sig.level = 0.05,
                             type = "two.sample",
                             alternative = "two.sided")
power_result
```

::: callout-warning
**⚠️ An assumption quietly made here**

`sd_pooled` averages the two lakes' SDs, which assumes they are equal. They are
not: sd(I3) = 28.3 mm but sd(I8) = 52.3 mm — nearly double. That is also why the
two-sample t-test is a **Welch** test (R's default), which does *not* assume
equal variances. Treat this power number as a rough guide, not a precise answer.
:::

## Scenario questions (from the old Practice Exercise 7)

Identify whether a Type I or Type II error might have occurred:

1. A researcher concludes that a new fishing regulation increased grayling size, when in fact it had no effect.
2. A study fails to detect a real decline in grayling population due to warming water, concluding there was no effect.
