---
title: "Module 4 — Graphing Effectively"
subtitle: "Boxplots, group comparisons, and saving a publication-ready figure"
description: "Boxplots with jittered raw points, mapping color and fill by group, plotting mean ± SE directly with stat_summary(), faceting, a quick theme tour, and a real ggsave() drill — all on the pine needle dataset."
categories: [Graphing, ggplot2, R]
order: 4
week: 4
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
    css: custom.css
  revealjs:
    output-ext: "slides.html"
    theme: [simple, custom.scss]
    slide-number: true
  docx:
    reference-doc: ../reference-doc.docx
  pptx: default
execute:
  echo: true
  warning: false
  message: false
  fig-width: 3
  fig-height: 3
  fig-dpi: 300
---

# Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Computed mean, median, SD, and SE — by hand and with `summarize()`
- Fixed the `length()` trap with `sum(!is.na())`
- Got tidy per-group tables with `group_by()` + `summarize()`

::: callout-note
**✅ Key idea from Module 3**

You now have real numbers describing lee vs. windward needles. Today we make those numbers **visible** — and save the result properly.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  The `ggplot` grammar, recapped
2.  Boxplots done right — with raw points
3.  Plotting mean ± SE directly
4.  Faceting and a quick theme tour
5.  `ggsave()`, for real this time
:::
::::
:::::::

# Part 1 · The `ggplot` Grammar, Recapped

::::::: columns
:::: {.column width="60%"}
Every `ggplot` is three pieces joined with `+`:

1.  **data** — which data frame
2.  **aes()** — which columns map to x, y, color, fill…
3.  **geom** — how to draw it

```{r}
library(tidyverse)

pine_df <- read_csv("data/pine_needles.csv")
```

::: callout-note
**✅ Key idea**

Everything today is the *same three pieces* — we're just adding more layers and more polish on top of them.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Recap from Module 1**

```{r}
ggplot(pine_df, aes(x = n_s, y = length_mm)) +
  geom_point()
```
:::
::::
:::::::

# Part 2 · Boxplot Anatomy

::::::: columns
:::: {.column width="60%"}
```{r}
pine_box_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side (n = lee, s = windward)",
    y = "Needle length (mm)",
    fill = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

pine_box_plot
```
::::

:::: {.column width="40%"}
**Boxplot anatomy:**

- Middle line = **median**
- Box edges = 25th and 75th percentile (**IQR**)
- Whiskers = 1.5 × IQR
- Dots beyond whiskers = **outliers**

📖 R4DS §9 — Layers
::::
:::::::

# Boxplot + Raw Points

::: callout-note
**🔮 Predict first:** With only 6 needles per side per tree, does a boxplot alone show you everything? What might it hide?
:::

::::::: columns
:::: {.column width="60%"}
```{r}
pine_jitter_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.6,
    size = 2
  ) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side of tree", y = "Needle length (mm)", fill = "Side"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

pine_jitter_plot
```
::::

:::: {.column width="40%"}
- `position_jitter(width = 0.15)` spreads points sideways so they don't overlap
- `seed = 42` — same jitter layout every time you render
- `alpha = 0.6` — semi-transparent, so overlapping points are still visible

::: callout-tip
Always show the raw points alongside a boxplot when *n* is small — every point matters, and a box alone can hide a bimodal or lopsided pattern.
:::
::::
:::::::

# Part 3 · Mapping Color by Group

::::::: columns
:::: {.column width="60%"}
```{r}
pine_team_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, color = group)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    size = 2.5, alpha = 0.8
  ) +
  labs(
    title = "Needle Length by Side, Colored by Team",
    x = "Side of tree", y = "Needle length (mm)", color = "Field team"
  ) +
  theme_minimal()

pine_team_plot
```
::::

:::: {.column width="40%"}
- `color = group` maps a **third variable** onto the plot — no new geom needed
- `ggplot` auto-builds the legend and picks colors for you
- Mapping inside `aes()` always means "vary this by data" — not "make everything this color"

::: callout-warning
**⚠️ Watch out!**

`color = "steelblue"` **outside** `aes()` sets one fixed color. `color = group` **inside** `aes()` maps colors to a variable. Mixing these up is one of the most common `ggplot` mistakes.
:::
::::
:::::::

# Part 4 · Mean ± SE — Plotted Directly

::::::: columns
:::: {.column width="60%"}
```{r}
pine_mean_se_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, color = n_s)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.3, size = 2
  ) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15, linewidth = 0.9
  ) +
  labs(
    title = "Mean ± SE Needle Length by Side",
    x = "Side of tree", y = "Needle length (mm)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

pine_mean_se_plot
```
::::

:::: {.column width="40%"}
**Two `stat_summary()` layers:**

| call | draws |
|---|---|
| `fun = mean` | one large point at the mean |
| `fun.data = mean_se` | error bars for ± 1 SE |

`stat_summary()` computes the mean and SE **for you**, straight from the raw data — no `summarize()` step required first.

::: callout-note
**✅ Key idea**

This is the exact same mean and SE you calculated by hand in Module 3 — now you can see them.
:::
::::
:::::::

# Part 5 · Faceting — One Panel Per Group

::::::: columns
:::: {.column width="60%"}
```{r}
pine_facet_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  facet_wrap(~group) +
  labs(
    title = "Needle Length by Side, Faceted by Team",
    x = "Side of tree", y = "Needle length (mm)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

pine_facet_plot
```
::::

:::: {.column width="40%"}
- `facet_wrap(~group)` gives each field team its **own small panel**
- Same x/y scale across panels — easy to compare
- Useful the moment "does every group show the same pattern?" is the real question

::: callout-tip
**🖐 Notice**

A faceted plot answers a different question than a single colored plot: not just "is there an overall pattern," but "does the pattern hold for *everyone*?"
:::
::::
:::::::

# Part 6 · A Quick Theme Tour

::::::: columns
:::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
pine_jitter_plot + theme_bw()
```

```{r}
#| fig-width: 3
#| fig-height: 2.6
pine_jitter_plot + theme_classic()
```
::::

:::: {.column width="40%"}
- `theme_minimal()` — light, few gridlines (what we've used so far)
- `theme_bw()` — white background, black-and-white frame
- `theme_classic()` — just x/y axis lines, no gridlines at all

::: callout-note
**✅ Key idea**

A theme changes *appearance* only — never the data or the statistics underneath. Pick one and use it consistently across a report.
:::
::::
:::::::

# Part 7 · Saving a Real Figure

::::::: columns
:::: {.column width="60%"}
```{r}
#| eval: false
ggsave(
  "figures/pine_needle_boxplot.png",
  plot   = pine_jitter_plot,
  width  = 3,
  height = 3,
  units  = "in",
  dpi    = 300
)
```

::: callout-warning
**⚠️ Watch out!**

`ggsave()` wants the **filename first**, then `plot =`. If you skip `plot =`, it saves whatever plot was drawn *last* — not necessarily the one you meant.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Always set `width`, `height`, `units`, and `dpi` explicitly. Letting `ggsave()` guess gives you a plot sized for your screen, not for a report or a poster.
:::
::::
:::::::

# Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Built boxplots with raw points overlaid, not hidden
- Mapped a third variable with `color =` / `fill =` inside `aes()`
- Plotted mean ± SE directly with `stat_summary()` — no separate summary table needed
- Faceted a plot with `facet_wrap()` to check every group at once
- Compared themes, and saved a real, correctly-sized figure with `ggsave()`

::: callout-tip
**🖐 Before next class**

Finish the worksheet: build a mean ± SE plot faceted by `group`, pick a theme, and save it to `figures/` at 3×3in, 300 dpi.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Where this leads**

You can now import, wrangle, describe, and visualize a dataset end to end. That full pipeline — raw data → clean → summarize → visualize — is the foundation every later statistical test builds on.
:::
::::
:::::::

# Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` loaded? Is `color`/`fill` inside `aes()` when it should be?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
