---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Nested ANOVA"
subtitle: "Hierarchical designs"
description: "Factorial vs. nested designs, the nested-ANOVA linear model and variance partitioning, and a naive-vs-correct worked example (sea urchin grazing and algae cover) using the traditional aov() approach — the modern mixed-model (lmer) alternative continues in the Linear Mixed Models lecture."
categories: [ANOVA]
order: 14
week: 9
author: "Bill Perry"

knitr:
  opts_chunk:
    paged.print: false
    collapse: true
format:
  html:
    downloads: [docx, pptx]  # This creates download links for all three
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true

editor:
  markdown:
    wrap: 72
---

```{r nested-anova-01}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
library(janitor)
library(tidyverse)

options(contrasts = c("contr.sum", "contr.poly"))
options(scipen = 999)  # no sci notation
```

## Where We Left Off

:::::: columns
:::: {.column width="60%"}
Covered last time — two-factor ANOVA:

- Example, linear model, analysis of variance
- Null hypotheses
- Interactions and main effects
- Unequal sample size
- Assumptions

::: callout-note
**✅ Key idea from last lecture**

In a factorial design, every level of B appears under every level of A.
Today: what happens when B is only ever nested *inside* one specific
level of A?
:::
::::

::: {.column width="40%"}
![XKCD comic listing p-value ranges with increasingly tongue-in-cheek
interpretations, from 'Highly Significant' at p under 0.03 down through
'On the edge of significance' and 'Hey, look at this interesting
subgroup analysis' at p \>= 0.1, captioned 'If all else fails, use
significant at a p\>0.05 level and hope no one notices' — a joke about
p-hacking and misusing significance
thresholds.](images/clipboard-1642768928.png){width="2.92in"}
:::
::::::

## Part 1 · Multifactor Designs {.divider}

## Two-Factor (2-Way) ANOVA, Recap

::::: columns
::: {.column width="60%"}
- We often consider more than 1 factor (independent categorical
  variables) to reduce unexplained variance and look at interactions
- 2-factor designs (2-way ANOVA) are very common in ecology
- Can have more factors (e.g., 3-way ANOVA) — interpretation gets tricky
- Most multifactor designs are: factorial, nested, or mixed
:::

::: {.column width="40%"}
![Diagram of a cow (representing manure/fertilizer) and a smiling sun
icon, each with an arrow pointing down to a small green seedling,
illustrating two factors that can each affect plant
growth.](images/clipboard-1173869376.png){width="2.92in"}
:::
:::::

## Factorial Versus Nested Designs

::::: columns
::: {.column width="60%"}
Consider two factors, A and B:

- **Factorial/crossed:** every level of B in every level of A
- **Nested/hierarchical:** levels of B occur only in 1 level of A
  - both fixed — rare
  - one fixed, one random
  - both random — rare in ecology, common in genetics
:::

::: {.column width="40%"}
![Diagram contrasting a crossed design, where both levels of Factor 1
(A, B) connect to every level of Factor 2 (a, b, c, d), against a nested
design below it, where each level of Factor 1 connects only to its own
subset of (random) Factor 2 levels, both feeding into individual
observations (Schielzth & Nakagawa
2013).](images/clipboard-758019042.png){width="2.92in"}
:::
:::::

## Part 2 · Factorial Designs — Examples {.divider}

## Factorial Designs Overview

::::: columns
::: {.column width="40%"}
Factorial designs: both factors are typically fixed (but not always).
:::

::: {.column width="60%"}
![Two grid diagrams of a factorial design with Factor A across the top
(4 levels) and Factor B down the side (3 levels): a compact 4x3 grid
marked 'XX' in every cell, and an expanded layout below showing each of
the 12 A-by-B combinations as its own column with two replicate marks
(X, X).](images/clipboard-723648250.png){width="3.96in"}
:::
:::::

## Factorial Design Example: Seedling Growth

::::: columns
::: {.column width="40%"}
Effects of light level on growth of seedlings of different size:

- 3 light levels (Factor A)
- 3 size classes (Factor B)
- 5 replicate seedlings in each cell
:::

::: {.column width="60%"}
![Photo of a small seedling above a table crossing Factor A (light: Low,
Medium, High) with Factor B (size: Small, Medium, Large), with each cell
showing 5 replicate seedling IDs (e.g., 1_LS...5_LS), illustrating the
seedling-growth factorial design
example.](images/clipboard-1057376698.png){width="3.96in"}
:::
:::::

## Factorial Design Example: Salamander Growth

::::: columns
::: {.column width="40%"}
Effects of food level and tadpole presence on larval salamander growth:

- 2 food levels (Factor A)
- Presence/absence of tadpoles (Factor B)
- 8 replicates in each cell
:::

::: {.column width="60%"}
![Photo of a larval salamander above a table crossing Factor A (food
level: Low, High) with Factor B (tadpoles: Absent, Present), with each
cell showing 8 replicate IDs (e.g., 1_LA...8_LA), illustrating the
salamander-growth factorial design
example.](images/clipboard-3697840818.png){width="3.96in"}
:::
:::::

## Factorial Design Example: Limpet Fecundity

::::: columns
::: {.column width="40%"}
Effect of season and density on limpet fecundity:

- 2 seasons (Factor A)
- 4 density treatments (Factor B)
- 3 replicates in each cell
:::

::: {.column width="60%"}
![Photo of limpets clustered on a rock above a table crossing Factor A
(season: Winter, Summer) with Factor B (density: 8, 15, 30, 45 per
plate), with each cell showing 3 replicate IDs (e.g., W.8.1...W.8.3),
illustrating the limpet-fecundity factorial design
example.](images/clipboard-1223732443.png){width="3.96in"}
:::
:::::

## Part 3 · Nested Designs — Examples {.divider}

## Nested Designs Overview

::::: columns
::: {.column width="40%"}
**Nested designs cover:** nested design examples, the linear model,
analysis of variance, null hypotheses, unbalanced designs, assumptions,
mixed-model ANOVA.

**Nested designs:** Factor A is usually fixed; Factor B is usually
random.
:::

::: {.column width="60%"}
![Grid layout with Factor A across the top (4 levels) and Factor B down
the side, each of the 4 levels of B uniquely numbered 1 through 12
nested within its own level of A, with two rows of X's marking replicate
observations under each B
level.](images/clipboard-3125256618.png){width="3.96in"}
:::
:::::

## Nested Design Example: Limpet Growth

::::: columns
::: {.column width="40%"}
Study on effects of enclosure size on limpet growth:

- 2 enclosure sizes (Factor A)
- 5 replicate enclosures (Factor B)
- 5 replicate limpets per enclosure
:::

::: {.column width="60%"}
![Photo of limpets on a rock above a table for Factor A (enclosure size:
Small, Large), each with 5 replicate enclosures (S1-S5 or L1-L5) and 5
replicate limpets per enclosure, illustrating the limpet-growth nested
design example.](images/clipboard-1867917958.png){width="3.96in"}
:::
:::::

## Nested Design Example: Reef Fish

::::: columns
::: {.column width="40%"}
Study on reef fish recruitment:

- 5 sites (Factor A)
- 6 transects at each site (Factor B)
- Replicate observations along each transect
:::

::: {.column width="60%"}
![Underwater photo of reef fish above coral, above a table for Factor A
(site: Site 1, Site 2), each with several transects (T1.1, T1.2, ...
T1.6) and replicate observations along each transect, illustrating the
reef-fish recruitment nested design
example.](images/clipboard-1621781711.png){width="3.96in"}
:::
:::::

## Nested Design Example: Sea Urchin Grazing

::::: columns
::: {.column width="40%"}
Effects of sea urchin grazing on biomass of filamentous algae:

- 4 urchin densities: 0%, 33%, 66%, 100% of natural density
- 4 patches of rocky bottom (3–4 m²) nested in each grazing level
- 5 replicate quadrats per patch

*This is our worked example for the rest of the lecture.*
:::

::: {.column width="60%"}
![Underwater photo of a black sea urchin on coral, above a table for
grazing level (None, Low, Medium, High) each with 4 patches nested
inside it and 5 replicate quadrats per patch, illustrating the sea
urchin grazing worked example used for the rest of this
lecture.](images/clipboard-3570106410.png){width="3.96in"}
:::
:::::

## Part 4 · The Nested Design Model {.divider}

## Nested Design: Linear Model Structure

::::: columns
::: {.column width="40%"}
Consider a nested design with:

- p levels of factor A (i = 1…p) — e.g., 4 grazing levels
- q levels of factor B (j = 1…q), nested within each level of A — e.g.,
  4 different patches per grazing level
- n replicates (k = 1…n) in each combination of A and B — e.g., 5
  replicate quadrats in each patch in each grazing level
:::

::: {.column width="60%"}
![Hierarchical tree diagram of the nested design structure: a top-level
factor A_1 branching into several B levels nested within it, each
branching further into individual replicate y observations, repeated for
a general level A_i.](images/clipboard-2460153431.png){width="3.96in"}
:::
:::::

## Calculating Means in a Nested Design

::::: columns
::: {.column width="40%"}
We can calculate several means:

- Overall mean (across all levels of A and B) = $\bar{y}$
- A mean for each level of A (across all levels of B in that A) =
  $\bar{y}_i$
- A mean for each level of B within each A = $\bar{y}_{j(i)}$
:::

::: {.column width="60%"}
![Grazing-level-by-patch table (grazing level: None, Low, Medium, High,
each with 4 patches numbered 1-16) with each patch cell listing 5
replicate quadrat IDs, the layout used to compute overall, A-level, and
B-within-A means.](images/clipboard-2098022400.png){width="3.96in"}
:::
:::::

## Nested Design Means, Visualized

![Full worked data table for the sea urchin example: for each grazing
level (A1-A4, 0-100%) the density mean is shown boxed in blue (e.g.,
39.2 for 0%), and for each patch nested within it the patch mean is
shown boxed in green, with the overall mean mu = 20.26 marked on the
right by a red bar spanning all
values.](images/clipboard-1474611051.png){width="3.96in"}

## The Nested Design Linear Model

:::::: columns
::: {.column width="60%"}
$$y_{ijk} = \mu + \alpha_i + \beta_{j(i)} + \varepsilon_{ijk}$$

- $y_{ijk}$ is the response variable — the value of the k-th replicate
  in the j-th level of B in the i-th level of A (e.g., algal biomass in
  the 3rd quadrat, in the 2nd patch, in the low-grazing treatment)
- $\mu$ is the overall mean (overall average algal biomass)
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

This looks almost identical to the two-way ANOVA model — the difference
is entirely in the subscript: $\beta_{j(i)}$, not $\beta_j$. B only
exists *inside* a specific A.
:::
::::
::::::

## Fixed and Random Effects in the Nested Model

:::::: columns
::: {.column width="60%"}
$$y_{ijk} = \mu + \alpha_i + \beta_{j(i)} + \varepsilon_{ijk}$$

- $\alpha_i$ is the fixed effect of factor A — the difference between
  the average biomass in all low-grazing-level quadrats and the overall
  mean
- $\beta_{j(i)}$ is the *random* effect of factor B nested within factor
  A — usually a random variable measuring variance among all possible
  levels of B within each level of A (variance among all possible
  patches that could have been used in the low-grazing treatment)
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

B being random is the whole point of nesting: you don't care about
*these specific* patches — you care about patch-to-patch variability in
general.
:::
::::
::::::

## The Error Term in the Nested Model

:::::: columns
::: {.column width="60%"}
$$y_{ijk} = \mu + \alpha_i + \beta_{j(i)} + \varepsilon_{ijk}$$

- $\varepsilon_{ijk}$ is the error term — the unexplained variance
  associated with the k-th replicate in the j-th level of B in the i-th
  level of A
- $\alpha_i$ is the effect of the i-th level of A: $\mu_i - \mu$
- $\varepsilon_{ijk}$: the difference between the observed algal biomass
  in the 3rd quadrat in the 2nd patch in the low-grazing treatment, and
  the *predicted* biomass (the average biomass in that 2nd patch in the
  low-grazing treatment)
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

The residual is measured against the *cell* mean (this specific patch),
not the treatment mean — that's what makes it "nested," arithmetically.
:::
::::
::::::

## Part 5 · Partitioning Variance {.divider}

## Analysis of Variance: Residual and Total

::::: columns
::: {.column width="40%"}
- $SS_{resid}$ is the difference between each observation and the mean
  for its level of factor B, summed over all observations
- $SS_{total} = SS_A + SS_B + SS_{resid}$
- SS can be turned into MS by dividing by the appropriate df
:::

::: {.column width="60%"}
![Nested ANOVA table listing Source (A, B(A), Residual, Total), each
with its sum-of-squares formula and degrees of freedom (p-1, p(q-1),
pq(n-1), pqn-1), with the Residual and Total rows outlined in red to
highlight how SS_residual is the deviation of each observation from its
own B(A) cell mean.](images/clipboard-2357756170.png){width="3.96in"}
:::
:::::

## Analysis of Variance: SSA

::::: columns
::: {.column width="40%"}
As before, we partition the variance in the response variable using SS.
$SS_A$ is the SS of differences between the mean in each level of A and
the overall mean.
:::

::: {.column width="60%"}
![The same nested ANOVA table with the A row outlined in red,
highlighting SS_A as the sum of squared deviations of each level-A mean
from the overall mean.](images/clipboard-1219870939.png){width="3.96in"}
:::
:::::

## Analysis of Variance: SSB

::::: columns
::: {.column width="40%"}
$SS_B$ is the SS of the difference between the mean in each level of B
and the mean of the corresponding level of A, summed across all levels
of A.
:::

::: {.column width="60%"}
![The same nested ANOVA table with the B(A) row outlined in red,
highlighting SS_B(A) as the sum of squared deviations of each
level-B-within-A mean from its corresponding level-A
mean.](images/clipboard-2467119661.png){width="3.96in"}
:::
:::::

## The Nested ANOVA Table

![The full sea urchin data table repeated (density means per grazing
level, patch means nested within each, overall mean y-bar = 20.26),
shown here alongside the nested ANOVA table structure it feeds
into.](images/clipboard-449121940.png){width="3.96in"}

## Part 6 · Hypotheses and Assumptions {.divider}

## Null Hypotheses: Factor A

::::: columns
::: {.column width="40%"}
Two hypotheses are tested on values of MS. **1. No effects of factor A**
— assuming A is fixed:

- $H_0(A)$: $\mu_1 = \mu_2 = ... = \mu_i = \mu$
- Same as in 1-factor ANOVA, using means from B factors nested within
  each level of A
- (no difference in algal biomass across all grazing levels:
  $\mu_{none} = \mu_{low} = \mu_{med} = \mu_{high}$)
:::

::: {.column width="60%"}
![Table for the 'A fixed, B random' case listing the expected mean
square and F-ratio formula for each source (A, B(A), Residual), showing
that A is tested against MS_B(A) rather than MS_Residual because B is a
random effect.](images/clipboard-1310685300.png){width="3.96in"}
:::
:::::

## Null Hypotheses: Factor B

::::: columns
::: {.column width="40%"}
**2. No effects of factor B nested in A** — assuming B is random:

- $H_0(B)$: $\sigma^2_\beta = 0$ (no variance added due to differences
  between all possible levels of B)
- (no variance added due to differences between patches)
:::

::: {.column width="60%"}
![Table for the 'A fixed, B random' case listing the expected mean
square and F-ratio formula for each source (A, B(A), Residual), showing
that A is tested against MS_B(A) rather than MS_Residual because B is a
random effect.](images/clipboard-1310685300.png){width="3.96in"}
:::
:::::

## Conclusions from the Analysis

::::: columns
::: {.column width="40%"}
**Conclusion:**

"Significant variation between replicate patches within each treatment,
but no significant difference in the amount of filamentous algae between
treatments."
:::

::: {.column width="60%"}
![Published nested ANOVA results table for the sea urchin study, showing
Treatment not significant (F = 2.72, p = 0.091) but Patches(treatment)
highly significant (F = 5.93, p \< 0.001), with a boxplot below showing
percent cover of filamentous algae declining from the None to High
urchin-density
treatments.](images/clipboard-3596669693.png){width="3.96in"}
:::
:::::

## Unbalanced Nested Designs

::::: columns
::: {.column width="40%"}
Unequal sample sizes can arise from:

- An uneven number of B levels within each A
- An uneven number of replicates within each level of B

Not a problem unless you have unequal variance or a large deviation from
normality.
:::

::: {.column width="60%"}
![The same grazing-level-by-patch layout table repeated, shown here in
the context of unequal sample sizes in unbalanced nested
designs.](images/clipboard-2098022400.png){width="3.96in"}
:::
:::::

## Nested Design Assumptions

:::::: columns
::: {.column width="60%"}
As usual, we assume: equal variance, normality, no outliers,
independence of observations.

Equal variance + normality need to be assessed at **both levels**:

- Since means for each level of B within each A are used for the H-test
  about A, need to assess whether *those means* meet normality and equal
  variance
- Examine residuals for the H-test about B
- Transformations can be used
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

This is the part people most often get wrong: checking assumptions on
the *raw data* isn't enough for a nested design — you need to check them
at the level each hypothesis test actually operates on.
:::
::::
::::::

## Part 7 · The Sea Urchin Example — Data {.divider}

## The Nested Design, the Hard Way

:::::: columns
::: {.column width="60%"}
This analysis examines the effects of varying sea urchin densities on
the percentage cover of filamentous algae. The experiment was designed
with:

- Four random patches
- Four urchin density treatments — **0%** (urchins removed), **33%**,
  **66%**, and **100%** of natural density (the unmanipulated control)
- Five replicate quadrats measured within each treatment-patch
  combination
:::

:::: {.column width="40%"}
::: callout-note
**📖 Reference**

Quinn & Keough (2002), *Experimental Design and Data Analysis for
Biologists* — this dataset and analysis follow their treatment of the
sea urchin/algae example.
:::
::::
::::::

## Data Overview

::::: columns
::: {.column width="50%"}
The data frame contains: `patch` (random patches 1–16 where treatments
were applied), `treat` (urchin density treatment), `quad` (replicate
quadrats within each treatment-patch combination), `algae` (percentage
cover of filamentous algae — the response variable).

```{r nested-anova-02}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_df <- read_csv("data/andrew.csv") %>% clean_names()

u_df <- u_df %>%
  # Map to Quinn & Keough's Table 9.1 densities. NOTE the file's dens_33 /
  # dens_66 labels are the reverse of the book's percentages -- verified by
  # matching patch means: dens_33 is the book's 66% group, dens_66 its 33%.
  mutate(treat = case_when(
         treat == "removal" ~ "0%",     # book  0% density, mean algae 39.2
         treat == "dens_66" ~ "33%",    # book 33% density, mean algae 19.0
         treat == "dens_33" ~ "66%",    # book 66% density, mean algae 21.6
         treat == "control" ~ "100%",   # book 100% density, mean algae 1.3
         TRUE ~ "other")) %>%
  mutate(treat = factor(treat,
                        levels = c("0%", "33%", "66%", "100%"),
                        labels = c("0%", "33%", "66%", "100%")),
         patch = as_factor(patch))

cat("Data\n\n")
head(u_df)
cat("\n\nTreatment levels:\n\n")
levels(u_df$treat)
```
:::

::: {.column width="50%"}
```{r nested-anova-03}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_summary <- u_df %>%
  group_by(treat) %>%
  summarise(
    n = n(),
    mean = mean(algae),
    sd = sd(algae),
    se = sd / sqrt(n),
    min = min(algae),
    max = max(algae)
  )
cat("Summary statistics by treatment:\n\n")
u_summary
```
:::
:::::

## A First Look at the Data

```{r nested-anova-04}
#| echo: false
#| fig-width: 3.6
#| fig-height: 3
#| fig-alt: "Scatterplot of algae cover by urchin grazing treatment, with individual quadrat points colored by patch and a black point marking the mean for each treatment, showing patch-to-patch scatter within each treatment level."
u_df %>% ggplot(aes(treat, algae)) +
  stat_summary(aes(treat, algae), fun = "mean", geom = "point", size = 3) +
  geom_point(aes(color = as.factor(patch)), position = position_dodge2(width = 0.3)) +
  theme_minimal(base_size = 9) +
  labs(color = "Patch")
```

## Part 8 · The Wrong Way — Naive Factorial Model {.divider}

## Manual Nested ANOVA — Setting Up

::::: columns
::: {.column width="60%"}
In this experimental design, `patch` is nested within `treat` because
each patch received only one treatment level — a hierarchical design
where the effect of patches must be considered *within* each treatment
(Quinn & Keough 2002).

We'll use a traditional nested ANOVA. You *could* make this easy by
averaging `algae` per patch and running a one-way ANOVA — but your power
is significantly reduced, and you lose estimates of variability among
patches.

**This first attempt is really a regular factorial ANOVA** — not
appropriate, since it's pseudoreplicated. Also, the F-value for Treat is
using $MS_{resid}$ when it should use $MS_{treat:patch}$.
:::

::: {.column width="40%"}
```{r nested-anova-05}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
u_fact_model <- aov(algae ~ treat + treat:patch, data = u_df)
cat("Factorial model summary:\n\n")
summary(u_fact_model)
```
:::
:::::

## Why the Factorial ANOVA Is Not Correct

::::: columns
::: {.column width="40%"}
- We can correct this by specifying the error term
- The MS needed for the F-value is really MS(patch nested in treatment),
  because those are the true replicates
- We can specify that — the problem is this approach can't really handle
  unbalanced designs
:::

::: {.column width="60%"}
```{r nested-anova-06}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
#| collapse: true
# Error(patch) -- not Error(treat:patch). Patch IDs are already unique
# (1-16) and each patch sits in exactly one treatment, so naming the patch
# alone is enough. Writing treat:patch asks for 4 x 16 = 64 combinations
# when only 16 exist, which makes R warn "Error() model is singular".
# Both give identical numbers; this one does it without the warning.
u_nested_model <- aov(algae ~ treat + Error(patch), data = u_df)
cat("Correctly specified nested ANOVA:\n\n")
summary(u_nested_model)
```
:::
:::::

## Same Numerator, Different Denominator

::::::: columns
:::: {.column width="60%"}
Both models used the **same 80 quadrats** and produced the **same sums of squares**. Only the F-test's denominator changed:

| | MS treat | ÷ MS error | = F | p |
|-----------|-----------|-----------|-----------|-----------|
| **Naive factorial** | 4810 | ÷ **299** (quadrats) | **16.11** | 0.00000007 |
| **Nested** | 4810 | ÷ **1770** (patches) | **2.72** | 0.091 |

::: callout-important
**The p-value moved by six orders of magnitude.**

The naive model tested the treatment against variation *between quadrats within a patch* — but quadrats are not what we assigned treatments to. **Patches are.**

There are 80 quadrats but only **16 patches**, four per treatment. The honest test has 12 denominator degrees of freedom, not 64.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ How to find the right denominator**

Ask: **what did I randomly assign the treatment to?**

That unit — patch here — supplies the error term. Everything measured *inside* it is a subsample, and subsamples tell you about measurement noise, not about treatment effects.
:::

::: callout-tip
**🖐 Notice `MS(patch)` is large**

1770 vs 299 — patches genuinely differ from one another. That variability is real biology, and the nested model makes you pay for it. The naive model pretended it was not there.
:::
::::
:::::::

## Where You Have Seen This Before

::::::: columns
:::: {.column width="60%"}
This is the **pine needle problem** again, one level up:

| Our pine study | This urchin study |
|------------------------|------------------------|
| 8 needles per side | 5 quadrats per patch |
| within one tree | within one patch |
| **4 trees** are the replicates | **16 patches** are the replicates |
| 64 rows, n = 4 | 80 rows, n = 16 |

In Week 1 we solved it by **averaging** each tree to one number. That works, and it is honest.

A nested ANOVA does something better: it keeps every quadrat, estimates *both* levels of variability, and still tests the treatment against the right error term.

::: callout-note
**✅ Key idea**

Averaging throws the subsamples away. Nesting *uses* them — to estimate how noisy your measurements are — while still refusing to count them as replicates.
:::
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ The limit of `aov(Error(...))`**

This syntax needs a (near-)balanced design. Ours is perfectly balanced — 4 patches per treatment, 5 quadrats per patch — so it works.

Real data usually is not. That is the problem **mixed models** solve, and it is where we go next.
:::
::::
:::::::

## How Do I Know If My Design Is Nested?

::::::: columns
:::: {.column width="60%"}
Three questions, in order. Ask them of **your own** data before you fit anything.

**1. What did I apply the treatment to?**
→ Whole patches. Not quadrats.

**2. Does every level of B appear under every level of A?**
→ Patch 7 only ever received the 66% density. It cannot appear under 100%.
→ **No ⇒ nested.** (Yes ⇒ crossed/factorial.)

**3. How many of those units do I have per treatment?**
→ 4 patches, not 20 quadrats. **That is your n.**

::: callout-important
**The answer to Q1 is your error term.**

```r
aov(algae ~ treat + Error(patch), data = u_df)
#                   ^^^^^^^^^^^^
#                   the thing treatments were assigned to
```
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 A quick tell**

If two rows of your data share a value for the thing you randomized, they are **not** independent replicates of it.

Two quadrats in patch 3 share patch 3. Two needles on one tree share that tree.
:::

::: callout-warning
**⚠️ Nested factors are random**

We are not asking "is patch 7 different from patch 12?" — those specific patches are of no interest. They are a **sample of possible patches**, so `patch` enters as a random effect.

The factor above it (`treat`) is fixed: those four densities are exactly the ones we care about.
:::
::::
:::::::

## Reporting a Nested ANOVA

::::::: columns
:::: {.column width="60%"}
::: callout-note
**How to report this result**

"Algal cover did not differ significantly among urchin density treatments (nested ANOVA, **F₃,₁₂ = 2.72, p = 0.091**), with patches treated as a random factor nested within treatment. Variation among patches within treatments was substantial, accounting for approximately **50% of the total variance** (σ²_patch = 294, σ²_within = 299; n = 4 patches per treatment, 5 quadrats per patch)."
:::

**The three things people get wrong:**

1. **The degrees of freedom.** It is **F(3, 12)** — the denominator df comes from the *patch* stratum, not the 64 on the "Within" line. Reporting F(3, 64) is reporting the wrong test.
2. **Silence about the nesting.** Say the word "nested" and name the random factor. A reader cannot tell from "F = 2.72" alone whether you did this correctly.
3. **Throwing away the patch variance.** It is a result, not a nuisance — here it is half the story.
::::

:::: {.column width="40%"}
```{r nested-anova-07}
#| echo: true
#| collapse: true
# Variance components, straight from
# the two mean squares
MS_patch  <- 1770.2   # Error: patch
MS_within <- 298.6    # Error: Within
n_quad    <- 5        # quadrats per patch

var_patch  <- (MS_patch - MS_within) / n_quad
var_within <- MS_within

cat("among patches:", round(var_patch, 1), "
")
cat("within patch: ", round(var_within, 1), "
")
cat("% among patches:",
    round(100 * var_patch /
          (var_patch + var_within), 1), "%
")
```

::: callout-tip
**🖐 Half the variance is *between patches***

Ignore that and you are pretending 80 quadrats are 80 independent observations. They are closer to 16.
:::
::::
:::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Told a **nested** design from a **crossed** one, and learned the three questions that decide it
- Found the right **error term** by asking what the treatment was applied to
- Watched the same data give **F = 16.11 (p < 0.0000001)** the wrong way and **F = 2.72 (p = 0.091)** the right way
- Estimated **variance components**, and found that half the variation lives between patches
- Wrote it up with the correct degrees of freedom

::: callout-important
**⚠️ The mistake this lecture exists to prevent**

Using the residual mean square as your denominator when the treatment was applied to something coarser than a single measurement. It inflates F, shrinks p, and it is the most common error in ecological field experiments (Hurlbert 1984).
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ This has been the plan all along**

| Week | Same idea, new tool |
|------|------|
| 1 | 64 needles, but 4 trees |
| 3 | two-stage `summarise()` |
| 4 | the paired t-test |
| 5 | power: 44 trees vs 7 |
| 9 | **nested ANOVA** |

Today it finally got its own name and its own model. **Sums of squares were never the problem — the denominator was.**
:::

::: callout-tip
**🖐 Before next class**

Take a dataset you actually care about — yours, or one from a paper — and answer the three questions on the first slide of this section. Write down what your true *n* is. If it is smaller than the number of rows, you have found a nested design.
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: are `patch` and `treat` **factors**? Are your nested unit IDs *unique* across the whole dataset (patch 1–16), or do they restart within each treatment (1–4, 1–4, …)? If they restart, R cannot tell nesting from crossing — renumber them, or write `Error(treat:patch)`
3.  Seeing **"Error() model is singular"**? You asked for more error-term combinations than your design contains — usually `Error(a:b)` where `Error(b)` is enough
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
