---
metadata-files:
  - ../../_templates/_lectures.yml
title: "Lecture: Graphing Effectively"
subtitle: "Boxplots, axis limits, a reusable themes file, multi-panel figures, and saving publication-ready output"
description: "Boxplots with jittered raw points, axis labels with labs(), mapping color and fill by group, setting axis limits honestly with coord_cartesian(), plotting mean ± SE with stat_summary(), sourcing a reusable themes file for publication-ready figures, combining panels with patchwork, and a real ggsave() drill — all on the pine needle data the class collected."
categories: [Graphing, ggplot2, R]
order: 4
week: 3
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 2.5
    fig-height: 2.5
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true
---

## Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Computed mean, median, SD, and SE — by hand and with `summarize()`
- Fixed the `length()` trap with `sum(!is.na())`
- Got tidy per-group tables with `group_by()` + `summarize()`

::: callout-note
**✅ Key idea from Describing Your Data**

You now have real numbers describing shady vs. sunny needles. Today we make those numbers **visible** — and save the result properly.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  The `ggplot` grammar, recapped
2.  Boxplots done right — with raw points
3.  Axis labels and titles with `labs()`
4.  Axis limits with `coord_cartesian()`
5.  Mapping color by group
6.  Mean ± SE with `stat_summary()`
7.  **A themes file you `source()` once** — publication-ready figures
8.  **`patchwork`** — many plots on one page
9.  `ggsave()`, for real this time

⏱ Histograms, density curves, faceting, and `scale_color_manual()` are in the notes as time allows — the worksheet has them either way.
:::
::::
:::::::

## Part 1 · The `ggplot` Grammar, Recapped {.divider}

## The `ggplot` Grammar, Recapped

::::::: columns
:::: {.column width="60%"}
Every `ggplot` is three pieces joined with `+`:

1.  **data** — which data frame
2.  **aes()** — which columns map to x, y, color, fill…
3.  **geom** — how to draw it

```{r graphing-effectively-01}
library(tidyverse)
library(readxl)

pine_df <- read_csv("data/pine_data.csv")
```

[📥 Download the data](data/pine_data.csv){.btn .btn-primary download="pine_data.csv"}

::: callout-note
**✅ Key idea**

Everything today is the *same three pieces* — we're just adding more layers and more polish on top of them.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, *The Analysis of Biological Data*, Ch. 2 — Displaying Data, is the core reference for this whole lecture.

[📖 R4DS §1 — Data visualization](https://r4ds.hadley.nz/data-visualize.html) and [§9 — Layers](https://r4ds.hadley.nz/layers.html)
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Recap from Getting Started**

```{r graphing-effectively-02}
#| fig-alt: "Scatterplot of pine needle length in millimeters by side of tree (side), with individual points shown for each side."
ggplot(pine_df, aes(x = side, y = needle_length_mm)) +
  geom_point()
```
:::
::::
:::::::

## Part 2 · Boxplot Anatomy {.divider}

## Boxplot Anatomy

::::: columns
::: {.column width="60%"}
```{r graphing-effectively-03}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree (shady vs. sunny), filled by side, with outlier points hidden. Titled 'Needle Length by Side of Tree'."
pine_box_plot <- ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side of tree",
    y = "Needle length (mm)",
    fill = "Side"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

pine_box_plot
```
:::

::: {.column width="40%"}
**Boxplot anatomy:**

- Middle line = **median**
- Box edges = 25th and 75th percentile (**IQR**)
- Whiskers = 1.5 × IQR
- Dots beyond whiskers = **outliers**

Whitlock & Schluter, Ch. 2 — Displaying Data, on the boxplot and the five-number summary it draws.

[📖 R4DS §9 — Layers](https://r4ds.hadley.nz/layers.html)
:::
:::::

## Boxplot + Raw Points

::: callout-note
**🔮 Predict first:** With only 8 needles per side per tree, does a boxplot alone show you everything? What might it hide?
:::

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-04}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with individual jittered data points overlaid on top of each box, so every raw observation is visible alongside the summary."
pine_jitter_plot <- ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15),
    alpha = 0.6,
    size = 2
  ) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side of tree",
    y = "Needle length (mm)",
    fill = "Side"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

pine_jitter_plot
```
:::

:::: {.column width="40%"}
- `position_jitter(width = 0.15)` spreads points sideways so they don't overlap
- `alpha = 0.6` — semi-transparent, so overlapping points are still visible

::: callout-tip
Always show the raw points alongside a boxplot when *n* is small — every point matters, and a box alone can hide a bimodal or lopsided pattern.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 2 — Displaying Data, "How to draw a good graph," makes exactly this argument: show the data, not only the summary.
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Parts 1, 2 and 3** — load the data, build the boxplot, add the raw points.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 2b · Axis Labels and Titles, Done Right {.divider}

## Before `labs()` — Never Ship This

::::::: columns
:::: {.column width="60%"}
```{r graphing-effectively-05}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with no labs() applied, so the axes are labeled with the raw column names side and needle_length_mm instead of readable text."
ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```

::: callout-important
The x-axis reads `side` and the y-axis reads `needle_length_mm` — the literal **column names**, not something a reader outside your lab would understand.
:::
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

A plot with default column-name axes is a plot that only makes sense to *you*, today. Six months from now, even you may not remember what `side` meant.
:::
::::
:::::::

## `labs()` — Every Piece of Text on the Plot

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-06}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with a full title, subtitle, readable axis labels, legend title, and data-source caption added via labs()."
ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    subtitle = "Four field teams, four trees",
    x = "Side of tree",
    y = "Needle length (mm)",
    fill = "Side",
    caption = "Data: pine_data.csv"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
**One function, five jobs:**

| Argument       | Controls                            |
|----------------|-------------------------------------|
| `title`        | main title                          |
| `subtitle`     | smaller text under the title        |
| `x` / `y`      | axis labels                         |
| `caption`      | small text, bottom-right            |
| `color`/`fill` | **legend title** for that aesthetic |

::: callout-note
**✅ Key idea**

`labs()` never touches the data or the statistics — only what a reader sees written on the plot. Change it freely.
:::

::: callout-note
**📖 Reference**

[📖 R4DS §11.1 — Labels](https://r4ds.hadley.nz/communication.html#labels)
:::
::::
::::::

## `labs()` — the Legend Title Trap

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-07}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Jittered scatterplot of needle length by side of tree, colored by field team, with the legend relabeled 'Field team'."
# Whichever aesthetic name you mapped, use that same name in labs()
ggplot(pine_df, aes(x = side, y = needle_length_mm, color = team)) +
  geom_point(position = position_jitter(width = 0.15)) +
  labs(color = "Field team") # matches aes(color = team)
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

The legend title comes from whatever you write for **the aesthetic you mapped** — `color = "Field team"` in `labs()` relabels a `color =` mapping. Write `fill = "..."` instead and it does nothing, because this plot never mapped `fill`.
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 3b** — write real titles, axis labels, and a legend title with `labs()`.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 2c · Setting the Axis Limits — `coord_cartesian()` {.divider}

## `coord_cartesian()` — Choosing What the Reader Sees

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-coord-1}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with the y-axis set to run from 0 to 25 mm using coord_cartesian, so the two boxes sit in the upper half of the panel against a zero baseline."
ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side of tree",
    y = "Needle length (mm)"
  ) +
  coord_cartesian(ylim = c(0, 25)) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
`ggplot` picks axis limits from your data. That is a fine default for exploring — and the wrong default the moment a reader has to judge *how big* a difference is.

- `coord_cartesian(ylim = c(0, 25))` — set the y-axis range
- `coord_cartesian(xlim = c(...))` — same for x
- Both at once: `coord_cartesian(xlim = c(...), ylim = c(...))`

::: callout-note
**✅ Key idea**

Starting a measurement axis at **zero** is what keeps a 1 mm difference from looking like a canyon. Decide the limits deliberately; don't inherit them.
:::
::::
::::::

## `coord_cartesian()` — Both Axes at Once

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-coord-2}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Scatterplot of needle width against needle length, colored by side of tree, with both axes set using coord_cartesian to x from 5 to 25 mm and y from 0.4 to 1.8 mm, a window that comfortably contains every point."
pine_scatter_plot <- ggplot(pine_df, aes(x = needle_length_mm, y = needle_width_mm,
                                         color = side)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(
    title = "Needle Width vs. Length",
    x = "Needle length (mm)",
    y = "Needle width (mm)",
    color = "Side"
  ) +
  coord_cartesian(xlim = c(5, 25), ylim = c(0.4, 1.8)) +
  theme_minimal(base_size = 9)

pine_scatter_plot
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

Two continuous axes, so both `xlim =` and `ylim =` are useful here. On the boxplot, `x` was a category — there is nothing to zoom.
:::

::: callout-note
**📖 Reference**

[📖 R4DS §11.4 — Scales](https://r4ds.hadley.nz/communication.html#scales)
:::
::::
::::::

## The Trap — `ylim()` Deletes Data, `coord_cartesian()` Does Not

::: callout-note
**🔮 Predict first:** Both plots below ask for a y-axis running 14 to 16 mm. One of them changes where the boxes are drawn. Which one, and why?
:::

::::::: columns
:::: {.column width="60%"}
```{r graphing-effectively-coord-3}
#| fig-width: 3
#| fig-height: 2.2
#| fig-alt: "Boxplot of needle length by side with ylim used to restrict the axis to 14 to 16 mm, which silently removes every needle outside that range and redraws the boxes from the surviving rows only."
# ❌ ylim() THROWS AWAY every row outside the range first
ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6) +
  ylim(14, 16) +
  labs(title = "ylim() — boxes redrawn from survivors") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "none")
```

```{r graphing-effectively-coord-4}
#| fig-width: 3
#| fig-height: 2.2
#| fig-alt: "The same boxplot with coord_cartesian used to zoom to 14 to 16 mm, which keeps every row in the calculation and simply crops the view, so the medians and quartiles are unchanged."
# ✅ coord_cartesian() keeps every row and just crops the view
ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6) +
  coord_cartesian(ylim = c(14, 16)) +
  labs(title = "coord_cartesian() — same boxes, zoomed") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "none")
```
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

`ylim(14, 16)` and `scale_y_continuous(limits = c(14, 16))` **filter the data before the geom draws it**. Every needle outside 14–16 mm is silently dropped, and the median, the quartiles, and the whiskers are all recomputed from what is left. R warns about "removed rows" and students read past it every time.

`coord_cartesian()` computes the statistics on **all** the data and then zooms the camera. The boxes do not move.
:::

::: callout-tip
**🖐 The rule**

Use `coord_cartesian()` to zoom. Use `filter()` when you actually mean to remove data — on purpose, in the data step, where you can see it.
:::

::: callout-note
**📖 Worth knowing — changing the axis *scale*, not just its limits**

Right-skewed data (a few very large values squashing everything else into a corner) is usually better read on a log axis:

``` r
+ scale_x_log10()      # the AXIS is logged; the data are untouched
```

That is different again from logging the data itself with `mutate(log_length = log(needle_length_mm))`, which changes the numbers every later calculation sees.

Nothing today needs it — our needles are not skewed. Try it on your scatterplot if you are curious, and we will come back to it when a distribution actually calls for it.
:::
::::
:::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 3c** — set your axis limits with `coord_cartesian()`. The `ylim()` trap stays on screen; you do not need to type it.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 3 · Mapping Color by Group {.divider}

## Mapping Color by Group

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-08}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Jittered scatterplot of needle length by side of tree, with points colored by field team, titled 'Needle Length by Side, Colored by Team'."
pine_team_plot <- ggplot(pine_df, aes(x = side, y = needle_length_mm, color = team)) +
  geom_point(
    position = position_jitter(width = 0.15),
    size = 2.5,
    alpha = 0.8
  ) +
  labs(
    title = "Needle Length by Side, Colored by Team",
    x = "Side of tree",
    y = "Needle length (mm)",
    color = "Field team"
  ) +
  theme_minimal(base_size = 9)

pine_team_plot
```
:::

:::: {.column width="40%"}
- `color = team` maps a **third variable** onto the plot — no new geom needed
- `ggplot` auto-builds the legend and picks colors for you
- Mapping inside `aes()` always means "vary this by data" — not "make everything this color"

::: callout-warning
**⚠️ Watch out!**

`color = "darkblue"` **outside** `aes()` sets one fixed color. `color = team` **inside** `aes()` maps colors to a variable. Mixing these up is one of the most common `ggplot` mistakes.
:::

::: callout-note
**📖 Worth knowing — where you put `aes()` matters too**

An `aes()` inside `ggplot()` is **global**: every layer inherits it. An `aes()` inside a single geom applies to **that layer only**:

``` r
# global — both layers colored by side
ggplot(pine_df, aes(x = side, y = needle_length_mm, color = side)) +
  geom_point() + stat_summary(fun = mean, geom = "point", size = 4)

# per-layer — points colored by team, the mean drawn in plain black
ggplot(pine_df, aes(x = side, y = needle_length_mm)) +
  geom_point(aes(color = team)) +
  stat_summary(fun = mean, geom = "point", size = 4)
```

Just know it exists for now — we will lean on it later, when a plot has to show raw data and a fitted model at the same time.
:::

::: callout-note
**📖 Reference**

[📖 R4DS §9.1 — Aesthetic mappings](https://r4ds.hadley.nz/layers.html#aesthetic-mappings)
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 4** — map `color =` / `fill =` to a third variable.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 3b · Histograms and Density — Seeing the Whole Distribution {.divider}

::: callout-note
**⏱ If time allows** — everything in this part is on the worksheet, so it is safe to work through on your own if we run short in class.
:::

## Histograms — the Shape Behind the Summary

::: callout-note
**🔮 Predict first:** In Describing Your Data we found the overall `mean(needle_length_mm)` ≈ 14.67 and the median ≈ 14.55 — nearly identical. What shape do you expect the full distribution to have — symmetric, or skewed?
:::

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-09}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Histogram of needle length in millimeters across all needles, with 2mm bins, showing the overall shape of the distribution."
ggplot(pine_df, aes(x = needle_length_mm)) +
  geom_histogram(binwidth = 2, fill = "darkblue", color = "white") +
  labs(
    title = "Distribution of Needle Length",
    x = "Needle length (mm)",
    y = "Count"
  ) +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
- `binwidth` sets the width of each bar in **data units** (here, 2mm)
- Too wide hides structure; too narrow shows noise — try a few values

::: callout-tip
**🖐 Try it yourself**

Change `binwidth = 2` to `binwidth = 0.5` and to `binwidth = 5`. Which tells the real story?
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 2 — Displaying Data, on frequency distributions and choosing bin width.

[📖 R4DS §10.3 — Variation](https://r4ds.hadley.nz/EDA.html#variation)
:::
::::
::::::

## Histograms by Group — `fill` and `facet_wrap()` Together

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-10}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Two stacked histograms of needle length, one panel for the shady side and one for the sunny side, faceted so each group's distribution is shown separately."
ggplot(pine_df, aes(x = needle_length_mm, fill = side)) +
  geom_histogram(binwidth = 2, color = "white", alpha = 0.7) +
  facet_wrap(~side, ncol = 1) +
  labs(x = "Needle length (mm)", y = "Count", fill = "Side") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Overlapping histograms on one panel are hard to read. Faceting stacks them into separate panels sharing an x-axis — same comparison, much clearer.
:::

::: callout-note
**📖 Reference**

[📖 R4DS §9.5 — Facets](https://r4ds.hadley.nz/layers.html#facets)
:::
::::
::::::

## Density Curves — a Smoothed Alternative

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-11}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Overlaid density curves of needle length for the shady and sunny sides, semi-transparent so both distributions are visible on one panel."
ggplot(pine_df, aes(x = needle_length_mm, fill = side)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Needle Length Density by Side",
    x = "Needle length (mm)",
    y = "Density",
    fill = "Side"
  ) +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
::: callout-tip
**📖 Histogram vs. density**

A density curve is a smoothed histogram, rescaled so the area under each curve sums to 1. It overlays cleanly for group comparisons — no faceting required, at the cost of hiding the raw bin counts.
:::

::: callout-note
**📖 Reference**

Gotelli & Ellison, *A Primer of Ecological Statistics*, Ch. 3 — Summary Statistics, on the shape of a distribution behind any mean and SD you report.
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 4b** — histograms and density curves. ⏱ Skip it if we are short; nothing later depends on it.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 4 · Mean ± SE — Plotted Directly {.divider}

## Mean ± SE — Plotted Directly

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-12}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Scatterplot of jittered raw needle-length points by side of tree, with a large point marking the mean and an error bar showing ± 1 standard error for each side."
pine_mean_se_plot <- ggplot(pine_df, aes(x = side, y = needle_length_mm, color = side)) +
  geom_point(
    position = position_jitter(width = 0.15),
    alpha = 0.3,
    size = 2
  ) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15,
    linewidth = 0.9
  ) +
  labs(
    title = "Mean ± SE Needle Length by Side",
    x = "Side of tree",
    y = "Needle length (mm)"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

pine_mean_se_plot
```
:::

:::: {.column width="40%"}
**Two `stat_summary()` layers:**

| call                 | draws                       |
|----------------------|-----------------------------|
| `fun = mean`         | one large point at the mean |
| `fun.data = mean_se` | error bars for ± 1 SE       |

`stat_summary()` computes the mean and SE **for you**, straight from the raw data — no `summarize()` step required first.

::: callout-warning
**⚠️ Convenient, but read the fine print**

"Straight from the raw data" means it used **all 64 needles**. Those are the *naive* error bars from Describing Your Data — the ones that treat 8 needles from one tree as 8 independent facts.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 4 — Estimating with Uncertainty, on what an error bar is an estimate of.\
Gotelli & Ellison, Ch. 6 — Designing Successful Field Studies, on why independence decides which rows you may average.
:::
::::
::::::

## The error bar is a claim — make it an honest one

::: callout-note
**🔮 Predict first:** We are about to draw the same two means, but with error bars built from the **4 tree means** instead of the 64 needles. Will the dots move? Will the bars?
:::

::::::: columns
:::: {.column width="60%"}
```{r graphing-effectively-13}
#| fig-width: 3.2
#| fig-height: 2.6
#| fig-alt: "Mean needle length by side of tree with error bars computed from the four team means rather than from all 64 needles, so the error bars are roughly twice as long as in the naive version."
team_means_df <- pine_df |>
  group_by(team, side) |>
  summarize(team_mean_mm = mean(needle_length_mm), .groups = "drop")

ggplot(team_means_df, aes(x = side, y = team_mean_mm, color = side)) +
  geom_point(size = 2, alpha = 0.6,
             position = position_jitter(width = 0.08)) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.15, linewidth = 0.9) +
  labs(title = "Mean ± SE, computed from 4 trees",
       x = "Side of tree", y = "Mean needle length (mm)") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```
::::

:::: {.column width="40%"}
::: callout-important
**Same dots. Longer bars.**

The means are identical (14.12 and 15.21). The SEs went from 0.49/0.46 to **1.01/0.79** — roughly double.

Each small point here is **one tree**, not one needle. That is the plot our design actually supports.
:::

::: callout-tip
**🖐 The habit**

Before you draw an error bar, ask what one point in it represents. `stat_summary()` will happily compute an SE over anything you hand it — including rows that aren't independent.
:::
::::
:::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 5** — mean ± SE with `stat_summary()`.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 5 · Faceting — One Panel Per Group {.divider}

::: callout-note
**⏱ If time allows** — the worksheet walks through this one too.
:::

## Faceting — One Panel Per Group

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-14}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplots of needle length by side of tree, faceted into a separate small panel for each field team, to compare whether the shady-vs-sunny pattern holds across teams."
pine_facet_plot <- ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  facet_wrap(~team) +
  labs(
    title = "Needle Length by Side, Faceted by Team",
    x = "Side of tree",
    y = "Needle length (mm)"
  ) +
  theme_minimal(base_size = 7) +
  theme(legend.position = "none")

pine_facet_plot
```
:::

:::: {.column width="40%"}
- `facet_wrap(~team)` gives each field team its **own small panel**
- Same x/y scale across panels — easy to compare
- Useful the moment "does every group show the same pattern?" is the real question

::: callout-tip
**🖐 Notice**

A faceted plot answers a different question than a single colored plot: not just "is there an overall pattern," but "does the pattern hold for *everyone*?"

[📖 R4DS §9.5 — Facets](https://r4ds.hadley.nz/layers.html#facets)
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 6** — facet by field team. ⏱ Skip it if we are short.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 6 · Themes — Write It Once, Use It Everywhere {.divider}

## The Built-in Themes

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-15}
#| fig-width: 3
#| fig-height: 2.4
#| fig-alt: "The boxplot-with-jittered-points figure re-rendered using theme_bw, a white background with a black-and-white frame."
pine_jitter_plot + theme_bw(base_size = 9)
```
:::

:::: {.column width="40%"}
- `theme_minimal()` — light, few gridlines (what we have used so far)
- `theme_bw()` — white background, black-and-white frame
- `theme_classic()` — just x/y axis lines, no gridlines

A theme changes **appearance only** — never the data or the statistics underneath.

::: callout-warning
**⚠️ The problem with the built-ins**

None of them is quite publication-ready, so you end up appending ten lines of `theme(axis.title = element_text(...), ...)` to every single plot — and retyping them in every script, slightly differently each time.
:::
::::
::::::

## The Themes File — Three Sizes, Written Once

::::::: columns
:::: {.column width="60%"}
Download **`r_themes_for_3_sizes.R`** from the activity page and save it in your **project folder**, next to your `.Rproj` file.

It defines three functions, each sized for a different final output:

| Function | Built for | Base font |
|---|---|---|
| `theme_small()` | ~3 × 3 in — a journal panel | 9 pt |
| `theme_regular()` | ~7 × 7 in — a report figure | 16 pt |
| `theme_large()` | ~16 × 16 in — a poster | 36 pt |

All three give you the same publication look: white background, a black panel border, black axis lines and ticks, bold axis titles, no gridlines.

::: callout-important
**Two steps, once each**

1. **Open the file and run it** (Ctrl/Cmd + Shift + Enter, or Source). Look at your Environment pane — `theme_small`, `theme_regular`, and `theme_large` appear as functions. That is all "running it" does: it defines them.
2. **From then on, `source()` it** at the top of any script that makes a figure. One line replaces the whole block of theme code you would otherwise paste into every script.
:::
::::

:::: {.column width="40%"}
```{r graphing-effectively-themes-1}
#| eval: false
# At the top of your script,
# with library(tidyverse):
source("r_themes_for_3_sizes.R")
```

::: callout-note
**✅ Why `source()` and not copy-paste**

`source()` runs that file top to bottom, exactly as if you had typed it. Fix a margin in the themes file once and **every** figure in every script picks up the change the next time you run it.

Copy-pasted theme code gives you fifteen slightly different themes and no way to fix them all.
:::

::: callout-warning
**⚠️ Watch out!**

`source()` needs the path to be right. With the file in your project root and an `.Rproj` open, `"r_themes_for_3_sizes.R"` is enough. Put it in a `scripts/` folder and it becomes `source("scripts/r_themes_for_3_sizes.R")`.
:::
::::
:::::::

```{r graphing-effectively-themes-setup}
#| echo: false
#| message: false
# Rendered from the content folder, so the file sits under data/ here.
# In your project it will be in the project root: source("r_themes_for_3_sizes.R")
source("data/r_themes_for_3_sizes.R")
```

## Using the Themes

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-themes-2}
#| fig-width: 3.2
#| fig-height: 2.8
#| fig-alt: "The needle length boxplot rendered with the custom theme_small function: white background, black panel border, bold axis titles, and no gridlines."
pine_theme_plot <- ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side of tree",
    y = "Needle length (mm)"
  ) +
  coord_cartesian(ylim = c(0, 25)) +
  theme_small() +
  theme(legend.position = "none")

pine_theme_plot
```
:::

:::: {.column width="40%"}
It is one more `+` layer, exactly like `theme_minimal()`:

``` r
+ theme_small()      # 3 × 3 in
+ theme_regular()    # 7 × 7 in
+ theme_large()      # poster
```

::: callout-tip
**🖐 Match the theme to the size you will save**

Pick the theme for the figure's **final printed size**, then pass that same size to `ggsave()`. A `theme_large()` plot saved at 3 × 3 in is unreadable, and `theme_small()` on a poster disappears.
:::

::: callout-note
**Still your plot**

Anything you add after the theme still wins — `theme_small() + theme(legend.position = "none")` keeps the theme and drops the legend.
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Parts 7 and 7b** — try the built-in themes, then download, run, and `source()` the themes file.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 7 · Many Plots on One Page — `patchwork` {.divider}

## `patchwork` — `+` for Whole Plots

::::::: columns
:::: {.column width="60%"}
```{r graphing-effectively-patchwork-1}
#| eval: false
install.packages("patchwork")   # once per computer
```

```{r graphing-effectively-patchwork-2}
#| message: false
library(patchwork)
```

Save each plot into an object, then combine them:

``` r
plot_a + plot_b     # side by side
plot_a / plot_b     # stacked
```

::: callout-note
**✅ Key idea**

`patchwork` reuses the operators you already know. `+` puts plots **beside** each other, `/` puts one **over** the other — and the result is still a plot object you can save.
:::
::::

:::: {.column width="40%"}
```{r graphing-effectively-patchwork-3}
#| fig-width: 5
#| fig-height: 2.4
#| fig-alt: "Two panels side by side: the boxplot with jittered raw points on the left and the mean plus or minus standard error plot on the right."
pine_jitter_plot + pine_mean_se_plot
```
::::
:::::::

## `patchwork` — Building a Real Figure

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-patchwork-4}
#| fig-width: 5
#| fig-height: 4
#| fig-alt: "A three-panel figure: the boxplot and the mean plus or minus standard error plot side by side on the top row, with the width-versus-length scatterplot spanning the bottom row, each panel tagged A, B, and C."
combined_plot <- (pine_jitter_plot + pine_mean_se_plot) / pine_scatter_plot +
  plot_annotation(tag_levels = "A")

combined_plot
```
:::

:::: {.column width="40%"}
- Parentheses group plots, exactly like arithmetic: `(a + b) / c` is two on top, one across the bottom
- `plot_annotation(tag_levels = "A")` tags the panels **A**, **B**, **C** — what a figure caption refers to
- `combined_plot` is one object, so `ggsave()` saves the whole multi-panel figure in one call

::: callout-tip
**🖐 Size it for the whole figure**

A three-panel figure needs roughly three panels' worth of space. Give `ggsave()` a `width`/`height` for the **combined** plot, not for one panel.
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 7c** — combine your plots with `patchwork`.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 8 · Saving a Real Figure {.divider}

## Saving a Real Figure

::::::: columns
:::: {.column width="60%"}
```{r graphing-effectively-17}
#| eval: false
# One panel, built with theme_small() — save it at the size the theme was written for
ggsave(
  "figures/pine_needle_boxplot.png",
  plot = pine_theme_plot,
  width = 3,
  height = 3,
  units = "in",
  dpi = 300
)

# The three-panel patchwork figure — bigger canvas, one call
ggsave(
  "figures/pine_needle_combined.png",
  plot = combined_plot,
  width = 7,
  height = 6,
  units = "in",
  dpi = 300
)
```

::: callout-warning
**⚠️ Watch out!**

`ggsave()` wants the **filename first**, then `plot =`. If you skip `plot =`, it saves whatever plot was drawn *last* — not necessarily the one you meant.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Always set `width`, `height`, `units`, and `dpi` explicitly. Letting `ggsave()` guess gives you a plot sized for your screen, not for a report or a poster.
:::

::: callout-note
**📖 Reference**

Whitlock & Schluter, Ch. 2 — Displaying Data, on what a figure has to carry to stand on its own in a report.

[📖 R4DS §11 — Communication](https://r4ds.hadley.nz/communication.html)
:::
::::
:::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 8** — save both figures with `ggsave()`.

Come back when it runs. If it does not run, that is the interesting part — bring the error.
:::

## Part 9 · Custom Colors and Legends — `scale_color_manual()` {.divider}

## `scale_color_manual()` — Choosing Your Own Colors

::: callout-tip
**🚀 Advanced / extra work** — `ggplot`'s default colors are fine for exploring data, but a real figure often needs specific colors (journal style, colorblind-safe palettes, matching a poster). This is how you take control.
:::

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-18}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Jittered scatterplot of needle length by side of tree with a black point marking each side's mean, using custom colors (blue for shady, orange for sunny) and a relabeled legend set via scale_color_manual()."
ggplot(pine_df, aes(x = side, y = needle_length_mm, color = side)) +
  geom_point(
    position = position_jitter(width = 0.15),
    size = 2.5
  ) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  scale_color_manual(
    name = "Side of tree",
    labels = c(shady = "Shady side", sunny = "Sunny side"),
    values = c(shady = "#2c7fb8", sunny = "#d95f0e")
  ) +
  labs(x = "Side of tree", y = "Needle length (mm)") +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
**Three arguments, matched by name:**

| Argument | Does what                       |
|----------|---------------------------------|
| `name`   | the legend title                |
| `labels` | text shown for each level       |
| `values` | the actual color for each level |

::: callout-note
**✅ Key idea**

`labels` and `values` are both **named vectors**, keyed by the *raw values in your data* (`shady`, `sunny`) — not by the pretty text you want to display. `ggplot` looks up each raw value and substitutes the label/color you gave it.
:::

::: callout-tip
**📖 Why these two colors?**

`#2c7fb8` (blue) and `#d95f0e` (orange) are a **blue–orange** pair, the safest two-colour choice for colour-vision deficiency: red–green confusion, which affects roughly 8% of men, leaves the blue–orange axis intact.

Don't take that on faith — check any pair you invent with a simulator (the `colorblindr` or `dichromat` packages), or start from a palette that was built for it: `scale_color_viridis_d()` or `RColorBrewer::brewer.pal(n, "Dark2")`.

[📖 R4DS §11.4 — Scales](https://r4ds.hadley.nz/communication.html#scales), including the ColorBrewer palettes.
:::
::::
::::::

## `scale_color_manual()` — Getting the Names Wrong

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-19}
#| eval: false
# Typo: "sun" instead of the real value "sunny"
ggplot(pine_df, aes(x = side, y = needle_length_mm, color = side)) +
  geom_point() +
  scale_color_manual(
    values = c(shady = "#2c7fb8", sun = "#d95f0e")
  )
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

- Every level that appears in your data **must** get an entry in `values`
  - (and in `labels`, if you supply it), spelled exactly as it appears in the column
  - check with `unique(pine_df$side)` first if you are not sure
  - A missing or misspelled level draws as grey with a warning.
:::
::::
::::::

## `scale_fill_manual()` — the `fill` Twin

:::::: columns
::: {.column width="60%"}
```{r graphing-effectively-20}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree using custom fill colors (blue for shady, orange for sunny) and a relabeled legend set via scale_fill_manual()."
ggplot(pine_df, aes(x = side, y = needle_length_mm, fill = side)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(
    name = "Side of tree",
    labels = c(shady = "Shady side", sunny = "Sunny side"),
    values = c(shady = "#2c7fb8", sunny = "#d95f0e")
  ) +
  labs(x = "Side of tree", y = "Needle length (mm)") +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

Same three arguments, same pattern — `scale_fill_manual()` is `scale_color_manual()` for the `fill` aesthetic instead of `color`. A boxplot's box is `fill`; its jittered points (if you added `geom_point(color = ...)`) would be `color`. A plot can need both at once.
:::
::::
::::::

## ▶ Your Turn — Code It Now

::: callout-important
**Stop here and write code.**

Worksheet **Part 8b** — 🚀 optional bonus: pick your own colors with `scale_color_manual()`, and check them against a colorblind-safe palette.
:::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Built boxplots with raw points overlaid, not hidden
- Wrote real axis labels, titles, and legend titles with `labs()` — never shipped a default column name
- Mapped a third variable with `color =` / `fill =` inside `aes()`
- Set axis limits deliberately with `coord_cartesian()` — and learned that `ylim()` deletes rows while `coord_cartesian()` only zooms
- Plotted mean ± SE directly with `stat_summary()` — and saw that an error bar is only as honest as the rows you feed it
- **`source()`d a themes file** so `theme_small()` / `theme_regular()` / `theme_large()` are one `+` layer away in every script, forever
- **Combined panels with `patchwork`** — `a + b`, `a / b`, `(a + b) / c`, tagged A/B/C
- Saved a real, correctly-sized figure with `ggsave()`
- *(As time allowed)* Histograms, density curves, `facet_wrap()`, and custom colors with `scale_color_manual()`

::: callout-tip
**🖐 Before next class**

Finish the worksheet: source the themes file, build a two-panel `patchwork` figure with `coord_cartesian()` limits you chose on purpose, and save it to `figures/` at 300 dpi.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Where this leads**

You can now import, wrangle, describe, and visualize a dataset end to end. That full pipeline — raw data → clean → summarize → visualize — is the foundation every later statistical test builds on.
:::
::::
:::::::

## Readings

::::::: columns
:::: {.column width="60%"}
**Before next class** — chapter PDFs are posted on **Canvas**:

- **Whitlock & Schluter**, *The Analysis of Biological Data* — **Ch. 2**, Displaying Data. Read "How to draw a good graph" closely: it is the standard every figure you hand in this term is held to
- **Gotelli & Ellison**, *A Primer of Ecological Statistics* — **Ch. 3**, Summary Statistics, for the distributions behind the boxplots and histograms

**If you want more:**

- Whitlock & Schluter, Ch. 4 — Estimating with Uncertainty, on what an error bar actually claims

::: callout-note
**✅ Key idea**

W&S tells you *which* graph is honest. R4DS tells you *how* to draw it. You need both.
:::
::::

:::: {.column width="40%"}
::: callout-note
**📖 R for Data Science (2e)**

Free online at [r4ds.hadley.nz](https://r4ds.hadley.nz/):

- [§1 — Data visualization](https://r4ds.hadley.nz/data-visualize.html)
- [§9 — Layers](https://r4ds.hadley.nz/layers.html) — geoms, aesthetics, positions, facets
- [§10 — Exploratory data analysis](https://r4ds.hadley.nz/EDA.html) — histograms and distribution shape
- [§11 — Communication](https://r4ds.hadley.nz/communication.html) — labels, scales, themes, saving
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` loaded? Is `color`/`fill` inside `aes()` when it should be?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

[📖 R4DS §8 — Workflow: getting help](https://r4ds.hadley.nz/workflow-help.html)

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
