---
title: "Lecture: Graphing Effectively"
subtitle: "Boxplots, group comparisons, and saving a publication-ready figure"
description: "Boxplots with jittered raw points, writing axis labels and titles properly with labs(), mapping color and fill by group, histograms and density curves, plotting mean ± SE directly with stat_summary(), faceting, a quick theme tour, custom colors and legends with scale_color_manual(), and a real ggsave() drill — all on the pine needle dataset."
categories: [Graphing, ggplot2, R]
order: 4
week: 3
author: "Bill Perry"

format:
  html:
    downloads: [docx, pptx]
  revealjs:
    output-ext: "slides.html"
  docx:
    reference-doc: ../../ms_templates/custom-reference-compact.docx
    fig-width: 3
    fig-height: 3
    fig-dpi: 300
    dpi: 300            # tell Pandoc the real render dpi so it sizes figures correctly
  pptx: default
execute:
  echo: true
---

## Where We Left Off

::::::: columns
:::: {.column width="60%"}
Last time you:

- Computed mean, median, SD, and SE — by hand and with `summarize()`
- Fixed the `length()` trap with `sum(!is.na())`
- Got tidy per-group tables with `group_by()` + `summarize()`

::: callout-note
**✅ Key idea from Describing Your Data**

You now have real numbers describing shady vs. sunny needles. Today we make those numbers **visible** — and save the result properly.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Today's roadmap**

1.  The `ggplot` grammar, recapped
2.  Boxplots done right — with raw points
3.  Axis labels and titles, done right
4.  Mapping color by group
5.  Histograms and density curves
6.  Plotting mean ± SE directly
7.  Faceting and a quick theme tour
8.  Custom colors and legends with `scale_color_manual()`
9.  `ggsave()`, for real this time
:::
::::
:::::::

## Part 1 · The `ggplot` Grammar, Recapped {.divider}

## The `ggplot` Grammar, Recapped

::::::: columns
:::: {.column width="60%"}
Every `ggplot` is three pieces joined with `+`:

1.  **data** — which data frame
2.  **aes()** — which columns map to x, y, color, fill…
3.  **geom** — how to draw it

```{r}
library(tidyverse)
library(readxl)

pine_df <- read_csv("data/pine_needles.csv")
```

::: callout-note
**✅ Key idea**

Everything today is the *same three pieces* — we're just adding more layers and more polish on top of them.
:::
::::

:::: {.column width="40%"}
::: callout-tip
**🖐 Recap from Getting Started**

```{r}
#| fig-alt: "Scatterplot of pine needle length in millimeters by side of tree (n_s), with individual points shown for each side."
ggplot(pine_df, aes(x = n_s, y = length_mm)) +
  geom_point()
```
:::
::::
:::::::

## Part 2 · Boxplot Anatomy {.divider}

## Boxplot Anatomy

::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree (shady vs. sunny), filled by side, with outlier points hidden. Titled 'Needle Length by Side of Tree'."
pine_box_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side (n = shady, s = sunny)",
    y = "Needle length (mm)",
    fill = "Side"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

pine_box_plot
```
:::

::: {.column width="40%"}
**Boxplot anatomy:**

- Middle line = **median**
- Box edges = 25th and 75th percentile (**IQR**)
- Whiskers = 1.5 × IQR
- Dots beyond whiskers = **outliers**

📖 R4DS §9 — Layers
:::
:::::

## Boxplot + Raw Points

::: callout-note
**🔮 Predict first:** With only 6 needles per side per tree, does a boxplot alone show you everything? What might it hide?
:::

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with individual jittered data points overlaid on top of each box, so every raw observation is visible alongside the summary."
pine_jitter_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.6,
    size = 2
  ) +
  labs(
    title = "Needle Length by Side of Tree",
    x = "Side of tree",
    y = "Needle length (mm)",
    fill = "Side"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

pine_jitter_plot
```
:::

:::: {.column width="40%"}
- `position_jitter(width = 0.15)` spreads points sideways so they don't overlap
- `seed = 42` — same jitter layout every time you render
- `alpha = 0.6` — semi-transparent, so overlapping points are still visible

::: callout-tip
Always show the raw points alongside a boxplot when *n* is small — every point matters, and a box alone can hide a bimodal or lopsided pattern.
:::
::::
::::::

## Part 2b · Axis Labels and Titles, Done Right {.divider}

## Before `labs()` — Never Ship This

::::::: columns
:::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with no labs() applied, so the axes are labeled with the raw column names n_s and length_mm instead of readable text."
ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```

::: callout-important
The x-axis reads `n_s` and the y-axis reads `length_mm` — the literal **column names**, not something a reader outside your lab would understand.
:::
::::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

A plot with default column-name axes is a plot that only makes sense to *you*, today. Six months from now, even you may not remember what `n_s` meant.
:::
::::
:::::::

## `labs()` — Every Piece of Text on the Plot

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree with a full title, subtitle, readable axis labels, legend title, and data-source caption added via labs()."
ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  labs(
    title = "Needle Length by Side of Tree",
    subtitle = "Four field teams, four trees",
    x = "Side of tree (shady = n, sunny = s)",
    y = "Needle length (mm)",
    fill = "Side",
    caption = "Data: pine_needles.csv"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
**One function, five jobs:**

| Argument       | Controls                            |
|----------------|-------------------------------------|
| `title`        | main title                          |
| `subtitle`     | smaller text under the title        |
| `x` / `y`      | axis labels                         |
| `caption`      | small text, bottom-right            |
| `color`/`fill` | **legend title** for that aesthetic |

::: callout-note
**✅ Key idea**

`labs()` never touches the data or the statistics — only what a reader sees written on the plot. Change it freely.
:::
::::
::::::

## `labs()` — the Legend Title Trap

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Jittered scatterplot of needle length by side of tree, colored by field team, with the legend relabeled 'Field team'."
# Whichever aesthetic name you mapped, use that same name in labs()
ggplot(pine_df, aes(x = n_s, y = length_mm, color = group)) +
  geom_point(position = position_jitter(width = 0.15, seed = 42)) +
  labs(color = "Field team") # matches aes(color = group)
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

The legend title comes from whatever you write for **the aesthetic you mapped** — `color = "Field team"` in `labs()` relabels a `color =` mapping. Write `fill = "..."` instead and it does nothing, because this plot never mapped `fill`.
:::
::::
::::::

## Part 3 · Mapping Color by Group {.divider}

## Mapping Color by Group

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Jittered scatterplot of needle length by side of tree, with points colored by field team, titled 'Needle Length by Side, Colored by Team'."
pine_team_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, color = group)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    size = 2.5,
    alpha = 0.8
  ) +
  labs(
    title = "Needle Length by Side, Colored by Team",
    x = "Side of tree",
    y = "Needle length (mm)",
    color = "Field team"
  ) +
  theme_minimal(base_size = 9)

pine_team_plot
```
:::

:::: {.column width="40%"}
- `color = group` maps a **third variable** onto the plot — no new geom needed
- `ggplot` auto-builds the legend and picks colors for you
- Mapping inside `aes()` always means "vary this by data" — not "make everything this color"

::: callout-warning
**⚠️ Watch out!**

`color = "darkblue"` **outside** `aes()` sets one fixed color. `color = group` **inside** `aes()` maps colors to a variable. Mixing these up is one of the most common `ggplot` mistakes.
:::
::::
::::::

## Part 3b · Histograms and Density — Seeing the Whole Distribution {.divider}

## Histograms — the Shape Behind the Summary

::: callout-note
**🔮 Predict first:** We found `mean(length_mm)` ≈ 17.7 in Describing Your Data, with a fairly small gap from the median. What shape do you expect the full distribution to have — symmetric, or skewed?
:::

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Histogram of needle length in millimeters across all needles, with 2mm bins, showing the overall shape of the distribution."
ggplot(pine_df, aes(x = length_mm)) +
  geom_histogram(binwidth = 2, fill = "darkblue", color = "white") +
  labs(
    title = "Distribution of Needle Length",
    x = "Needle length (mm)",
    y = "Count"
  ) +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
- `binwidth` sets the width of each bar in **data units** (here, 2mm)
- Too wide hides structure; too narrow shows noise — try a few values

::: callout-tip
**🖐 Try it yourself**

Change `binwidth = 2` to `binwidth = 0.5` and to `binwidth = 5`. Which tells the real story?
:::
::::
::::::

## Histograms by Group — `fill` and `facet_wrap()` Together

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Two stacked histograms of needle length, one panel for the shady side and one for the sunny side, faceted so each group's distribution is shown separately."
ggplot(pine_df, aes(x = length_mm, fill = n_s)) +
  geom_histogram(binwidth = 2, color = "white", alpha = 0.7) +
  facet_wrap(~n_s, ncol = 1) +
  labs(x = "Needle length (mm)", y = "Count", fill = "Side") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "none")
```
:::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Overlapping histograms on one panel are hard to read. Faceting stacks them into separate panels sharing an x-axis — same comparison, much clearer.
:::
::::
::::::

## Density Curves — a Smoothed Alternative

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Overlaid density curves of needle length for the shady and sunny sides, semi-transparent so both distributions are visible on one panel."
ggplot(pine_df, aes(x = length_mm, fill = n_s)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Needle Length Density by Side",
    x = "Needle length (mm)",
    y = "Density",
    fill = "Side"
  ) +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
::: callout-tip
**📖 Histogram vs. density**

A density curve is a smoothed histogram, rescaled so the area under each curve sums to 1. It overlays cleanly for group comparisons — no faceting required, at the cost of hiding the raw bin counts.
:::
::::
::::::

## Part 4 · Mean ± SE — Plotted Directly {.divider}

## Mean ± SE — Plotted Directly

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Scatterplot of jittered raw needle-length points by side of tree, with a large point marking the mean and an error bar showing ± 1 standard error for each side."
pine_mean_se_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, color = n_s)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    alpha = 0.3,
    size = 2
  ) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15,
    linewidth = 0.9
  ) +
  labs(
    title = "Mean ± SE Needle Length by Side",
    x = "Side of tree",
    y = "Needle length (mm)"
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

pine_mean_se_plot
```
:::

:::: {.column width="40%"}
**Two `stat_summary()` layers:**

| call                 | draws                       |
|----------------------|-----------------------------|
| `fun = mean`         | one large point at the mean |
| `fun.data = mean_se` | error bars for ± 1 SE       |

`stat_summary()` computes the mean and SE **for you**, straight from the raw data — no `summarize()` step required first.

::: callout-note
**✅ Key idea**

This is the exact same mean and SE you calculated by hand in Describing Your Data — now you can see them.
:::
::::
::::::

## Part 5 · Faceting — One Panel Per Group {.divider}

## Faceting — One Panel Per Group

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplots of needle length by side of tree, faceted into a separate small panel for each field team, to compare whether the shady-vs-sunny pattern holds across teams."
pine_facet_plot <- ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  facet_wrap(~group) +
  labs(
    title = "Needle Length by Side, Faceted by Team",
    x = "Side of tree",
    y = "Needle length (mm)"
  ) +
  theme_minimal(base_size = 7) +
  theme(legend.position = "none")

pine_facet_plot
```
:::

:::: {.column width="40%"}
- `facet_wrap(~group)` gives each field team its **own small panel**
- Same x/y scale across panels — easy to compare
- Useful the moment "does every group show the same pattern?" is the real question

::: callout-tip
**🖐 Notice**

A faceted plot answers a different question than a single colored plot: not just "is there an overall pattern," but "does the pattern hold for *everyone*?"
:::
::::
::::::

## Part 6 · A Quick Theme Tour {.divider}

## A Quick Theme Tour

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "The boxplot-with-jittered-points figure re-rendered using theme_bw, a white background with a black-and-white frame."
pine_jitter_plot + theme_bw(base_size = 9)
```

```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "The boxplot-with-jittered-points figure re-rendered using theme_classic, showing only x and y axis lines with no gridlines."
pine_jitter_plot + theme_classic(base_size = 9)
```
:::

:::: {.column width="40%"}
- `theme_minimal()` — light, few gridlines (what we've used so far)
- `theme_bw()` — white background, black-and-white frame
- `theme_classic()` — just x/y axis lines, no gridlines at all

::: callout-note
**✅ Key idea**

A theme changes *appearance* only — never the data or the statistics underneath. Pick one and use it consistently across a report.
:::
::::
::::::

## Part 7 · Saving a Real Figure {.divider}

## Saving a Real Figure

::::::: columns
:::: {.column width="60%"}
```{r}
#| eval: false
ggsave(
  "figures/pine_needle_boxplot.png",
  plot = pine_jitter_plot,
  width = 3,
  height = 3,
  units = "in",
  dpi = 300
)
```

::: callout-warning
**⚠️ Watch out!**

`ggsave()` wants the **filename first**, then `plot =`. If you skip `plot =`, it saves whatever plot was drawn *last* — not necessarily the one you meant.
:::
::::

:::: {.column width="40%"}
::: callout-note
**✅ Key idea**

Always set `width`, `height`, `units`, and `dpi` explicitly. Letting `ggsave()` guess gives you a plot sized for your screen, not for a report or a poster.
:::
::::
:::::::

## Part 8 · Custom Colors and Legends — `scale_color_manual()` {.divider}

## `scale_color_manual()` — Choosing Your Own Colors

::: callout-tip
**🚀 Advanced / extra work** — `ggplot`'s default colors are fine for exploring data, but a real figure often needs specific colors (journal style, colorblind-safe palettes, matching a poster). This is how you take control.
:::

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Jittered scatterplot of needle length by side of tree with a black point marking each side's mean, using custom colors (blue for shady, orange for sunny) and a relabeled legend set via scale_color_manual()."
ggplot(pine_df, aes(x = n_s, y = length_mm, color = n_s)) +
  geom_point(
    position = position_jitter(width = 0.15, seed = 42),
    size = 2.5
  ) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  scale_color_manual(
    name = "Side of tree",
    labels = c(n = "Shady (sheltered)", s = "Sunny"),
    values = c(n = "#2c7fb8", s = "#d95f0e")
  ) +
  labs(x = "Side of tree", y = "Needle length (mm)") +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
**Three arguments, matched by name:**

| Argument | Does what                       |
|----------|---------------------------------|
| `name`   | the legend title                |
| `labels` | text shown for each level       |
| `values` | the actual color for each level |

::: callout-note
**✅ Key idea**

`labels` and `values` are both **named vectors**, keyed by the *raw values in your data* (`n`, `s`) — not by the pretty text you want to display. `ggplot` looks up each raw value and substitutes the label/color you gave it.
:::
::::
::::::

## `scale_color_manual()` — Getting the Names Wrong

:::::: columns
::: {.column width="60%"}
```{r}
#| eval: false
# Typo: "sun" instead of the real value "s"
ggplot(pine_df, aes(x = n_s, y = length_mm, color = n_s)) +
  geom_point() +
  scale_color_manual(
    values = c(n = "#2c7fb8", sun = "#d95f0e")
  )
```
:::

:::: {.column width="40%"}
::: callout-warning
**⚠️ Watch out!**

-  Every level that appears in your data **must** get an entry in `values`
  - (and in `labels`, if you supply it), spelled exactly as it appears in the column
  - check with `unique(pine_df$n_s)` first if you're not sure
  - A missing or misspelled level draws as grey with a warning.
:::
::::
::::::

## `scale_fill_manual()` — the `fill` Twin

:::::: columns
::: {.column width="60%"}
```{r}
#| fig-width: 3
#| fig-height: 2.6
#| fig-alt: "Boxplot of needle length by side of tree using custom fill colors (blue for shady, orange for sunny) and a relabeled legend set via scale_fill_manual()."
ggplot(pine_df, aes(x = n_s, y = length_mm, fill = n_s)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(
    name = "Side of tree",
    labels = c(n = "Shady (sheltered)", s = "Sunny"),
    values = c(n = "#2c7fb8", s = "#d95f0e")
  ) +
  labs(x = "Side of tree", y = "Needle length (mm)") +
  theme_minimal(base_size = 9)
```
:::

:::: {.column width="40%"}
::: callout-tip
**🖐 Notice**

Same three arguments, same pattern — `scale_fill_manual()` is `scale_color_manual()` for the `fill` aesthetic instead of `color`. A boxplot's box is `fill`; its jittered points (if you added `geom_point(color = ...)`) would be `color`. A plot can need both at once.
:::
::::
::::::

## Wrap-up

::::::: columns
:::: {.column width="60%"}
Today you:

- Built boxplots with raw points overlaid, not hidden
- Wrote real axis labels, titles, and legend titles with `labs()` — never shipped a default column name
- Mapped a third variable with `color =` / `fill =` inside `aes()`
- Built histograms and density curves to see a whole distribution's shape
- Plotted mean ± SE directly with `stat_summary()` — no separate summary table needed
- Faceted a plot with `facet_wrap()` to check every group at once
- Compared themes, and saved a real, correctly-sized figure with `ggsave()`
- *(Advanced)* Set exact colors and legend text by hand with `scale_color_manual()` / `scale_fill_manual()`

::: callout-tip
**🖐 Before next class**

Finish the worksheet: build a mean ± SE plot faceted by `group`, pick a theme, and save it to `figures/` at 3×3in, 300 dpi.
:::
::::

:::: {.column width="40%"}
::: callout-note
**Where this leads**

You can now import, wrangle, describe, and visualize a dataset end to end. That full pipeline — raw data → clean → summarize → visualize — is the foundation every later statistical test builds on.
:::
::::
:::::::

## Getting unstuck

When code breaks — and it will, that is normal:

1.  Read the **error message** out loud; it usually names the line
2.  Check the usual suspects: `library(tidyverse)` loaded? Is `color`/`fill` inside `aes()` when it should be?
3.  `?function_name` opens the help page
4.  Bring the **exact error** (copy-paste it) to class or office hours

::: callout-note
**✅ Key idea**

Every working scientist googles error messages daily. Getting stuck is not failing — it is the job.
:::
